---
title: "Clustering_Annotation_FindAllMarkers"
author: "SLB"
date: "2025-01-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# 7. Clustering association

1.  Use 'SingleR' is a widely used tool that compares gene expression
    profiles of your clusters to predefined reference databases. Load an
    appropriate reference database. For example:

-   MouseRNAseqData: Reference for murine tissues.
-   ImmGenData: Immunological reference.
-   HumanPrimaryCellAtlasData: For human cells.

2.  Use scType for fast annotation 'scType' is a simpler, more
    customizable tool that uses user-defined or database-available
    marker lists.

    Comparison between 'SingleR' and 'scType'

    | Feature           | SingleR                        | scType            |
    |-------------------|--------------------------------|-------------------|
    | **Ease of use**   | More robust but more technical | Simpler           |
    | **Database**      | Use predefined references      | Use marker lists  |
    | **Customization** | Limited                        | Very customizable |
    | **Speed**         | Moderate                       | Very fast         |

    : Recommendation If you have a well-characterized dataset (such as
    mouse brain transcriptomics), use SingleR with a database such as
    MouseRNAseqData. If you prefer greater flexibility or speed, try
    scType with specific marker lists.



## Create Summarized Experiment from FindAllMarkes function

'FindAllMarkers' will search for differentially expressed genes between clusters. It will search for characterist genes markers for every cluster. On the other hand 'FindMarkers' will search for differentially expressed genes between conditions (i.e. genotype) inside every cluster. 

# Clean up enviroment
```{r}
options(future.globals.maxSize = 28000 * 1024^2) # 16 GB limit
plan("sequential") # Use 4 workers

#Clean enviroment and memory
# Especifica los nombres de los objetos que deseas conservar
keep_objects <- c("seurat_objects", "image_number", "output_path", "data_select", "cluster.all.markers", "cluster_markers_df", "cluster_list")
# Elimina todos los objetos excepto los especificados
rm(list = setdiff(ls(), keep_objects))
gc() #Clean memory
# Set up parameters
set.seed(123)
options(scipen = -1)  # Habilitar notación científica en R  
source("code/global_variables.R")



```

```{r}

# Asegúrate de que el assay está configurado correctamente
DefaultAssay(data_select) <- "RNA"

# Extraer los genes significativos encontrados por FindAllMarkers
significant_genes <- cluster.all.markers %>% 
  dplyr::filter(p_val_adj < 0.05) %>% 
  dplyr::select(gene) %>% 
  dplyr::distinct()%>% 
  dplyr::pull(gene)
  
# Extraer la matriz de expresión para esos genes (utilizando 'layer' en lugar de 'slot')
expression_matrix <- GetAssayData(data_select, layer = "data")[significant_genes, ]

# Extraer las identidades de los clústeres
cluster.ids <- Idents(data_select)

# Crear un objeto SummarizedExperiment con la matriz de expresión
all_markers_data <- SummarizedExperiment(
  assays = list(logcounts = as.matrix(expression_matrix)),
  rowData = DataFrame(gene = rownames(expression_matrix)),
  colData = DataFrame(cluster = cluster.ids)
)


#Alternatively use FindMarkers output. Subtitute all_markers_data for cluster.markers
# Extraer la matriz de expresión y las identidades de los clusters
#cluster.markers <- GetAssayData(data_select, slot = "data")
#cluster.ids <- Idents(data_select)
```

## Allen Brain Atlas clustering

Download matrix Gene expression matrix and Table cell metadata from
Allen Brain Atlas Cell Types Database: RNA-Seq Data
(<https://portal.brain-map.org/atlases-and-data/rnaseq>) Whole Cortex &
Hippocampus - 10x genomics (2020) with 10x-SMART-seq taxonomy (2021)

### Medians 
```{r}
# Cargar los archivos CSV
median_data <- fread("rawdata/medians.csv")
# Verificar las primeras filas
head(median_data)

# Revisar dimensiones
dim(median_data)

# Convertir a matriz
median_matrix <- as.matrix(median_data[ , -1, with = FALSE])  # Asumiendo que la primera columna son los nombres de genes
rownames(median_matrix) <- median_data[[1]]  # Asignar nombres de genes a las filas

# Crear referencia de SingleR
# Crear un nuevo objeto SummarizedExperiment con logcounts
reference_median <- SummarizedExperiment(
    assays = list(logcounts = log1p(median_matrix)), # Aplicar log-transformation
    rowData = DataFrame(gene = rownames(median_matrix)),
    colData = DataFrame(label.main = colnames(median_matrix))
)
# Revisar las etiquetas
colData(reference_median)
# Confirmar que `label.main` contiene los nombres de los tipos celulares
reference_median$label.main <- colnames(median_matrix)

```


```{r}
# Comparar con la base de datos de referencia
annotations_median <- SingleR(test = all_markers_data, 
                       ref = reference_median, 
                       labels = reference_median$label.main,
                       clusters = cluster.ids)

'#annotations_median <- SingleR(test = cluster.markers, 
                       ref = reference_median, 
                       labels = reference_median$label.main,
                       clusters = cluster.ids)
'


# Revisar resultados
head(annotations_median)


```

#### Plot Median

```{r}
new.cluster.median <- annotations_median$labels
names(new.cluster.median) <- levels(data_select)
data_select_median <- RenameIdents(data_select, new.cluster.median)
```

```{r}
# Visualiza con las nuevas identidades
plot <- DimPlot(data_select_median, reduction = "umap", label = TRUE, repel = TRUE)
print(plot)
save_plot("UMAP_median_cluster_plot", plot)

```

### Trimmed_Means
```{r}
# Cargar los archivos CSV
trimmed_means_data <- fread("rawdata/trimmed_means.csv")

# Verificar las primeras filas
head(trimmed_means_data)

# Revisar dimensiones
dim(trimmed_means_data)

# Convertir a matriz
trimmed_means_matrix <- as.matrix(trimmed_means_data[ , -1, with = FALSE])  
rownames(trimmed_means_matrix) <- trimmed_means_data[[1]]

reference_trimmed <- SummarizedExperiment(
    assays = list(logcounts = log1p(trimmed_means_matrix)), # Aplicar log-transformation
    rowData = DataFrame(gene = rownames(trimmed_means_matrix)),
    colData = DataFrame(label.main = colnames(trimmed_means_matrix))
)
# Revisar las etiquetas
colData(reference_trimmed)

# Confirmar que `label.main` contiene los nombres de los tipos celulares
reference_trimmed$label.main <- colnames(trimmed_means_matrix)
```

```{r}
# Comparar con la base de datos de referencia
annotations_trimmed <- SingleR(test = all_markers_data, 
                       ref = reference_trimmed, 
                       labels = reference_trimmed$label.main,
                       clusters = cluster.ids)


# Revisar resultados
head(annotations_trimmed)
```

#### Plot Trimmed

```{r}
new.cluster.trimmed <- annotations_trimmed$labels
names(new.cluster.trimmed) <- levels(data_select)
data_select_trimmed <- RenameIdents(data_select, new.cluster.trimmed)
```

```{r}
# Visualiza con las nuevas identidades
plot <- DimPlot(data_select_trimmed, reduction = "umap", label = TRUE, repel = TRUE)
print(plot)
save_plot("UMAP_trimmed_cluster_plot", plot)

```

## MouseRNAseqData

```{r eval=FALSE, include=FALSE}


ref <- MouseRNAseqData()  # Change according needs
# Extraer la matriz de expresión y las identidades de los clusters
cluster.markers <- GetAssayData(data_select, slot = "data")
cluster.ids <- Idents(data_select)

# Comparar con la base de datos de referencia
annotations_ms <- SingleR(test = all_markers_data, 
                       ref = ref, 
                       labels = ref$label.fine,
                       clusters = cluster.ids)

# Ver los resultados
head(annotations_ms)
```

```{r eval=FALSE, include=FALSE}
new.cluster.ids <- annotations_ms$labels
names(new.cluster.ids) <- levels(data_select)
data_select_ms <- RenameIdents(data_select, new.cluster.ids)
```

```{r eval=FALSE, include=FALSE}
# Visualiza con las nuevas identidades
plot <- DimPlot(data_select_ms, reduction = "umap", label = TRUE, repel = TRUE)
print(plot)
save_plot("UMAP_simple_cluster_plot", plot)

```

## Cerebellum types

Taken from supplementary material from: Kozareva, V., Martin, C.,
Osorno, T. et al. A transcriptomic atlas of mouse cerebellar cortex
comprehensively defines cell types. Nature 598, 214–219 (2021).
<https://doi.org/10.1038/s41586-021-03220-z>

Supplementary Table 2: Highly differentially expressed genes across all
clusters. Genes are limited to top 500 differentially expressed genes
per subtype. Column names represent the following: “avgExpr” is the mean
gene expression in that subtype; “logFC” is the log fold change between
cells within the subtype compared to remaining cells; “pval” is the
p-value, calculated with the two-tailed Gaussian approximation of U
statistic; “padj” is the Benjamini-Hochberg adjusted p-value; “pct_in”
is the percent of cells in the subtype expressing the gene; “pct_out” is
the percent of cells outside the subtype expressing the gene.

```{r eval=FALSE, include=FALSE}
# Leer el archivo Excel (asumiendo que la primera hoja contiene los datos)
file_path <- "rawdata/cb_cell_types.xlsx"
data <- read_excel(file_path)

# Transformar los datos: agrupar subtipos similares y pivotar
cerebellum_data <- data %>%
  dplyr::select(gene, subtype, avgExpr) %>%  # Seleccionar solo las columnas necesarias
  dplyr::mutate(subtype = str_replace(subtype, "_\\d+$", "")) %>%  # Eliminar sufijos _1, _2, etc.
  dplyr::group_by(gene, subtype) %>%  # Agrupar por gen y subtipo
  dplyr::summarise(avgExpr = sum(avgExpr, na.rm = TRUE), .groups = "drop") %>%  # Sumar valores de expresión
  pivot_wider(
    names_from = subtype,  # Usar la columna "subtype" como nombres de columnas
    values_from = avgExpr,  # Usar los valores de la columna "avgExpr" como contenido de las celdas
    values_fill = list(avgExpr = 0)  # Asignar 0 a los valores NA
  )%>%
  dplyr::mutate(Purkinje = coalesce(Purkinje_Aldoc, 0) + coalesce(Purkinje_Anti_Aldoc, 0)) %>%  # Combinar columnas Purkinje_Aldoc y Purkinje_Anti_Aldoc
  dplyr::select(-Purkinje_Aldoc, -Purkinje_Anti_Aldoc)  # Eliminar las columnas originales
# Verificar las primeras filas
head(cerebellum_data)
# Revisar dimensiones
dim(cerebellum_data)

# Convertir a matriz
cerebellum_matrix <- as.matrix(cerebellum_data[ , -1, with = FALSE])  # Asumiendo que la primera columna son los nombres de genes
rownames(cerebellum_matrix) <- cerebellum_data[[1]]  # Asignar nombres de genes a las filas

# Crear referencia de SingleR
# Crear un nuevo objeto SummarizedExperiment con logcounts
reference_cerebellum <- SummarizedExperiment(
    assays = list(logcounts = log1p(cerebellum_matrix)), # Aplicar log-transformation
    rowData = DataFrame(gene = rownames(cerebellum_matrix)),
    colData = DataFrame(label.main = colnames(cerebellum_matrix))
)
# Revisar las etiquetas
colData(reference_cerebellum)

# Confirmar que `label.main` contiene los nombres de los tipos celulares
reference_cerebellum$label.main <- colnames(cerebellum_matrix)

```

```{r eval=FALSE, include=FALSE}
# Comparar con la base de datos de referencia
annotations_cerebellum <- SingleR(test = all_markers_data, 
                       ref = reference_cerebellum, 
                       labels = reference_cerebellum$label.main,
                       clusters = cluster.ids)


# Revisar resultados
head(annotations_cerebellum)

```

```{r eval=FALSE, include=FALSE}
new.cluster.cerebellum <- annotations_cerebellum$labels
names(new.cluster.cerebellum) <- levels(data_select)
data_select_cerebellum <- RenameIdents(data_select, new.cluster.cerebellum)
```

```{r eval=FALSE, include=FALSE}
# Visualiza con las nuevas identidades
plot <- DimPlot(data_select_cerebellum, reduction = "umap", label = TRUE, repel = TRUE)
print(plot)
save_plot("UMAP_cerebellum_cluster_plot", plot)
```

## Save Clusters

```{r}
# Convertir las anotaciones a un data.frame para unirlas con cluster_markers_df
annotations_median_df <- as.data.frame(annotations_median)
annotations_median_df$cluster <- rownames(annotations_median_df)  # Añadir los IDs de los clusters como columna
# Convertir las anotaciones a un data.frame para unirlas con cluster_markers_df
annotations_trimmed_df <- as.data.frame(annotations_trimmed)
annotations_trimmed_df$cluster <- rownames(annotations_trimmed_df)  # Añadir los IDs de los clusters como columna
# Convertir las anotaciones a un data.frame para unirlas con cluster_markers_df
#annotations_ms_df <- as.data.frame(annotations_ms)
#annotations_ms_df$cluster <- rownames(annotations_ms_df)  # Añadir los IDs de los clusters como columna

# Convertir las anotaciones a un data.frame para unirlas con cluster_markers_df
#annotations_cerebellum_df <- as.data.frame(annotations_cerebellum)
#annotations_cerebellum_df$cluster <- rownames(annotations_cerebellum_df)  # Añadir los IDs de los clusters como columna


cell_types_median <- annotations_median_df$labels
cell_types_trimmed <- annotations_trimmed_df$labels
#cell_types_ms <- annotations_ms_df$labels
#cell_types_cerebellum <- annotations_cerebellum_df$labels

# Añadir columna 'labels'
#cluster_markers_df$labels_median <- cell_types_median[cluster_markers_df$cluster + 1]
#cluster_markers_df$labels_trimmed <- cell_types_trimmed[cluster_markers_df$cluster + 1]
#cluster_markers_df$labels_ms <- cell_types_ms[cluster_markers_df$cluster + 1]
#cluster_markers_df$labels_cerebellum <- cell_types_cerebellum[cluster_markers_df$cluster + 1]

cluster.all.markers$labels_median <- cell_types_median[cluster.all.markers$cluster ]
cluster.all.markers$labels_trimmed <- cell_types_trimmed[cluster.all.markers$cluster]
#cluster.all.markers$labels_ms <- cell_types_ms[cluster.all.markers$cluster]
#cluster.all.markers$labels_cerebellum <- cell_types_cerebellum[cluster.all.markers$cluster]

#Generate a list 
cluster_list_all_markers <- cluster.all.markers %>%
  group_split(cluster) %>%
  set_names(paste0("cluster_", unique(cluster.all.markers$cluster)))


#Save xlsx
cluster_list_all_markers %>% 
write_xlsx(paste0(output_path,"tables/","cluster_list_all_markers.xlsx"))
#Save RData

# Save using the name
save(cluster_list_all_markers, file = paste0(output_path, "RData/", "cluster_list_all_markers", ".RData"))



```

```{r}
#Clean enviroment and memory
# Especifica los nombres de los objetos que deseas conservar
keep_objects <- c("data_select", "image_number", "output_path", "cluster_list")
# Elimina todos los objetos excepto los especificados
rm(list = setdiff(ls(), keep_objects))
gc() #Clean memory
# Set up parameters
set.seed(123)
options(scipen = -1)  # Habilitar notación científica en R  
source("code/global_variables.R")  
source("code/01_sc_functions.R")

```

Now, cluster_list will be used as input for the next steps.
