---
editor_options: 
  markdown: 
    wrap: 72
---

# 0. Loading Libraries

## Loading functions

```{r}
output_path <- "output/SC_UbCRE/merged/"
source("code/global_variables.R")
```

```{r}
#source("code/00_packages.R")
```

# 1. Importing data

We start by reading in the data. The Read10X() function reads in the
output of the cellranger pipeline from 10X, returning a unique molecular
identified (UMI) count matrix. The values in this matrix represent the
number of molecules for each feature (i.e. gene; row) that are detected
in each cell (column). Note that more recent versions of cellranger now
also output using the h5 file format, which can be read in using the
Read10X_h5() function in Seurat.

```{r}
tic()
#import data
# Load the PBMC dataset
ctrl_1 <- Seurat::Read10X(data.dir = "rawdata/SC/AWGC-1_Ctrl-1_results")
ctrl_2 <- Seurat::Read10X(data.dir = "rawdata/SC/AWGC-2_Ctrl-2_results")
ko_1 <- Seurat::Read10X(data.dir = "rawdata/SC/AWGC-3_KOut-1_results")
ko_2 <- Seurat::Read10X(data.dir = "rawdata/SC/AWGC-4_KOut-2_results")

# Initialize the Seurat object with the raw (non-normalized data).
ubc_c1 <- CreateSeuratObject(counts = ctrl_1, project = "WT_1", min.cells = 3, min.features = 200)
ubc_c2 <- CreateSeuratObject(counts = ctrl_2, project = "WT_2", min.cells = 3, min.features = 200)
ubc_ko1 <- CreateSeuratObject(counts = ko_1, project = "KO_1", min.cells = 3, min.features = 200)
ubc_ko2 <- CreateSeuratObject(counts = ko_2, project = "KO_2", min.cells = 3, min.features = 200)



# Merge all four Seurat objects into a single object
data_merge <- merge(ubc_c1, y = c(ubc_c2, ubc_ko1, ubc_ko2), 
                  add.cell.ids = c("wt_1", "wt_2", "ko_1", "ko_2"), 
                  project = "UBC_experiment")


# Join layer of the merged object
data_merge <- JoinLayers(data_merge)

toc()
```

## Saving combined data as HDF5

```{r}
tic()
SaveH5Seurat(data_merge,
             filename = paste0(output_path, "RData/","UBC_experiment.h5Seurat"),
             overwrite = TRUE, 
             verbose = TRUE)
toc()
```

# 2. Data preprocessing for each library

## QC and selecting cells for further analysis

Seurat allows you to easily explore QC metrics and filter cells based on
any user-defined criteria. A few QC metrics commonly used by the
community include \## Identify mitochondrial genes The number of unique
genes detected in each cell. Low-quality cells or empty droplets will
often have very few genes Cell doublets or multiplets may exhibit an
aberrantly high gene count Similarly, the total number of molecules
detected within a cell (correlates strongly with unique genes) The
percentage of reads that map to the mitochondrial genome Low-quality /
dying cells often exhibit extensive mitochondrial contamination We
calculate mitochondrial QC metrics with the 'PercentageFeatureSet()'
function, which calculates the percentage of counts originating from a
set of features We use the set of all genes starting with 'MT-' as a set
of mitochondrial genes

```{r}
# The [[ operator can add columns to object metadata. This is a great place to stash QC stats
  data_merge[["percent.mt"]] <- PercentageFeatureSet(data_merge, pattern = "^mt-")
```

# QC cell Pre-Filter

```{r}
# Visualize QC metrics as a violin plot
# Initialize figure counter
figure_counter <- 1

  # Generate the violin plot for QC metrics
  QC_VlnPlot <- VlnPlot(data_merge, features = c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.ribo"), ncol = 4)
  
  # Print the plot
  print(QC_VlnPlot)
  
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_QC_VlnPlot_prefilter")
  
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), QC_VlnPlot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), QC_VlnPlot, width = 8, height = 6, units = "in")
  
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable

# `figure_counter` now holds the next available figure number for subsequent steps

```

```{r}
# FeatureScatter is typically used to visualize feature-feature relationships, but can be used
# for anything calculated by the object, i.e. columns in object metadata, PC scores etc.
plot1 <- FeatureScatter(data_merge, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(data_merge, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot <- plot1 + plot2

  # Print the features plot
  print(plot)
  
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_Features_Plot_prefilter")
  
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 8, height = 6, units = "in")
  
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable

```

# QC Post-filtering

We filter cells that have unique feature counts over 7,500 or less than
850 We filter cells that have \>1% mitochondrial counts

```{r}
  data_merge <- subset(data_merge, subset = nFeature_RNA > 850 & nFeature_RNA < 7500 & percent.mt < 1)
```

```{r}
# Visualize QC metrics as a violin plot
# Initialize figure counter
figure_counter <- 1

  # Generate the violin plot for QC metrics
  QC_VlnPlot <- VlnPlot(data_merge, features = c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.ribo"), ncol = 4)
  
  # Print the plot
  print(QC_VlnPlot)
  
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_QC_VlnPlot_filter")
  
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), QC_VlnPlot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), QC_VlnPlot, width = 8, height = 6, units = "in")
  
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable

# `figure_counter` now holds the next available figure number for subsequent steps

```

```{r}
# FeatureScatter is typically used to visualize feature-feature relationships, but can be used
# for anything calculated by the object, i.e. columns in object metadata, PC scores etc.
plot1 <- FeatureScatter(data_merge, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(data_merge, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot <- plot1 + plot2

  # Print the features plot
  print(plot)
  
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_Features_Plot_filter")
  
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 8, height = 6, units = "in")
  
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable

```

# 3.Normalizing the data

```{r}
tic()
  
  all.genes <- rownames(data_merge)
  # Normalize, find variable features, and scale the data
  data_merge <- NormalizeData(data_merge, normalization.method = "LogNormalize", scale.factor = 10000) %>%
          FindVariableFeatures(selection.method = "vst", nfeatures = 2000) 
  
  # Identify the 10 most highly variable genes
  top10 <- head(VariableFeatures(data_merge), 10)
  # plot variable features with and without labels
  plot <- VariableFeaturePlot(data_merge)
  plot <- LabelPoints(plot = plot, points = top10, repel = TRUE)
  plot
 # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_Highly_Varible_Genes")
  
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 8, height = 6, units = "in")
  
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
  
  data_merge <- ScaleData(data_merge, features = all.genes)


toc()
```

# 4.Linear dimensional reduction

Next we perform PCA on the scaled data. By default, only the previously
determined variable features are used as input, but can be defined using
features argument if you wish to choose a different subset (if you do
want to use a custom subset of features, make sure you pass these to
ScaleData first).

For the first principal components, Seurat outputs a list of genes with
the most positive and negative loadings, representing modules of genes
that exhibit either correlation (or anti-correlation) across
single-cells in the dataset.

```{r}
tic()

#Add PCA 

  data_merge <- RunPCA(data_merge, features = VariableFeatures(object = data_merge))
  
  # Examine and visualize PCA results a few different ways
  print(data_merge[["pca"]], dims = 1:5, nfeatures = 5)
  
  plot <- VizDimLoadings(data_merge, dims = 1:2, reduction = "pca")
  print(plot)
 # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_VimPCA_Loadings")
  
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 8, height = 6, units = "in")
  
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
  
toc()

```

## Elbow Plot

Determine the ‘dimensionality’ of the dataset To overcome the extensive
technical noise in any single feature for scRNA-seq data, Seurat
clusters cells based on their PCA scores, with each PC essentially
representing a ‘metafeature’ that combines information across a
correlated feature set. The top principal components therefore represent
a robust compression of the dataset. However, how many components should
we choose to include? 10? 20? 100?

In Macosko et al, we implemented a resampling test inspired by the
JackStraw procedure. While still available in Seurat (see previous
vignette), this is a slow and computationally expensive procedure, and
we is no longer routinely used in single cell analysis.

An alternative heuristic method generates an ‘Elbow plot’: a ranking of
principle components based on the percentage of variance explained by
each one (ElbowPlot() function). In this example, we can observe an
‘elbow’ around PC9-10, suggesting that the majority of true signal is
captured in the first 10 PCs.

```{r}
tic()

  #ElbowPlot
  plot <- ElbowPlot(data_merge)
  print(plot)
 # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_ElbowPlot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 8, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
  
toc()


```

## Cluster cells

Seurat applies a graph-based clustering approach, building upon initial
strategies in (Macosko et al). Importantly, the distance metric which
drives the clustering analysis (based on previously identified PCs)
remains the same. However, our approach to partitioning the cellular
distance matrix into clusters has dramatically improved. Our approach
was heavily inspired by recent manuscripts which applied graph-based
clustering approaches to scRNA-seq data [SNN-Cliq, Xu and Su,
Bioinformatics, 2015] and CyTOF data [PhenoGraph, Levine et al., Cell,
2015]. Briefly, these methods embed cells in a graph structure - for
example a K-nearest neighbor (KNN) graph, with edges drawn between cells
with similar feature expression patterns, and then attempt to partition
this graph into highly interconnected ‘quasi-cliques’ or ‘communities’.

As in PhenoGraph, we first construct a KNN graph based on the euclidean
distance in PCA space, and refine the edge weights between any two cells
based on the shared overlap in their local neighborhoods (Jaccard
similarity). This step is performed using the FindNeighbors() function,
and takes as input the previously defined dimensionality of the dataset
(first 10 PCs).

To cluster the cells, we next apply modularity optimization techniques
such as the Louvain algorithm (default) or SLM [SLM, Blondel et al.,
Journal of Statistical Mechanics], to iteratively group cells together,
with the goal of optimizing the standard modularity function. The
FindClusters() function implements this procedure, and contains a
resolution parameter that sets the ‘granularity’ of the downstream
clustering, with increased values leading to a greater number of
clusters. We find that setting this parameter between 0.4-1.2 typically
returns good results for single-cell datasets of around 3K cells.
Optimal resolution often increases for larger datasets. The clusters can
be found using the Idents() function.

```{r}
data_merge <- FindNeighbors(data_merge, dims = 1:11)
data_merge <- FindClusters(data_merge, resolution = 0.5)
```

## UMAP dimensions

```{r}
tic()
  #UMAP
  data_merge <- RunUMAP(data_merge, dims = 1:11)
toc()
```

## Plot UMAP & UMAP Features

```{r}
tic()
 # note that you can set `label = TRUE` or use the LabelClusters function to help label individual clusters
  plot <- DimPlot(data_merge, reduction = "umap", label = TRUE)

  print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_UMAP")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 8, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
    


# note that you can set `label = TRUE` or use the LabelClusters function to help label individual clusters
  plot <- DimPlot(data_merge, reduction = "umap", label = TRUE)
  plot <- DimPlot(data_merge, reduction = "umap", split.by="orig.ident", ncol = 2)
  
  print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_UMAP_ident")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 8, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
  
  
  #UMAP by features
  features <- c("nFeature_RNA", "nCount_RNA")
  feature_plot <- FeaturePlot(data_merge, features = features)
    print(feature_plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_Feature_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), feature_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), feature_plot, width = 8, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
toc()

```

## UMAP HVGs

```{r}
top <-head(VariableFeatures(data_merge),10)

  #UMAP by features
  feature_plot <- FeaturePlot(data_merge, features = top, ncol = 5)
    print(feature_plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_Feature_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), feature_plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), feature_plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
  
```

## UMAP Dnajc5

```{r}
  #UMAP by features
  feature_plot <- FeaturePlot(data_merge, features = "Dnajc5",split.by="orig.ident", ncol = 2 )
    print(feature_plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_Feature_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), feature_plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), feature_plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
  
```

```{r}
tic()
SaveH5Seurat(data_merge,
             filename = paste0(output_path, "RData/","UBC_experiment.h5Seurat"),
             overwrite = TRUE, 
             verbose = TRUE)
toc()
```

#5. Finding differentially expressed features (cluster biomarkers)
Seurat can help you find markers that define clusters via differential
expression (DE). By default, it identifies positive and negative markers
of a single cluster (specified in ident.1), compared to all other cells.
FindAllMarkers() automates this process for all clusters, but you can
also test groups of clusters vs. each other, or against all cells.

In Seurat v5, we use the presto package (as described here and available
for installation here), to dramatically improve the speed of DE
analysis, particularly for large datasets. For users who are not using
presto, you can examine the documentation for this function
(?FindMarkers) to explore the min.pct and logfc.threshold parameters,
which can be increased in order to increase the speed of DE testing.

```{r}
# find markers for every cluster compared to all remaining cells, report only the positive
# ones
# Encuentra marcadores de todos los clusters
cluster.markers <- FindAllMarkers(data_merge, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)

data_merge.markers <- FindAllMarkers(data_merge, 
                              logfc.threshold = 0.1,
                              min.pct = 0.1,
                              only.pos = TRUE)
```

```{r}
#Generate a list 
cluster_list <- data_merge.markers %>%
  group_split(cluster) %>%
  set_names(paste0("cluster_", unique(data_merge.markers$cluster)))

cluster_list %>% 
write_xlsx(paste0(output_path,"tables/","cluster_list.xlsx"))
```

## HeatMap

```{r}
data_merge.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10


plot <- DoHeatmap(data_merge, features = top10$gene) 
print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "HeatMap_cluster_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
```

# 6. Clustering association

1.  Use 'SingleR' is a widely used tool that compares gene expression
    profiles of your clusters to predefined reference databases. Load an
    appropriate reference database. For example:

-   MouseRNAseqData: Reference for murine tissues.
-   ImmGenData: Immunological reference.
-   HumanPrimaryCellAtlasData: For human cells.

2.  Use scType for fast annotation 'scType' is a simpler, more
    customizable tool that uses user-defined or database-available
    marker lists.

    Comparison between 'SingleR' and 'scType'
    
    | Feature           | SingleR                         | scType            |
    |-------------------|---------------------------------|-------------------|
    | **Ease of use**   | More robust but more technical  |  Simpler          |
    | **Database**      | Use predefined references       | Use marker lists  |
    | **Customization** | Limited                         | Very customizable |
    | **Speed**         | Moderate                        |  Very fast        |

    : Recommendation If you have a well-characterized dataset (such as
    mouse brain transcriptomics), use SingleR with a database such as
    MouseRNAseqData. If you prefer greater flexibility or speed, try
    scType with specific marker lists.


## Allen Brain Atlas clustering
Download matrix Gene expression matrix and Table cell metadata from Allen Brain Atlas Cell Types Database: RNA-Seq Data (https://portal.brain-map.org/atlases-and-data/rnaseq)
Whole Cortex & Hippocampus - 10x genomics (2020)
with 10x-SMART-seq taxonomy (2021)
```{r}

# Extraer la matriz de expresión y las identidades de los clusters
cluster.markers <- GetAssayData(data_merge, slot = "data")
cluster.ids <- Idents(data_merge)

# Cargar los archivos CSV
median_data <- fread("rawdata/medians.csv")
trimmed_means_data <- fread("rawdata/trimmed_means.csv")

# Verificar las primeras filas
head(median_data)
head(trimmed_means_data)

# Revisar dimensiones
dim(median_data)
dim(trimmed_means_data)

# Convertir a matriz
median_matrix <- as.matrix(median_data[ , -1, with = FALSE])  # Asumiendo que la primera columna son los nombres de genes
rownames(median_matrix) <- median_data[[1]]  # Asignar nombres de genes a las filas

trimmed_means_matrix <- as.matrix(trimmed_means_data[ , -1, with = FALSE])  # Similar al anterior
rownames(trimmed_means_matrix) <- trimmed_means_data[[1]]

# Crear referencia de SingleR
# Crear un nuevo objeto SummarizedExperiment con logcounts
reference_median <- SummarizedExperiment(
    assays = list(logcounts = log1p(median_matrix)), # Aplicar log-transformation
    rowData = DataFrame(gene = rownames(median_matrix)),
    colData = DataFrame(label.main = colnames(median_matrix))
)
# Revisar las etiquetas
colData(reference_median)

# Confirmar que `label.main` contiene los nombres de los tipos celulares
reference_median$label.main <- colnames(median_matrix)

reference_trimmed <- SummarizedExperiment(
    assays = list(logcounts = log1p(trimmed_means_matrix)), # Aplicar log-transformation
    rowData = DataFrame(gene = rownames(trimmed_means_matrix)),
    colData = DataFrame(label.main = colnames(trimmed_means_matrix))
)
# Revisar las etiquetas
colData(reference_trimmed)

# Confirmar que `label.main` contiene los nombres de los tipos celulares
reference_trimmed$label.main <- colnames(trimmed_means_matrix)
```


```{r}
# Comparar con la base de datos de referencia
annotations_median <- SingleR(test = cluster.markers, 
                       ref = reference_median, 
                       labels = reference_median$label.main,
                       clusters = cluster.ids)


# Revisar resultados
head(annotations_median)
```


```{r}
# Comparar con la base de datos de referencia
annotations_trimmed <- SingleR(test = cluster.markers, 
                       ref = reference_trimmed, 
                       labels = reference_trimmed$label.main,
                       clusters = cluster.ids)


# Revisar resultados
head(annotations_trimmed)
```

## Plot Median
```{r}
new.cluster.median <- annotations_median$labels
names(new.cluster.median) <- levels(data_merge)
data_merge_median <- RenameIdents(data_merge, new.cluster.median)
```


```{r}
# Visualiza con las nuevas identidades
plot <- DimPlot(data_merge_median, reduction = "umap", label = TRUE, repel = TRUE)
print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "UMAP_simple_cluster_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable

```

## Plot Trimmed
```{r}
new.cluster.trimmed <- annotations_trimmed$labels
names(new.cluster.trimmed) <- levels(data_merge)
data_merge_trimmed <- RenameIdents(data_merge, new.cluster.median)
```


```{r}
# Visualiza con las nuevas identidades
plot <- DimPlot(data_merge_trimmed, reduction = "umap", label = TRUE, repel = TRUE)
print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "UMAP_simple_cluster_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable

```


## MouseRNAseqData
```{r}


ref <- MouseRNAseqData()  # Change according needs
# Extraer la matriz de expresión y las identidades de los clusters
cluster.markers <- GetAssayData(data_merge, slot = "data")
cluster.ids <- Idents(data_merge)

# Comparar con la base de datos de referencia
annotations_ms <- SingleR(test = cluster.markers, 
                       ref = ref, 
                       labels = ref$label.main,
                       clusters = cluster.ids)

# Ver los resultados
head(annotations)
```


```{r}
new.cluster.ids <- annotations_ms$labels
names(new.cluster.ids) <- levels(data_merge)
data_merge_ms <- RenameIdents(data_merge, new.cluster.ids)
```


```{r}
# Visualiza con las nuevas identidades
plot <- DimPlot(data_merge_ms, reduction = "umap", label = TRUE, repel = TRUE)
print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "UMAP_simple_cluster_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable

```
