

# 0. Loading Libraries

## Loading functions
```{r}
output_path <- "output/SC_UbCRE/Integrated/"
source("code/global_variables.R")
```


```{r}
source("code/00_packages.R")
```

# 1. Importing data
We start by reading in the data. The Read10X() function reads in the output of the cellranger pipeline from 10X, returning a unique molecular identified (UMI) count matrix. The values in this matrix represent the number of molecules for each feature (i.e. gene; row) that are detected in each cell (column). Note that more recent versions of cellranger now also output using the h5 file format, which can be read in using the Read10X_h5() function in Seurat.
```{r}
tic()
#import data
# Load the PBMC dataset
ctrl_1 <- Seurat::Read10X(data.dir = "rawdata/SC/AWGC-1_Ctrl-1_results")
ctrl_2 <- Seurat::Read10X(data.dir = "rawdata/SC/AWGC-2_Ctrl-2_results")
ko_1 <- Seurat::Read10X(data.dir = "rawdata/SC/AWGC-3_KOut-1_results")
ko_2 <- Seurat::Read10X(data.dir = "rawdata/SC/AWGC-4_KOut-2_results")

# Initialize the Seurat object with the raw (non-normalized data).
ubc_c1 <- CreateSeuratObject(counts = ctrl_1, project = "wt_1", min.cells = 3, min.features = 200)
ubc_c2 <- CreateSeuratObject(counts = ctrl_2, project = "wt_2", min.cells = 3, min.features = 200)
ubc_ko1 <- CreateSeuratObject(counts = ko_1, project = "ko_1", min.cells = 3, min.features = 200)
ubc_ko2 <- CreateSeuratObject(counts = ko_2, project = "ko_2", min.cells = 3, min.features = 200)

toc()
```

Create vector with SeurtalObjects names for automatization

```{r}
seurat_names <- c("ubc_c1", "ubc_c2", "ubc_ko1", "ubc_ko2")
```

```{r}
tic()
# Export results ####
lapply(seurat_names, function(x) {
  obj <- get(x)  # Retrieve the object
  save(obj, file = paste0(output_path, "RData/", x, ".RData"))
})
toc()
```
# 2. Data preprocessing for each library
## QC and selecting cells for further analysis
Seurat allows you to easily explore QC metrics and filter cells based on any user-defined criteria. A few QC metrics commonly used by the community include
## Identify mitochondrial genes
The number of unique genes detected in each cell.
Low-quality cells or empty droplets will often have very few genes
Cell doublets or multiplets may exhibit an aberrantly high gene count
Similarly, the total number of molecules detected within a cell (correlates strongly with unique genes)
The percentage of reads that map to the mitochondrial genome
Low-quality / dying cells often exhibit extensive mitochondrial contamination
We calculate mitochondrial QC metrics with the 'PercentageFeatureSet()' function, which calculates the percentage of counts originating from a set of features
We use the set of all genes starting with 'MT-' as a set of mitochondrial genes

```{r}
# The [[ operator can add columns to object metadata. This is a great place to stash QC stats
lapply(seurat_names, function(x) {
  obj <- get(x)  # Retrieve the Seurat object
  obj[["percent.mt"]] <- PercentageFeatureSet(obj, pattern = "^MT-")
  assign(x, obj, envir = .GlobalEnv)  # Reassign the modified object back to the global environment
})



```
## QC cell Pre-Filter

```{r}
# Visualize QC metrics as a violin plot
# Initialize figure counter
figure_counter <- 1

# Create and save the violin plots with incremented figure numbers
lapply(seq_along(seurat_names), function(i) {
  # Retrieve the Seurat object
  obj <- get(seurat_names[i])
  
  # Generate the violin plot for QC metrics
  QC_VlnPlot <- VlnPlot(obj, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
  
  # Print the plot
  print(QC_VlnPlot)
  
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_QC_VlnPlot_prefilter_", seurat_names[i])
  
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), QC_VlnPlot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), QC_VlnPlot, width = 8, height = 6, units = "in")
  
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
})

# `figure_counter` now holds the next available figure number for subsequent steps

```


```{r}


lapply(seq_along(seurat_names), function(i) {
  # Retrieve the Seurat object
  obj <- get(seurat_names[i])
  
 # FeatureScatter is typically used to visualize feature-feature relationships, but can be used
# for anything calculated by the object, i.e. columns in object metadata, PC scores etc.
plot1 <- FeatureScatter(obj, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot <- plot1 + plot2

  # Print the features plot
  print(plot)
  
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_Features_Plot_prefilter_", seurat_names[i])
  
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 8, height = 6, units = "in")
  
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
})




```

## QC cell Filter
We filter cells that have unique feature counts over 7,500 or less than 850
We filter cells that have >1% mitochondrial counts

```{r}

lapply(seurat_names, function(x) {
  obj <- get(x)  # Retrieve the Seurat object
  obj <- subset(obj, subset = nFeature_RNA > 850 & nFeature_RNA < 7500 & percent.mt < 1)
  assign(x, obj, envir = .GlobalEnv)  # Reassign the modified object back to the global environment
})


```

```{r}
# Visualize QC metrics as a violin plot
# Initialize figure counter
figure_counter <- 1

# Create and save the violin plots with incremented figure numbers
lapply(seq_along(seurat_names), function(i) {
  # Retrieve the Seurat object
  obj <- get(seurat_names[i])
  
  # Generate the violin plot for QC metrics
  QC_VlnPlot <- VlnPlot(obj, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
  
  # Print the plot
  print(QC_VlnPlot)
  
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_QC_VlnPlot_filter_", seurat_names[i])
  
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), QC_VlnPlot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), QC_VlnPlot, width = 8, height = 6, units = "in")
  
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
})

# `figure_counter` now holds the next available figure number for subsequent steps

```
```{r}


lapply(seq_along(seurat_names), function(i) {
  # Retrieve the Seurat object
  obj <- get(seurat_names[i])
  
 # FeatureScatter is typically used to visualize feature-feature relationships, but can be used
# for anything calculated by the object, i.e. columns in object metadata, PC scores etc.
plot1 <- FeatureScatter(obj, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot <- plot1 + plot2

  # Print the features plot
  print(plot)
  
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_Features_Plot_filter_", seurat_names[i])
  
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 8, height = 6, units = "in")
  
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
})




```

# 3.Normalizing the data

```{r}
tic()

#Filtering, normalization and scaling for all libraries
lapply(seq_along(seurat_names), function(x) {
  # Retrieve the Seurat object by name
  obj <- get(seurat_names[x])
  all.genes <- rownames(obj)
  
  # Normalize, find variable features, and scale the data
  norm <- NormalizeData(obj, normalization.method = "LogNormalize", scale.factor = 10000) %>%
          FindVariableFeatures(selection.method = "vst", nfeatures = 2000) 
  
  # Identify the 10 most highly variable genes
  top10 <- head(VariableFeatures(norm), 10)
  # plot variable features with and without labels
  plot <- VariableFeaturePlot(norm)
  plot <- LabelPoints(plot = plot, points = top10, repel = TRUE)
  
 # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_Highly_Varible_Genes_", seurat_names[x])
  
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 8, height = 6, units = "in")
  
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
  
  
  
  norm <- ScaleData(norm, features = all.genes)
  
  # Assign the new object with the suffix "norm_" to the global environment
  assign(paste0("norm_", seurat_names[x]), norm, envir = .GlobalEnv)
  save(norm, file = paste0(output_path, "RData/", paste0("norm_", seurat_names[x]), ".RData"))
})
toc()
```


# 4.Linear dimensional reduction
Next we perform PCA on the scaled data. By default, only the previously determined variable features are used as input, but can be defined using features argument if you wish to choose a different subset (if you do want to use a custom subset of features, make sure you pass these to ScaleData first).

For the first principal components, Seurat outputs a list of genes with the most positive and negative loadings, representing modules of genes that exhibit either correlation (or anti-correlation) across single-cells in the dataset.

```{r}
tic()
# Modify the vector by adding the "norm_" prefix
seurat_names <- paste0("norm_", seurat_names)


#Add PCA 
lapply(seq_along(seurat_names), function(x) {
  # Retrieve the Seurat object by name
  obj <- get(seurat_names[x])
  
  obj <- RunPCA(obj, features = VariableFeatures(object = obj))
  
  # Examine and visualize PCA results a few different ways
  print(obj[["pca"]], dims = 1:5, nfeatures = 5)
  
  plot <- VizDimLoadings(obj, dims = 1:2, reduction = "pca")
  print(plot)
 # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_VimPCA_Loadings_", seurat_names[x])
  
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 8, height = 6, units = "in")
  
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
  
  assign(seurat_names[x], obj, envir = .GlobalEnv)
  })
toc()

```

## Elbow Plot
Determine the ‘dimensionality’ of the dataset
To overcome the extensive technical noise in any single feature for scRNA-seq data, Seurat clusters cells based on their PCA scores, with each PC essentially representing a ‘metafeature’ that combines information across a correlated feature set. The top principal components therefore represent a robust compression of the dataset. However, how many components should we choose to include? 10? 20? 100?

In Macosko et al, we implemented a resampling test inspired by the JackStraw procedure. While still available in Seurat (see previous vignette), this is a slow and computationally expensive procedure, and we is no longer routinely used in single cell analysis.

An alternative heuristic method generates an ‘Elbow plot’: a ranking of principle components based on the percentage of variance explained by each one (ElbowPlot() function). In this example, we can observe an ‘elbow’ around PC9-10, suggesting that the majority of true signal is captured in the first 10 PCs.

```{r}
tic()
#Add PCA 
lapply(seq_along(seurat_names), function(x) {
  # Retrieve the Seurat object by name
  obj <- get(seurat_names[x])
  
  #ElbowPlot
  plot <- ElbowPlot(obj)
  print(plot)
 # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_ElbowPlot_", seurat_names[x])
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 8, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
  
  })
toc()


```

## Saving Results
```{r}
tic()
# Export results ####
lapply(seurat_names, function(x) {
  obj <- get(x)  # Retrieve the object
  save(obj, file = paste0(output_path, "RData/", x, ".RData"))
})
toc()
```
### Load RDatas
```{r}
tic()
# Load results ####
lapply(seurat_names, function(x) {
  obj <- get(x)  # Retrieve the object
  load(obj, file = paste0(output_path, "RData/", x, ".RData"))
})
toc()


```


## Cluster cells
Seurat applies a graph-based clustering approach, building upon initial strategies in (Macosko et al). Importantly, the distance metric which drives the clustering analysis (based on previously identified PCs) remains the same. However, our approach to partitioning the cellular distance matrix into clusters has dramatically improved. Our approach was heavily inspired by recent manuscripts which applied graph-based clustering approaches to scRNA-seq data [SNN-Cliq, Xu and Su, Bioinformatics, 2015] and CyTOF data [PhenoGraph, Levine et al., Cell, 2015]. Briefly, these methods embed cells in a graph structure - for example a K-nearest neighbor (KNN) graph, with edges drawn between cells with similar feature expression patterns, and then attempt to partition this graph into highly interconnected ‘quasi-cliques’ or ‘communities’.

As in PhenoGraph, we first construct a KNN graph based on the euclidean distance in PCA space, and refine the edge weights between any two cells based on the shared overlap in their local neighborhoods (Jaccard similarity). This step is performed using the FindNeighbors() function, and takes as input the previously defined dimensionality of the dataset (first 10 PCs).

To cluster the cells, we next apply modularity optimization techniques such as the Louvain algorithm (default) or SLM [SLM, Blondel et al., Journal of Statistical Mechanics], to iteratively group cells together, with the goal of optimizing the standard modularity function. The FindClusters() function implements this procedure, and contains a resolution parameter that sets the ‘granularity’ of the downstream clustering, with increased values leading to a greater number of clusters. We find that setting this parameter between 0.4-1.2 typically returns good results for single-cell datasets of around 3K cells. Optimal resolution often increases for larger datasets. The clusters can be found using the Idents() function.

```{r}
tic()
#Add PCA 
lapply(seq_along(seurat_names), function(x) {
  # Retrieve the Seurat object by name
  obj <- get(seurat_names[x])
  #Clustering
  
  obj <- FindNeighbors(obj, dims = 1:11)
  obj <- FindClusters(obj, resolution = 0.5)
  
  #UMAP
  obj <- RunUMAP(obj, dims = 1:11)
  # note that you can set `label = TRUE` or use the LabelClusters function to help label individual clusters
  plot <- DimPlot(obj, reduction = "umap", label = TRUE)
  print(plot)
 # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_UMAP_", seurat_names[x])
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 8, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
  
  })
toc()
```


```{r}
#Save Normalized and dimentionally reduced data
tic()
# Export results ####
lapply(seurat_names, function(x) {
  obj <- get(x)  # Retrieve the object
  save(obj, file = paste0(output_path, "RData/", x, ".RData"))
})
toc()

tic()
# Load results ####
lapply(seurat_names, function(x) {
  obj <- get(x)  # Retrieve the object
  load(obj, file = paste0(output_path, "RData/", x, ".RData"))
})
toc()
```


# 4. Library integration

```{r}
tic()
# Encuentra anclajes de integración
anchors <- FindIntegrationAnchors(object.list = lapply(seurat_names, get))
# Integra los datos
data_int <- IntegrateData(anchorset = anchors)

# Realiza una reducción de dimensionalidad
data_int <- ScaleData(data_int) %>% RunPCA() %>% RunUMAP(dims = 1:11)

# Agrupa las células en clústeres
data_int <- FindNeighbors(data_int, dims = 1:11) %>% FindClusters(resolution = 0.5)


toc()
```
## Save data_int object

```{r}
tic()
SaveH5Seurat(data_int,
             filename = paste0(output_path, "RData/","UBC_experiment.h5Seurat"),
             overwrite = TRUE, 
             verbose = TRUE)
toc()
```


# 5.Plot UMAP & UMAP Features

```{r}
tic()
 # note that you can set `label = TRUE` or use the LabelClusters function to help label individual clusters
  plot <- DimPlot(data_int, reduction = "umap", label = TRUE)

  print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_UMAP")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 8, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
    


# note that you can set `label = TRUE` or use the LabelClusters function to help label individual clusters
  plot <- DimPlot(data_int, reduction = "umap", label = TRUE)
  plot <- DimPlot(data_int, reduction = "umap", split.by="orig.ident", ncol = 2)
  
  print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_UMAP_ident")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 8, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
  
  
  #UMAP by features
  features <- c("nFeature_RNA", "nCount_RNA")
  feature_plot <- FeaturePlot(data_int, features = features)
    print(feature_plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_Feature_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), feature_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), feature_plot, width = 8, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
toc()

```

## UMAP HVGs

```{r}
top <-head(VariableFeatures(data_int),10)

  #UMAP by features
  feature_plot <- FeaturePlot(data_int, features = top, ncol = 5)
    print(feature_plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_Feature_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), feature_plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), feature_plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
  
```

## UMAP Dnajc5

```{r}
  #UMAP by features
  feature_plot <- FeaturePlot(data_int, features = "Dnajc5",split.by="orig.ident", ncol = 2 )
    print(feature_plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_Feature_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), feature_plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), feature_plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
  
```
### Saving Seurat Object

```{r}
tic()
#save(data_int, file = paste0(output_path, "RData/", data_int, ".RData"))

SaveH5Seurat(data_int,
             filename = paste0(output_path, "RData/","UBC_experiment.h5Seurat"),
             overwrite = TRUE, 
             verbose = TRUE)
toc()
```

### Load Seurat object
```{r}
# Load the .h5Seurat file
file_path <- paste0(output_path, "RData/UBC_experiment.h5Seurat")
data_int <- LoadH5Seurat(file_path)
```


# 6. Finding differentially expressed features (cluster biomarkers)
Seurat can help you find markers that define clusters via differential
expression (DE). By default, it identifies positive and negative markers
of a single cluster (specified in ident.1), compared to all other cells.
FindAllMarkers() automates this process for all clusters, but you can
also test groups of clusters vs. each other, or against all cells.

In Seurat v5, we use the presto package (as described here and available
for installation here), to dramatically improve the speed of DE
analysis, particularly for large datasets. For users who are not using
presto, you can examine the documentation for this function
(?FindMarkers) to explore the min.pct and logfc.threshold parameters,
which can be increased in order to increase the speed of DE testing.

```{r}
# find markers for every cluster compared to all remaining cells, report only the positive
# ones
# Encuentra marcadores de todos los clusters
data_int.markers <- FindAllMarkers(data_int, 
                              logfc.threshold = 0.1,
                              min.pct = 0.1,
                              only.pos = TRUE)
```

```{r}
#Generate a list 
cluster_list <- data_int.markers %>%
  group_split(cluster) %>%
  set_names(paste0("cluster_", unique(data_int.markers$cluster)))

cluster_list %>% 
write_xlsx(paste0(output_path,"tables/","cluster_list.xlsx"))
```

## HeatMap

```{r}
data_int.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10


plot <- DoHeatmap(data_int, features = top10$gene) 
print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "HeatMap_cluster_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
```


# 8. Saving

```{r}
tic()
SaveH5Seurat(data_int,
             filename = paste0(output_path, "RData/","UBC_experiment.h5Seurat"),
             overwrite = TRUE, 
             verbose = TRUE)
toc()
```

# 7. Clustering association

1.  Use 'SingleR' is a widely used tool that compares gene expression
    profiles of your clusters to predefined reference databases. Load an
    appropriate reference database. For example:

-   MouseRNAseqData: Reference for murine tissues.
-   ImmGenData: Immunological reference.
-   HumanPrimaryCellAtlasData: For human cells.

2.  Use scType for fast annotation 'scType' is a simpler, more
    customizable tool that uses user-defined or database-available
    marker lists.

    Comparison between 'SingleR' and 'scType'
    
    | Feature           | SingleR                         | scType            |
    |-------------------|---------------------------------|-------------------|
    | **Ease of use**   | More robust but more technical  |  Simpler          |
    | **Database**      | Use predefined references       | Use marker lists  |
    | **Customization** | Limited                         | Very customizable |
    | **Speed**         | Moderate                        |  Very fast        |

    : Recommendation If you have a well-characterized dataset (such as
    mouse brain transcriptomics), use SingleR with a database such as
    MouseRNAseqData. If you prefer greater flexibility or speed, try
    scType with specific marker lists.


## Allen Brain Atlas clustering
Download matrix Gene expression matrix and Table cell metadata from Allen Brain Atlas Cell Types Database: RNA-Seq Data (https://portal.brain-map.org/atlases-and-data/rnaseq)
Whole Cortex & Hippocampus - 10x genomics (2020)
with 10x-SMART-seq taxonomy (2021)
```{r}

# Extraer la matriz de expresión y las identidades de los clusters
cluster.markers <- GetAssayData(data_int, slot = "data")
cluster.ids <- Idents(data_int)

# Cargar los archivos CSV
median_data <- fread("rawdata/medians.csv")
trimmed_means_data <- fread("rawdata/trimmed_means.csv")

# Verificar las primeras filas
head(median_data)
head(trimmed_means_data)

# Revisar dimensiones
dim(median_data)
dim(trimmed_means_data)

# Convertir a matriz
median_matrix <- as.matrix(median_data[ , -1, with = FALSE])  # Asumiendo que la primera columna son los nombres de genes
rownames(median_matrix) <- median_data[[1]]  # Asignar nombres de genes a las filas

trimmed_means_matrix <- as.matrix(trimmed_means_data[ , -1, with = FALSE])  # Similar al anterior
rownames(trimmed_means_matrix) <- trimmed_means_data[[1]]

# Crear referencia de SingleR
# Crear un nuevo objeto SummarizedExperiment con logcounts
reference_median <- SummarizedExperiment(
    assays = list(logcounts = log1p(median_matrix)), # Aplicar log-transformation
    rowData = DataFrame(gene = rownames(median_matrix)),
    colData = DataFrame(label.main = colnames(median_matrix))
)
# Revisar las etiquetas
colData(reference_median)

# Confirmar que `label.main` contiene los nombres de los tipos celulares
reference_median$label.main <- colnames(median_matrix)

reference_trimmed <- SummarizedExperiment(
    assays = list(logcounts = log1p(trimmed_means_matrix)), # Aplicar log-transformation
    rowData = DataFrame(gene = rownames(trimmed_means_matrix)),
    colData = DataFrame(label.main = colnames(trimmed_means_matrix))
)
# Revisar las etiquetas
colData(reference_trimmed)

# Confirmar que `label.main` contiene los nombres de los tipos celulares
reference_trimmed$label.main <- colnames(trimmed_means_matrix)
```


```{r}
# Comparar con la base de datos de referencia
annotations_median <- SingleR(test = cluster.markers, 
                       ref = reference_median, 
                       labels = reference_median$label.main,
                       clusters = cluster.ids)


# Revisar resultados
head(annotations_median)
```


```{r}
# Comparar con la base de datos de referencia
annotations_trimmed <- SingleR(test = cluster.markers, 
                       ref = reference_trimmed, 
                       labels = reference_trimmed$label.main,
                       clusters = cluster.ids)


# Revisar resultados
head(annotations_trimmed)
```

## Plot Median
```{r}
new.cluster.median <- annotations_median$labels
names(new.cluster.median) <- levels(data_int)
data_int_median <- RenameIdents(data_int, new.cluster.median)
```


```{r}
# Visualiza con las nuevas identidades
plot <- DimPlot(data_int_median, reduction = "umap", label = TRUE, repel = TRUE)
print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "UMAP_simple_cluster_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable

```

## Plot Trimmed
```{r}
new.cluster.trimmed <- annotations_trimmed$labels
names(new.cluster.trimmed) <- levels(data_int)
data_int_trimmed <- RenameIdents(data_int, new.cluster.median)
```


```{r}
# Visualiza con las nuevas identidades
plot <- DimPlot(data_int_trimmed, reduction = "umap", label = TRUE, repel = TRUE)
print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "UMAP_simple_cluster_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable

```


## MouseRNAseqData
```{r}


ref <- MouseRNAseqData()  # Change according needs
# Extraer la matriz de expresión y las identidades de los clusters
cluster.markers <- GetAssayData(data_int, slot = "data")
cluster.ids <- Idents(data_int)

# Comparar con la base de datos de referencia
annotations_ms <- SingleR(test = cluster.markers, 
                       ref = ref, 
                       labels = ref$label.main,
                       clusters = cluster.ids)

# Ver los resultados
head(annotations)
```


```{r}
new.cluster.ids <- annotations_ms$labels
names(new.cluster.ids) <- levels(data_int)
data_int_ms <- RenameIdents(data_int, new.cluster.ids)
```


```{r}
# Visualiza con las nuevas identidades
plot <- DimPlot(data_int_ms, reduction = "umap", label = TRUE, repel = TRUE)
print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "UMAP_simple_cluster_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable

```
