

# 0. Loading Libraries

# Set up global parameters
```{r}
rm(list=ls()) #Clean enviroment
gc() #Clean memory

# Set up parameters
image_number <- 1
output_path <- "output/SC_UBC/Integrated/"
set.seed(123)
options(scipen = -1)  # Habilitar notación científica en R
```

## Loading functions
```{r}
source("code/global_variables.R")
source("code/00_packages.R")
source("code/01_sc_functions.R")
```
# 1. Importing data
We start by reading in the data. The Read10X() function reads in the output of the cellranger pipeline from 10X, returning a unique molecular identified (UMI) count matrix. The values in this matrix represent the number of molecules for each feature (i.e. gene; row) that are detected in each cell (column). Note that more recent versions of cellranger now also output using the h5 file format, which can be read in using the Read10X_h5() function in Seurat.
```{r}
tic()
#import data
# Load the dataset. Select the folder for every library.

c1 <- Seurat::Read10X(data.dir = "rawdata/SC/AWGC-1_Ctrl-1_results")
c2 <- Seurat::Read10X(data.dir = "rawdata/SC/AWGC-2_Ctrl-2_results")
ko1 <- Seurat::Read10X(data.dir = "rawdata/SC/AWGC-3_KOut-1_results")
ko2 <- Seurat::Read10X(data.dir = "rawdata/SC/AWGC-4_KOut-2_results")

# Initialize the Seurat object with the raw (non-normalized data).
c1 <- CreateSeuratObject(counts = c1, project = "wt_1", min.cells = 3, min.features = 200)
c2 <- CreateSeuratObject(counts = c2, project = "wt_2", min.cells = 3, min.features = 200)
ko1 <- CreateSeuratObject(counts = ko1, project = "ko_1", min.cells = 3, min.features = 200)
ko2 <- CreateSeuratObject(counts = ko2, project = "ko_2", min.cells = 3, min.features = 200)

c1$condition <- "WT"
c2$condition <- "WT"
ko1$condition <- "KO"
ko2$condition <- "KO"



#Put all SeuratObjects in a vector and in a list for automatization
#seurat_names <- c("ubc_c1", "ubc_c2", "ubc_ko1", "ubc_ko2")
# Create a named list of Seurat objects
seurat_objects <- list(
  c1 = c1,
  c2 = c2,
  ko1 = ko1,
  ko2 = ko2
)
# Remove the individual objects from the global environment
rm(c1, c2, ko1, ko2)

# Optionally, run garbage collection to free up memory
gc()

```

```{r}
tic()
# Export results ####
  save(seurat_objects, file = paste0(output_path, "RData/", "seurat_objects", ".RData"))  # Save using the name

toc()

load(paste0(output_path,"RData/seurat_objects.RData"))

```
# 2. Data preprocessing for each library

Preprocessing taken from Seurat web page and 
https://hbctraining.github.io/scRNA-seq/lessons/04_SC_quality_control.html
 
## QC and selecting cells for further analysis
Seurat allows you to easily explore QC metrics and filter cells based on any user-defined criteria. A few QC metrics commonly used by the community include
## Identify mitochondrial genes
The number of unique genes detected in each cell.
Low-quality cells or empty droplets will often have very few genes
Cell doublets or multiplets may exhibit an aberrantly high gene count
Similarly, the total number of molecules detected within a cell (correlates strongly with unique genes)
The percentage of reads that map to the mitochondrial genome
Low-quality / dying cells often exhibit extensive mitochondrial contamination
We calculate mitochondrial QC metrics with the 'PercentageFeatureSet()' function, which calculates the percentage of counts originating from a set of features
We use the set of all genes starting with 'mt-' as a set of mitochondrial genes

```{r}
# Add QC stats to Seurat objects in the list
seurat_objects <- lapply(names(seurat_objects), function(name) {
  obj <- seurat_objects[[name]]  # Access the object from the list
  # Add percentage of mitochondrial genes
  obj$percent.mt <- PercentageFeatureSet(obj, pattern = "^mt-")  
  # Add number of genes per UMI for each cell to metadata
  obj$log10GenesPerUMI <- log10(obj$nFeature_RNA) / log10(obj$nCount_RNA)
  # Add percentage of ribosomal genes
  obj$percent.ribosomal <- PercentageFeatureSet(obj,pattern="^Rp[ls]") 
  obj  # Return the modified object
}) %>% setNames(names(seurat_objects))  # Preserve the original names in the list
```

## QC cell Pre-Filter


```{r}
# Combine metadata from all Seurat objects into a single dataframe
combined_meta <- bind_rows(
  seurat_objects$c1@meta.data %>% mutate(library = "c1"),
  seurat_objects$c2@meta.data %>% mutate(library = "c2"),
  seurat_objects$ko1@meta.data %>% mutate(library = "ko1"),
  seurat_objects$ko2@meta.data %>% mutate(library = "ko2")
)

```


### Number of cells per sample

```{r}
# Visualize the number of cells per library
combined_plot <- combined_meta %>%
  ggplot(aes(x = library, fill = condition)) +
  geom_bar() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("Number of Cells per Library") +
  xlab("Library") +
  ylab("Number of Cells")

print(combined_plot)

  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Number_of_Cells")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")
  
  image_number <- image_number+1
  
```

### UMI counts (transcripts) per cell
The UMI counts per cell (nCount_RNA) should generally be above 500, that is the low end of what we expect. If UMI counts are between 500-1000 counts, it is usable but the cells probably should have been sequenced more deeply.



```{r}
# Visualize the number UMIs/transcripts per cell
# Visualize the number of cells per library
combined_plot <- combined_meta %>%
  ggplot(aes(x = nCount_RNA, color=library, fill = library)) +
  geom_density(alpha = 0.2) + 
  scale_x_log10() + 
  theme_classic() +
  ylab("UMI per cell") +
  geom_vline(xintercept = 500) 


print(combined_plot)

  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_UMI_per_cell")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")

    image_number <- image_number+1
```
We can see that majority of our cells in both samples have 1000 UMIs or greater, which is great.


### Genes detected per cell
We have similar expectations for gene detection as for UMI detection, although it may be a bit lower than UMIs. For high quality data, the proportional histogram should contain a single large peak that represents cells that were encapsulated. If we see a small shoulder to the right of the major peak (not present in our data), or a bimodal distribution of the cells, that can indicate a couple of things. It might be that there are a set of cells that failed for some reason. It could also be that there are biologically different types of cells (i.e. quiescent cell populations, less complex cells of interest), and/or one type is much smaller than the other (i.e. cells with high counts may be cells that are larger in size). Therefore, this threshold should be assessed with other metrics that we describe in this lesson.

```{r}
# Visualize the distribution of genes detected per cell via histogram
combined_plot <- combined_meta %>% 
  	ggplot(aes(color=library, x=nFeature_RNA, fill= library)) + 
  	geom_density(alpha = 0.2) + 
  	theme_classic() +
  	scale_x_log10() + 
  	geom_vline(xintercept = 300)

print(combined_plot)

  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Genes_per_cell_denstity")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")
  image_number <- image_number+1
# Visualize the distribution of genes detected per cell via boxplot
combined_plot <- combined_meta %>% 
  	ggplot(aes(x=library, y=log10(nFeature_RNA), fill=library)) + 
  	geom_boxplot() + 
  	theme_classic() +
  	theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  	theme(plot.title = element_text(hjust=0.5, face="bold")) +
  	ggtitle("NCells vs NGenes")

print(combined_plot)

  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Genes_per_cell_boxplot")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")
  image_number <- image_number+1
```
### UMIs vs. genes detected
Two metrics that are often evaluated together are the number of UMIs and the number of genes detected per cell. Here, we have plotted the number of genes versus the numnber of UMIs coloured by the fraction of mitochondrial reads. Mitochondrial read fractions are only high (light blue color) in particularly low count cells with few detected genes. This could be indicative of damaged/dying cells whose cytoplasmic mRNA has leaked out through a broken membrane, and thus, only mRNA located in the mitochondria is still conserved. These cells are filtered out by our count and gene number thresholds. Jointly visualizing the count and gene thresholds shows the joint filtering effect.

Cells that are poor quality are likely to have low genes and UMIs per cell, and correspond to the data points in the bottom left quadrant of the plot. Good cells will generally exhibit both higher number of genes per cell and higher numbers of UMIs.

With this plot we also evaluate the slope of the line, and any scatter of data points in the bottom right hand quadrant of the plot. These cells have a high number of UMIs but only a few number of genes. These could be dying cells, but also could represent a population of a low complexity celltype (i.e red blood cells).

```{r}
# Visualize the correlation between genes detected and number of UMIs and determine whether strong presence of cells with low numbers of genes/UMIs
combined_plot <- combined_meta %>% 
  	ggplot(aes(x=nCount_RNA, y=nFeature_RNA, color=percent.mt)) + 
  	geom_point() + 
	scale_colour_gradient(low = "gray90", high = "black") +
  	stat_smooth(method=lm) +
  	scale_x_log10() + 
  	scale_y_log10() + 
  	theme_classic() +
  	geom_vline(xintercept = 500) +
  	geom_hline(yintercept = 250) +
  	facet_wrap(~library)


print(combined_plot)

  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Genes_vs_Cell")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")
  image_number <- image_number+1
```

### Mitochondrial counts ratio
This metric can identify whether there is a large amount of mitochondrial contamination from dead or dying cells. We define poor quality samples for mitochondrial counts as cells which surpass the 0.2 mitochondrial ratio mark, unless of course you are expecting this in your sample.


```{r}
# Visualize the distribution of mitochondrial gene expression detected per cell
combined_plot <- combined_meta %>% 
  	ggplot(aes(color=library, x=percent.mt, fill=library)) + 
  	geom_density(alpha = 0.2) + 
  	scale_x_log10() + 
  	theme_classic() +
  	geom_vline(xintercept = 1) #Threshold of 1% of mitochondrial genes

print(combined_plot)

  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Density_MitoGenes_Cell")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")
  image_number <- image_number+1

```
  	
###	Complexity
We can see the samples where we sequenced each cell less have a higher overall complexity, that is because we have not started saturating the sequencing for any given gene for these samples. Outlier cells in these samples might be cells that have a less complex RNA species than other cells. Sometimes we can detect contamination with low complexity cell types like red blood cells via this metric. Generally, we expect the novelty score to be above 0.80.

```{r}
# Visualize the overall complexity of the gene expression by visualizing the genes detected per UMI
combined_plot <- combined_meta %>%
  	ggplot(aes(x=log10GenesPerUMI, color = library, fill=library)) +
  	geom_density(alpha = 0.2) +
  	theme_classic() +
  	geom_vline(xintercept = 0.8)

print(combined_plot)
  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Complexity")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")
  image_number <- image_number+1

```

  	
### Percentage of mitochondrial genes

```{r}
source("code/01_sc_functions.R")
# Load the function from the R script
#source("01_qc_functions.R")

# Generate QC plots for objects in the list
lapply(names(seurat_objects), function(name) {
  obj <- seurat_objects[[name]]  # Access the object from the list
  image_number <<- generate_qc_plots(obj, name, image_number, output_path, suffix = "prefilter")
})
```


## QC cell Filter
We filter cells that have unique feature counts over 7,500 or less than 850
We filter cells that have >1% mitochondrial counts

```{r}

# Filter Seurat objects based on QC metrics
seurat_objects <- lapply(names(seurat_objects), function(name) {
  obj <- seurat_objects[[name]]  # Access the object from the list
  obj <- subset(obj, subset = nFeature_RNA > 850 & nFeature_RNA < 7500 & percent.mt < 1)  # Apply QC filters
  obj  # Return the filtered object
}) %>% setNames(names(seurat_objects))  # Preserve the original names in the list

```






```{r}
# Combine metadata from all Seurat objects into a single dataframe
combined_meta_filter <- bind_rows(
  seurat_objects$c1@meta.data %>% mutate(library = "c1"),
  seurat_objects$c2@meta.data %>% mutate(library = "c2"),
  seurat_objects$ko1@meta.data %>% mutate(library = "ko1"),
  seurat_objects$ko2@meta.data %>% mutate(library = "ko2")
)

```


### Number of cells per sample

```{r}
# Visualize the number of cells per library
combined_plot <- combined_meta_filter %>%
  ggplot(aes(x = library, fill = condition)) +
  geom_bar() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("Number of Cells per Library") +
  xlab("Library") +
  ylab("Number of Cells")

print(combined_plot)

  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Number_of_Cells_filter")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")
  
  image_number <- image_number+1
  
```

### UMI counts (transcripts) per cell
The UMI counts per cell (nCount_RNA) should generally be above 500, that is the low end of what we expect. If UMI counts are between 500-1000 counts, it is usable but the cells probably should have been sequenced more deeply.



```{r}
# Visualize the number UMIs/transcripts per cell
# Visualize the number of cells per library
combined_plot <- combined_meta_filter %>%
  ggplot(aes(x = nCount_RNA, color=library, fill = library)) +
  geom_density(alpha = 0.2) + 
  scale_x_log10() + 
  theme_classic() +
  ylab("UMI per cell") +
  geom_vline(xintercept = 500) 


print(combined_plot)

  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_UMI_per_cell_filter")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")

    image_number <- image_number+1
```
We can see that majority of our cells in both samples have 1000 UMIs or greater, which is great.


### Genes detected per cell
We have similar expectations for gene detection as for UMI detection, although it may be a bit lower than UMIs. For high quality data, the proportional histogram should contain a single large peak that represents cells that were encapsulated. If we see a small shoulder to the right of the major peak (not present in our data), or a bimodal distribution of the cells, that can indicate a couple of things. It might be that there are a set of cells that failed for some reason. It could also be that there are biologically different types of cells (i.e. quiescent cell populations, less complex cells of interest), and/or one type is much smaller than the other (i.e. cells with high counts may be cells that are larger in size). Therefore, this threshold should be assessed with other metrics that we describe in this lesson.

```{r}
# Visualize the distribution of genes detected per cell via histogram
combined_plot <- combined_meta_filter %>% 
  	ggplot(aes(color=library, x=nFeature_RNA, fill= library)) + 
  	geom_density(alpha = 0.2) + 
  	theme_classic() +
  	scale_x_log10() + 
  	geom_vline(xintercept = 300)

print(combined_plot)

  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Genes_per_cell_denstity_filter")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")
  image_number <- image_number+1
# Visualize the distribution of genes detected per cell via boxplot
combined_plot <- combined_meta_filter %>% 
  	ggplot(aes(x=library, y=log10(nFeature_RNA), fill=library)) + 
  	geom_boxplot() + 
  	theme_classic() +
  	theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  	theme(plot.title = element_text(hjust=0.5, face="bold")) +
  	ggtitle("NCells vs NGenes")

print(combined_plot)

  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Genes_per_cell_boxplot_filter")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")
  image_number <- image_number+1
```
### UMIs vs. genes detected
Two metrics that are often evaluated together are the number of UMIs and the number of genes detected per cell. Here, we have plotted the number of genes versus the numnber of UMIs coloured by the fraction of mitochondrial reads. Mitochondrial read fractions are only high (light blue color) in particularly low count cells with few detected genes. This could be indicative of damaged/dying cells whose cytoplasmic mRNA has leaked out through a broken membrane, and thus, only mRNA located in the mitochondria is still conserved. These cells are filtered out by our count and gene number thresholds. Jointly visualizing the count and gene thresholds shows the joint filtering effect.

Cells that are poor quality are likely to have low genes and UMIs per cell, and correspond to the data points in the bottom left quadrant of the plot. Good cells will generally exhibit both higher number of genes per cell and higher numbers of UMIs.

With this plot we also evaluate the slope of the line, and any scatter of data points in the bottom right hand quadrant of the plot. These cells have a high number of UMIs but only a few number of genes. These could be dying cells, but also could represent a population of a low complexity celltype (i.e red blood cells).

```{r}
# Visualize the correlation between genes detected and number of UMIs and determine whether strong presence of cells with low numbers of genes/UMIs
combined_plot <- combined_meta_filter %>% 
  	ggplot(aes(x=nCount_RNA, y=nFeature_RNA, color=percent.mt)) + 
  	geom_point() + 
	scale_colour_gradient(low = "gray90", high = "black") +
  	stat_smooth(method=lm) +
  	scale_x_log10() + 
  	scale_y_log10() + 
  	theme_classic() +
  	geom_vline(xintercept = 500) +
  	geom_hline(yintercept = 250) +
  	facet_wrap(~library)


print(combined_plot)

  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Genes_vs_Cell_filter")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")
  image_number <- image_number+1
```


  	
###	Complexity
We can see the samples where we sequenced each cell less have a higher overall complexity, that is because we have not started saturating the sequencing for any given gene for these samples. Outlier cells in these samples might be cells that have a less complex RNA species than other cells. Sometimes we can detect contamination with low complexity cell types like red blood cells via this metric. Generally, we expect the novelty score to be above 0.80.

```{r}
# Visualize the overall complexity of the gene expression by visualizing the genes detected per UMI
combined_plot <- combined_meta_filter %>%
  	ggplot(aes(x=log10GenesPerUMI, color = library, fill=library)) +
  	geom_density(alpha = 0.2) +
  	theme_classic() +
  	geom_vline(xintercept = 0.8)

print(combined_plot)
  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Complexity_filter")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")
  image_number <- image_number+1

```
### Mitochondrial counts ratio
This metric can identify whether there is a large amount of mitochondrial contamination from dead or dying cells. We define poor quality samples for mitochondrial counts as cells which surpass the 0.2 mitochondrial ratio mark, unless of course you are expecting this in your sample.


```{r}
# Visualize the distribution of mitochondrial gene expression detected per cell
combined_plot <- combined_meta_filter %>% 
  	ggplot(aes(color=library, x=percent.mt, fill=library)) + 
  	geom_density(alpha = 0.2) + 
  	scale_x_log10() + 
  	theme_classic() +
  	geom_vline(xintercept = 1) #Threshold of 1% of mitochondrial genes

print(combined_plot)

  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Density_MitoGenes_Cell_filter")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")
  image_number <- image_number+1

```
  	
  	
### Percentage of mitochondrial genes

```{r}
# Generate QC plots for objects in the list
lapply(names(seurat_objects), function(name) {
  obj <- seurat_objects[[name]]  # Access the object from the list
  image_number <<- generate_qc_plots(obj, name, image_number, output_path, suffix = "filter")
})
```

# 3.Normalizing the data

```{r}
  # Plan for parallelization (adjust workers based on machine resources)
  plan(multicore, workers = 4)  # Adjust based on your system
  options(future.globals.maxSize = 8 * 1024^3) #(Gb)
  seurat_names <- names(seurat_objects)
# Process Seurat objects
norm_seurat_objects <- normalize_seurat_objects(seurat_objects = seurat_objects)


```


```{r}
tic()
# Export results ####
  save(norm_seurat_objects, file = paste0(output_path, "RData/", "Norm_seurat_objects", ".RData"))  # Save using the name
toc()
```
## Highly Variable genes plot
```{r}
# Generate HV plots for objects in the list
lapply(names(norm_seurat_objects), function(name) {
  obj <- norm_seurat_objects[[name]]  # Access the object from the list
  image_number <<- generate_HV_plots(obj, name, image_number, output_path)
})

```
# 4.Linear dimensional reduction
Next we perform PCA on the scaled data. By default, only the previously determined variable features are used as input, but can be defined using features argument if you wish to choose a different subset (if you do want to use a custom subset of features, make sure you pass these to ScaleData first).

For the first principal components, Seurat outputs a list of genes with the most positive and negative loadings, representing modules of genes that exhibit either correlation (or anti-correlation) across single-cells in the dataset.

```{r}


norm_seurat_objects <- add_pca(
  norm_seurat_objects = norm_seurat_objects,
  output_path = output_path,
  image_number = image_number
)

```

## Elbow Plot
Determine the ‘dimensionality’ of the dataset
To overcome the extensive technical noise in any single feature for scRNA-seq data, Seurat clusters cells based on their PCA scores, with each PC essentially representing a ‘metafeature’ that combines information across a correlated feature set. The top principal components therefore represent a robust compression of the dataset. However, how many components should we choose to include? 10? 20? 100?

In Macosko et al, we implemented a resampling test inspired by the JackStraw procedure. While still available in Seurat (see previous vignette), this is a slow and computationally expensive procedure, and we is no longer routinely used in single cell analysis.

An alternative heuristic method generates an ‘Elbow plot’: a ranking of principle components based on the percentage of variance explained by each one (ElbowPlot() function). In this example, we can observe an ‘elbow’ around PC9-10, suggesting that the majority of true signal is captured in the first 10 PCs.

```{r}

elbow_plots(
  seurat_objects = norm_seurat_objects,
  output_path = output_path,
  image_number = image_number
)


```

## Saving Results
```{r}
tic()
# Export results ####
  save(norm_seurat_objects, file = paste0(output_path, "RData/", "Norm_seurat_objects", ".RData"))  # Save using the name
toc()
```
### Load RDatas
```{r}
tic()
# Load results ####
load(paste0(output_path,"RData/Norm_seurat_objects.RData"))
toc()


```


## Cluster cells
Seurat applies a graph-based clustering approach, building upon initial strategies in (Macosko et al). Importantly, the distance metric which drives the clustering analysis (based on previously identified PCs) remains the same. However, our approach to partitioning the cellular distance matrix into clusters has dramatically improved. Our approach was heavily inspired by recent manuscripts which applied graph-based clustering approaches to scRNA-seq data [SNN-Cliq, Xu and Su, Bioinformatics, 2015] and CyTOF data [PhenoGraph, Levine et al., Cell, 2015]. Briefly, these methods embed cells in a graph structure - for example a K-nearest neighbor (KNN) graph, with edges drawn between cells with similar feature expression patterns, and then attempt to partition this graph into highly interconnected ‘quasi-cliques’ or ‘communities’.

As in PhenoGraph, we first construct a KNN graph based on the euclidean distance in PCA space, and refine the edge weights between any two cells based on the shared overlap in their local neighborhoods (Jaccard similarity). This step is performed using the FindNeighbors() function, and takes as input the previously defined dimensionality of the dataset (first 11 PCs).

To cluster the cells, we next apply modularity optimization techniques such as the Louvain algorithm (default) or SLM [SLM, Blondel et al., Journal of Statistical Mechanics], to iteratively group cells together, with the goal of optimizing the standard modularity function. The FindClusters() function implements this procedure, and contains a resolution parameter that sets the ‘granularity’ of the downstream clustering, with increased values leading to a greater number of clusters. We find that setting this parameter between 0.4-1.2 typically returns good results for single-cell datasets of around 3K cells. Optimal resolution often increases for larger datasets. The clusters can be found using the Idents() function.

```{r}

  plan(multicore, workers = 4)  # Adjust based on your system
  options(future.globals.maxSize = 24 * 1024^3) #(Gb)
  gc()
seurat_names <- names(norm_seurat_objects)
# Perform clustering and UMAP on Seurat objects
clustered_seurat_objects <- cluster_and_umap(
  seurat_objects = norm_seurat_objects,  # Output from the PCA step
  output_path = output_path,
  image_number = image_number,
  dims = 1:10,  # Customize dimensions if needed
  resolution = 0.8,  # Customize resolution if needed
  components = 2L
  )
```


```{r}
#Save Normalized and dimentionally reduced data
tic()
# Export results ####
# Save using the name
  save(clustered_seurat_objects, file = paste0(output_path, "RData/", "clustered_seurat_objects", ".RData"))  

### Load RDatas

# Load results ####
load(paste0(output_path,"RData/clustered_seurat_objects.RData"))
toc()

rm(norm_seurat_objects, seurat_objects, combined_meta, combined_meta_filter, combined_plot)
gc()
```


# 4. Library integration

```{r}
tic()
# Plan for parallelization (adjust workers based on machine resources)
  plan(multicore, workers = 4)  # Adjust based on your system
  options(future.globals.maxSize = 16 * 1024^3) #(Gb)
  gc()
  
# Integrate data and perform clustering/UMAP in parallel
data_int <- integrate_and_cluster_parallel(
  seurat_objects = clustered_seurat_objects,
  seurat_names = seurat_names,
  output_path = output_path,
  image_number = image_number)

toc()
```
## Save data_int object

```{r}
#Save Normalized and dimentionally reduced data
tic()
# Export results ####
# Save using the name
  save(data_int, file = paste0(output_path, "RData/", "data_int", ".RData"))  

### Load RDatas

# Load results ####
load(paste0(output_path,"RData/data_int.RData"))
toc()

rm(clustered_seurat_objects)
gc()
```


# 5.Plot UMAP & UMAP Features

```{r}
tic()
 # note that you can set `label = TRUE` or use the LabelClusters function to help label individual clusters
  plot <- DimPlot(data_int, reduction = "umap", label = TRUE)

  print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_UMAP")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 8, height = 6, units = "in")
  # Increment the figure counter
  image_number <<- image_number + 1  # Use <<- to update the external variable
    


    plot <- DimPlot(data_int, reduction = "umap", split.by="condition", ncol = 2, label = TRUE)
  
  print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_UMAP_ident")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 8, height = 6, units = "in")
  # Increment the figure counter
  image_number <<- image_number + 1  # Use <<- to update the external variable
  
  
  #UMAP by features
  features <- c("nFeature_RNA", "nCount_RNA")
  feature_plot <- FeaturePlot(data_int, features = features)
    print(feature_plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Feature_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), feature_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), feature_plot, width = 8, height = 6, units = "in")
  # Increment the figure counter
  image_number <<- image_number + 1  # Use <<- to update the external variable
toc()

```

## UMAP HVGs

```{r}
top <-head(VariableFeatures(data_int),10)

  #UMAP by features
  feature_plot <- FeaturePlot(data_int, features = top, ncol = 5)
    print(feature_plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Feature_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), feature_plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), feature_plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  image_number <<- image_number + 1  # Use <<- to update the external variable
  
```

## UMAP Dnajc5

```{r}
  #UMAP by features
  feature_plot <- FeaturePlot(data_int, features = "Dnajc5",split.by="condition", ncol = 2 )
    print(feature_plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Feature_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), feature_plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), feature_plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  image_number <<- image_number + 1  # Use <<- to update the external variable
  
```



# 6. Finding differentially expressed features (cluster biomarkers)
Seurat can help you find markers that define clusters via differential
expression (DE). By default, it identifies positive and negative markers
of a single cluster (specified in ident.1), compared to all other cells.
FindAllMarkers() automates this process for all clusters, but you can
also test groups of clusters vs. each other, or against all cells.

In Seurat v5, we use the presto package (as described here and available
for installation here), to dramatically improve the speed of DE
analysis, particularly for large datasets. For users who are not using
presto, you can examine the documentation for this function
(?FindMarkers) to explore the min.pct and logfc.threshold parameters,
which can be increased in order to increase the speed of DE testing.
```{r}
# Ensure default assay is set to 'integrated'
DefaultAssay(data_int) <- "integrated"

Idents(data_int) <- "seurat_clusters"

# Crear una lista para guardar los resultados
cluster_list <- list()

# Iterar sobre los clústeres
for (cluster in unique(Idents(data_int))) {
  # Filtrar por clúster
  cells_in_cluster <- WhichCells(data_int, idents = cluster)
  
  # Subconjunto de células del clúster actual
  cluster_subset <- subset(data_int, cells = cells_in_cluster)
  
  # Cambiar las identidades al factor "condition"
  Idents(cluster_subset) <- "condition"
  
  # Realizar análisis diferencial entre KO y WT
  markers <- FindMarkers(cluster_subset, 
                         ident.1 = "KO", 
                         ident.2 = "WT",
                         test.use = "wilcox_limma", # Cambia si prefieres otro test
                         logfc.threshold = 0.25, 
                         min.pct = 0.1)
  
  # Guardar resultados
  cluster_list[[paste0("Cluster_", cluster)]] <- markers
}

```




```{r}
# Convertir la lista en un data.frame con una columna adicional 'cluster' (valor numérico)
cluster_markers_df <- do.call(rbind, lapply(names(cluster_list), function(cluster) {
  # Extraer el dataframe del cluster actual y añadir una columna con el número del cluster
  markers <- cluster_list[[cluster]]
  markers$cluster <- as.numeric(gsub("Cluster_", "", cluster)) # Eliminar el prefijo y convertir a numérico
  return(markers)
}))
# Convertir las filas (rownames) en una columna llamada 'gene'
cluster_markers_df$gene <- rownames(cluster_markers_df)

# Reiniciar los nombres de las filas
rownames(cluster_markers_df) <- NULL


cluster_markers_df <- cluster_markers_df %>% 
  dplyr::select(gene, everything())

# Ver el resultado
head(cluster_markers_df)


```




```{r}


VolcanoEnhanced <- EnhancedVolcano::EnhancedVolcano(cluster_markers_df,
                                  lab = cluster_markers_df$gene,
                                  x = 'avg_log2FC',
                                  y = 'p_val_adj',
                                  title = 'WT vs KO Differential Genes',
                                  pCutoff = 1e-16)



```


```{r}
#Generate a list 

cluster_list %>% 
write_xlsx(paste0(output_path,"tables/","cluster_list.xlsx"))
```

## HeatMap

```{r}
cluster_markers_df %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > FC) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10


plot <- DoHeatmap(data_int, features = top10$gene) 
print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_HeatMap_cluster_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  image_number <<- image_number + 1  # Use <<- to update the external variable
```


# 8. Saving


```{r}
#Save Normalized and dimentionally reduced data
tic()
# Export results ####
# Save using the name
  save(data_int, file = paste0(output_path, "RData/", "data_int", ".RData"))  

### Load RDatas

# Load results ####
load(paste0(output_path,"RData/data_int.RData"))
toc()

gc()
```

# 7. Clustering association

1.  Use 'SingleR' is a widely used tool that compares gene expression
    profiles of your clusters to predefined reference databases. Load an
    appropriate reference database. For example:

-   MouseRNAseqData: Reference for murine tissues.
-   ImmGenData: Immunological reference.
-   HumanPrimaryCellAtlasData: For human cells.

2.  Use scType for fast annotation 'scType' is a simpler, more
    customizable tool that uses user-defined or database-available
    marker lists.

    Comparison between 'SingleR' and 'scType'
    
    | Feature           | SingleR                         | scType            |
    |-------------------|---------------------------------|-------------------|
    | **Ease of use**   | More robust but more technical  | Simpler           |
    | **Database**      | Use predefined references       | Use marker lists  |
    | **Customization** | Limited                         | Very customizable |
    | **Speed**         | Moderate                        | Very fast         |

    : Recommendation If you have a well-characterized dataset (such as
    mouse brain transcriptomics), use SingleR with a database such as
    MouseRNAseqData. If you prefer greater flexibility or speed, try
    scType with specific marker lists.


## Allen Brain Atlas clustering
Download matrix Gene expression matrix and Table cell metadata from Allen Brain Atlas Cell Types Database: RNA-Seq Data (https://portal.brain-map.org/atlases-and-data/rnaseq)
Whole Cortex & Hippocampus - 10x genomics (2020)
with 10x-SMART-seq taxonomy (2021)
```{r}

# Extraer la matriz de expresión y las identidades de los clusters
cluster.markers <- GetAssayData(data_int, slot = "data")
cluster.ids <- Idents(data_int)

# Cargar los archivos CSV
median_data <- fread("rawdata/medians.csv")
trimmed_means_data <- fread("rawdata/trimmed_means.csv")

# Verificar las primeras filas
head(median_data)
head(trimmed_means_data)

# Revisar dimensiones
dim(median_data)
dim(trimmed_means_data)

# Convertir a matriz
median_matrix <- as.matrix(median_data[ , -1, with = FALSE])  # Asumiendo que la primera columna son los nombres de genes
rownames(median_matrix) <- median_data[[1]]  # Asignar nombres de genes a las filas

trimmed_means_matrix <- as.matrix(trimmed_means_data[ , -1, with = FALSE])  # Similar al anterior
rownames(trimmed_means_matrix) <- trimmed_means_data[[1]]

# Crear referencia de SingleR
# Crear un nuevo objeto SummarizedExperiment con logcounts
reference_median <- SummarizedExperiment(
    assays = list(logcounts = log1p(median_matrix)), # Aplicar log-transformation
    rowData = DataFrame(gene = rownames(median_matrix)),
    colData = DataFrame(label.main = colnames(median_matrix))
)
# Revisar las etiquetas
colData(reference_median)

# Confirmar que `label.main` contiene los nombres de los tipos celulares
reference_median$label.main <- colnames(median_matrix)

reference_trimmed <- SummarizedExperiment(
    assays = list(logcounts = log1p(trimmed_means_matrix)), # Aplicar log-transformation
    rowData = DataFrame(gene = rownames(trimmed_means_matrix)),
    colData = DataFrame(label.main = colnames(trimmed_means_matrix))
)
# Revisar las etiquetas
colData(reference_trimmed)

# Confirmar que `label.main` contiene los nombres de los tipos celulares
reference_trimmed$label.main <- colnames(trimmed_means_matrix)
```

```{r}
# Comparar con la base de datos de referencia
annotations_median <- SingleR(test = cluster.markers, 
                       ref = reference_median, 
                       labels = reference_median$label.main,
                       clusters = cluster.ids)


# Revisar resultados
head(annotations_median)


```


```{r}
# Comparar con la base de datos de referencia
annotations_trimmed <- SingleR(test = cluster.markers, 
                       ref = reference_trimmed, 
                       labels = reference_trimmed$label.main,
                       clusters = cluster.ids)


# Revisar resultados
head(annotations_trimmed)
```


## Plot Median
```{r}
new.cluster.median <- annotations_median$labels
names(new.cluster.median) <- levels(data_int)
data_int_median <- RenameIdents(data_int, new.cluster.median)
```


```{r}
# Visualiza con las nuevas identidades
plot <- DimPlot(data_int_median, reduction = "umap", label = TRUE, repel = TRUE)
print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_UMAP_median_cluster_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  image_number <<- image_number + 1  # Use <<- to update the external variable

```

## Plot Trimmed
```{r}
new.cluster.trimmed <- annotations_trimmed$labels
names(new.cluster.trimmed) <- levels(data_int)
data_int_trimmed <- RenameIdents(data_int, new.cluster.trimmed)
```


```{r}
# Visualiza con las nuevas identidades
plot <- DimPlot(data_int_trimmed, reduction = "umap", label = TRUE, repel = TRUE)
print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_UMAP_trimmed_cluster_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  image_number <<- image_number + 1  # Use <<- to update the external variable

```


## MouseRNAseqData
```{r}


ref <- MouseRNAseqData()  # Change according needs
# Extraer la matriz de expresión y las identidades de los clusters
cluster.markers <- GetAssayData(data_int, slot = "data")
cluster.ids <- Idents(data_int)

# Comparar con la base de datos de referencia
annotations_ms <- SingleR(test = cluster.markers, 
                       ref = ref, 
                       labels = ref$label.main,
                       clusters = cluster.ids)

# Ver los resultados
head(annotations_ms)
```


```{r}
new.cluster.ids <- annotations_ms$labels
names(new.cluster.ids) <- levels(data_int)
data_int_ms <- RenameIdents(data_int, new.cluster.ids)
```


```{r}
# Visualiza con las nuevas identidades
plot <- DimPlot(data_int_ms, reduction = "umap", label = TRUE, repel = TRUE)
print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_UMAP_simple_cluster_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  image_number <<- image_number + 1  # Use <<- to update the external variable

```
## Save Clusters

```{r}
# Convertir las anotaciones a un data.frame para unirlas con cluster_markers_df
annotations_median_df <- as.data.frame(annotations_median)
annotations_median_df$cluster <- rownames(annotations_median_df)  # Añadir los IDs de los clusters como columna
# Convertir las anotaciones a un data.frame para unirlas con cluster_markers_df
annotations_trimmed_df <- as.data.frame(annotations_trimmed)
annotations_trimmed_df$cluster <- rownames(annotations_trimmed_df)  # Añadir los IDs de los clusters como columna


# Convertir las anotaciones a un data.frame para unirlas con cluster_markers_df
annotations_ms_df <- as.data.frame(annotations_ms)
annotations_ms_df$cluster <- rownames(annotations_ms_df)  # Añadir los IDs de los clusters como columna


cell_types_median <- annotations_median_df$labels
cell_types_trimmed <- annotations_trimmed_df$labels
cell_types_ms <- annotations_ms_df$labels

# Añadir columna 'labels'
cluster_markers_df$labels_median <- cell_types_median[cluster_markers_df$cluster + 1]
cluster_markers_df$labels_trimmed <- cell_types_trimmed[cluster_markers_df$cluster + 1]
cluster_markers_df$labels_ms <- cell_types_ms[cluster_markers_df$cluster + 1]
#Generate a list 
cluster_list <- cluster_markers_df %>%
  group_split(cluster) %>%
  set_names(paste0("cluster_", unique(cluster_markers_df$cluster)))
#Save xlsx
cluster_list %>% 
write_xlsx(paste0(output_path,"tables/","cluster_list.xlsx"))
#Save RData

# Save using the name
save(cluster_list, file = paste0(output_path, "RData/", "cluster_list", ".RData"))
```


```{r}
#Clean enviroment and memory
# Especifica los nombres de los objetos que deseas conservar
keep_objects <- c("cluster_list", "image_number", "output_path")
# Elimina todos los objetos excepto los especificados
rm(list = setdiff(ls(), keep_objects))
gc() #Clean memory
# Set up parameters
set.seed(123)
options(scipen = -1)  # Habilitar notación científica en R  
source("code/global_variables.R")  
### Load RDatas
# Load results ####
load(paste0(output_path,"RData/cluster_list.RData"))

```

Now, cluster_list will be used as input for the next steps.



# 9. Vulcano Plots

```{r}
filtration <- 0
source("code/02_vulcano_plots.R")

```


# 10. GO

```{r}

source("code/001_GO.R")
```

# 11. Strings
```{r}
source("code/003_strings.R")
```

# 12. GSE
```{r}
source("code/004_gse.R")
```

# 13. HeatMaps
```{r}

# Gene list
gene_list <- c("Hmgcr", "Srebf2", "Apoe","Lrp1", "Lrp1b", "Abca1")

source("code/06_Heatmap.R")
```

# 14. HeatMaps GO Types
```{r}
#source("code/07_HeatMap_GO_types.R")
```

# 15. EnrichR
```{r}
source("code/08_EnrichR.R")
```

