---
title: "Single Cell UbC-Cre CSP Flox"
author: "Santiago López Begines"
date: "2025-02-09"
output:  
  html_document:
    highlight: tango
    fig_width: 8
    fig_height: 7
    fig_caption: true
    code_folding: hide 
    number_sections: true
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
editor_options: 
  markdown: 
    wrap: sentence
---

# 0. Loading Libraries

# Set up global parameters

```{r}

rm(list=ls()) #Clean enviroment
gc() #Clean memory

# Set up parameters
image_number <- 1
output_path <- "output/SC_UBC/SCT/"
source("code/global_variables.R")
source("code/00_packages.R")
source("code/01_sc_functions.R")
set.seed(123)
options(scipen = -1)  # Habilitar notación científica en R
options(digits = 3)
# Increase memory limit for future package
options(future.globals.maxSize = 14000 * 1024^2) # 16 GB limit
#plan("multisession", workers = 4) # Use 4 workers
plan("sequential") # Sequential to save RAM
```

# 1. Importing data

We start by reading in the data.
The Read10X() function reads in the output of the cellranger pipeline from 10X, returning a unique molecular identified (UMI) count matrix.
The values in this matrix represent the number of molecules for each feature (i.e. gene; row) that are detected in each cell (column).
Note that more recent versions of cellranger now also output using the h5 file format, which can be read in using the Read10X_h5() function in Seurat.

```{r}
tic()
#import data
# Load the dataset. Select the folder for every library.
c1 <- Seurat::Read10X(data.dir = "rawdata/SC/AWGC-1_Ctrl-1_results")
c2 <- Seurat::Read10X(data.dir = "rawdata/SC/AWGC-2_Ctrl-2_results")
ko1 <- Seurat::Read10X(data.dir = "rawdata/SC/AWGC-3_KOut-1_results")
ko2 <- Seurat::Read10X(data.dir = "rawdata/SC/AWGC-4_KOut-2_results")

# Initialize the Seurat object with the raw (non-normalized data).
c1 <- CreateSeuratObject(counts = c1, project = "wt_1", min.cells = 3, min.features = 200)
c2 <- CreateSeuratObject(counts = c2, project = "wt_2", min.cells = 3, min.features = 200)
ko1 <- CreateSeuratObject(counts = ko1, project = "ko_1", min.cells = 3, min.features = 200)
ko2 <- CreateSeuratObject(counts = ko2, project = "ko_2", min.cells = 3, min.features = 200)

c1$condition <- "WT"
c2$condition <- "WT"
ko1$condition <- "KO"
ko2$condition <- "KO"



#Put all SeuratObjects in a vector and in a list for automatization
#seurat_names <- c("ubc_c1", "ubc_c2", "ubc_ko1", "ubc_ko2")
# Create a named list of Seurat objects
seurat_objects <- list(
  c1 = c1,
  c2 = c2,
  ko1 = ko1,
  ko2 = ko2
)
# Remove the individual objects from the global environment
rm(c1, c2, ko1, ko2)

# Optionally, run garbage collection to free up memory
gc()

```

```{r}
tic()
# Export results ####
  save(seurat_objects, file = paste0(output_path, "RData/", "seurat_objects", ".RData"))  # Save using the name

toc()

load(paste0(output_path,"RData/seurat_objects.RData"))

```

# 2. Data preprocessing for each library

Preprocessing taken from Seurat web page and <https://hbctraining.github.io/scRNA-seq/lessons/04_SC_quality_control.html>

## QC and selecting cells for further analysis

Seurat allows you to easily explore QC metrics and filter cells based on any user-defined criteria.
A few QC metrics commonly used by the community include \## Identify mitochondrial genes The number of unique genes detected in each cell.
Low-quality cells or empty droplets will often have very few genes Cell doublets or multiplets may exhibit an aberrantly high gene count Similarly, the total number of molecules detected within a cell (correlates strongly with unique genes) The percentage of reads that map to the mitochondrial genome Low-quality / dying cells often exhibit extensive mitochondrial contamination We calculate mitochondrial QC metrics with the 'PercentageFeatureSet()' function, which calculates the percentage of counts originating from a set of features We use the set of all genes starting with 'mt-' as a set of mitochondrial genes

```{r}
# Add QC stats to Seurat objects in the list
seurat_objects <- lapply(names(seurat_objects), function(name) {
  obj <- seurat_objects[[name]]  # Access the object from the list
  # Add percentage of mitochondrial genes
  obj$percent.mt <- PercentageFeatureSet(obj, pattern = "^mt-")  
  # Add number of genes per UMI for each cell to metadata
  obj$log10GenesPerUMI <- log10(obj$nFeature_RNA) / log10(obj$nCount_RNA)
  # Add percentage of ribosomal genes
  obj$percent.ribosomal <- PercentageFeatureSet(obj,pattern="^Rp[ls]") 
  obj  # Return the modified object
}) %>% setNames(names(seurat_objects))  # Preserve the original names in the list
```

## QC cell Pre-Filter

```{r}
# Combine metadata from all Seurat objects into a single dataframe
combined_meta <- bind_rows(
  seurat_objects$c1@meta.data %>% mutate(library = "c1"),
  seurat_objects$c2@meta.data %>% mutate(library = "c2"),
  seurat_objects$ko1@meta.data %>% mutate(library = "ko1"),
  seurat_objects$ko2@meta.data %>% mutate(library = "ko2")
)

```

### Number of cells per sample

```{r}
# Visualize the number of cells per library
combined_plot <- combined_meta %>%
  ggplot(aes(x = library, fill = condition)) +
  geom_bar() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("Number of Cells per Library") +
  xlab("Library") +
  ylab("Number of Cells")

print(combined_plot)

  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Number_of_Cells")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")
  
  image_number <- image_number+1
  
```

### Distribution of Dnajc5 per library

```{r}
# Add a library identifier to each object for grouping
seurat_objects <- lapply(names(seurat_objects), function(name) {
  obj <- seurat_objects[[name]]
  obj$library <- name  # Add library name to metadata
  obj  # Return updated object
}) %>% setNames(names(seurat_objects))

# Merge all Seurat objects into one for combined plotting
merged_obj <- merge(seurat_objects[[1]], y = seurat_objects[-1])

# Plot the expression of "Dnajc5" across all libraries
plot <- VlnPlot(merged_obj, features = "Dnajc5", assay = "RNA", group.by = "library") +
  ggtitle("Expression of Dnajc5 Across Libraries") +
  theme_minimal()
print(plot)
  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Dnajc5_Expression")
  ggsave(paste0(filename, tiff_extension), plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 8, height = 6, units = "in")
  
  image_number <- image_number+1
```

### UMI counts (transcripts) per cell

The UMI counts per cell (nCount_RNA) should generally be above 500, that is the low end of what we expect.
If UMI counts are between 500-1000 counts, it is usable but the cells probably should have been sequenced more deeply.

```{r}
# Visualize the number UMIs/transcripts per cell
# Visualize the number of cells per library
combined_plot <- combined_meta %>%
  ggplot(aes(x = nCount_RNA, color=library, fill = library)) +
  geom_density(alpha = 0.2) + 
  scale_x_log10() + 
  theme_classic() +
  ylab("Density") +
  geom_vline(xintercept = 500) 


print(combined_plot)

  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_UMI_per_cell")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")

    image_number <- image_number+1
```

We can see that majority of our cells in both samples have 1000 UMIs or greater, which is great.

### Genes detected per cell

We have similar expectations for gene detection as for UMI detection, although it may be a bit lower than UMIs.
For high quality data, the proportional histogram should contain a single large peak that represents cells that were encapsulated.
If we see a small shoulder to the right of the major peak (not present in our data), or a bimodal distribution of the cells, that can indicate a couple of things.
It might be that there are a set of cells that failed for some reason.
It could also be that there are biologically different types of cells (i.e. quiescent cell populations, less complex cells of interest), and/or one type is much smaller than the other (i.e. cells with high counts may be cells that are larger in size).
Therefore, this threshold should be assessed with other metrics that we describe in this lesson.

```{r}
# Visualize the distribution of genes detected per cell via histogram
combined_plot <- combined_meta %>% 
  	ggplot(aes(color=library, x=nFeature_RNA, fill= library)) + 
  	geom_density(alpha = 0.2) + 
  	theme_classic() +
  	scale_x_log10() + 
  	geom_vline(xintercept = 300)+
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) 

print(combined_plot)

  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Genes_per_cell_denstity")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")
  image_number <- image_number+1
# Visualize the distribution of genes detected per cell via boxplot
combined_plot <- combined_meta %>% 
  	ggplot(aes(x=library, y=log10(nFeature_RNA), fill=library)) + 
  	geom_boxplot() + 
  	theme_classic() +
  	theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  	theme(plot.title = element_text(hjust=0.5, face="bold")) +
  	ggtitle("NCells vs NGenes")

print(combined_plot)

  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Genes_per_cell_boxplot")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")
  image_number <- image_number+1
```

### UMIs vs. genes detected

Two metrics that are often evaluated together are the number of UMIs and the number of genes detected per cell.
Here, we have plotted the number of genes versus the numnber of UMIs coloured by the fraction of mitochondrial reads.
Mitochondrial read fractions are only high (light blue color) in particularly low count cells with few detected genes.
This could be indicative of damaged/dying cells whose cytoplasmic mRNA has leaked out through a broken membrane, and thus, only mRNA located in the mitochondria is still conserved.
These cells are filtered out by our count and gene number thresholds.
Jointly visualizing the count and gene thresholds shows the joint filtering effect.

Cells that are poor quality are likely to have low genes and UMIs per cell, and correspond to the data points in the bottom left quadrant of the plot.
Good cells will generally exhibit both higher number of genes per cell and higher numbers of UMIs.

With this plot we also evaluate the slope of the line, and any scatter of data points in the bottom right hand quadrant of the plot.
These cells have a high number of UMIs but only a few number of genes.
These could be dying cells, but also could represent a population of a low complexity celltype (i.e red blood cells).

```{r}
# Visualize the correlation between genes detected and number of UMIs and determine whether strong presence of cells with low numbers of genes/UMIs
combined_plot <- combined_meta %>% 
  	ggplot(aes(x=nCount_RNA, y=nFeature_RNA, color=percent.mt)) + 
  	geom_point() + 
	scale_colour_gradient(low = "gray90", high = "black") +
  	stat_smooth(method=lm) +
  	scale_x_log10() + 
  	scale_y_log10() + 
  	theme_classic() +
  	geom_vline(xintercept = 500) +
  	geom_hline(yintercept = 250) +
  	facet_wrap(~library)


print(combined_plot)

  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Genes_vs_Cell")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")
  image_number <- image_number+1
```

### Mitochondrial counts ratio

This metric can identify whether there is a large amount of mitochondrial contamination from dead or dying cells.
We define poor quality samples for mitochondrial counts as cells which surpass the 0.2 mitochondrial ratio mark, unless of course you are expecting this in your sample.

```{r}
# Visualize the distribution of mitochondrial gene expression detected per cell
combined_plot <- combined_meta %>% 
  	ggplot(aes(color=library, x=percent.mt, fill=library)) + 
  	geom_density(alpha = 0.2) + 
  	scale_x_log10() + 
  	theme_classic() +
    geom_vline(xintercept = 1)+#Threshold of 1% of mitochondrial genes
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

print(combined_plot)

  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Density_MitoGenes_Cell")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")
  image_number <- image_number+1

```

### Complexity

We can see the samples where we sequenced each cell less have a higher overall complexity, that is because we have not started saturating the sequencing for any given gene for these samples.
Outlier cells in these samples might be cells that have a less complex RNA species than other cells.
Sometimes we can detect contamination with low complexity cell types like red blood cells via this metric.
Generally, we expect the novelty score to be above 0.80.

```{r}
# Visualize the overall complexity of the gene expression by visualizing the genes detected per UMI
combined_plot <- combined_meta %>%
  	ggplot(aes(x=log10GenesPerUMI, color = library, fill=library)) +
  	geom_density(alpha = 0.2) +
  	theme_classic() +
  	geom_vline(xintercept = 0.8)+
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

print(combined_plot)
  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Complexity")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")
  image_number <- image_number+1

```

### Percentage of mitochondrial genes

```{r}
source("code/01_sc_functions.R")
# Load the function from the R script
#source("01_qc_functions.R")

# Generate QC plots for objects in the list
lapply(names(seurat_objects), function(name) {
  obj <- seurat_objects[[name]]  # Access the object from the list
  image_number <<- generate_qc_plots(obj, name, image_number, output_path, suffix = "prefilter")
})
```

## QC cell Filter

We filter cells that have unique feature counts over 7,500 or less than 850 We filter cells that have \>1% mitochondrial counts

*Malat1*: It is a highly expressed lncRNA and can distort variability.
*Xist*: Used to identify female cells due to its role in X chromosome inactivation.
*mt-Co1, mt-Co2*: Mitochondrial genes that can indicate cells under mitochondrial stress or poor quality.
*mt-Nd1 - mt-Nd6, mt-Cytb*: Additional mitochondrial genes that are usually highly expressed.

```{r}
#Remove non-informative genes that could lead to downstream bias.
genes_to_exclude <- c("Malat1", "Xist", "mt-Co1", "mt-Co2", "mt-Nd1", "mt-Nd2", "mt-Nd3", "mt-Nd4", "mt-Nd5", "mt-Nd6", "mt-Cytb")

# Filter Seurat objects based on QC metrics
filter_seurat_objects <- lapply(names(seurat_objects), function(name) {
  obj <- seurat_objects[[name]]  # Access the object from the list
  obj <- subset(obj, features = setdiff(rownames(obj), genes_to_exclude))
  obj <- subset(obj, subset = nFeature_RNA > 850 & nFeature_RNA < 7500 & percent.mt < 1)  # Apply QC filters
  obj  # Return the filtered object
}) %>% setNames(names(seurat_objects))  # Preserve the original names in the list

```

```{r}
presents <- genes_to_exclude[genes_to_exclude %in% rownames(filter_seurat_objects)]

```

```{r}
# Combine metadata from all Seurat objects into a single dataframe
combined_meta_filter <- bind_rows(
  filter_seurat_objects$c1@meta.data %>% mutate(library = "c1"),
  filter_seurat_objects$c2@meta.data %>% mutate(library = "c2"),
  filter_seurat_objects$ko1@meta.data %>% mutate(library = "ko1"),
  filter_seurat_objects$ko2@meta.data %>% mutate(library = "ko2")
)

```

### Number of cells per sample

```{r}
# Visualize the number of cells per library
combined_plot <- combined_meta_filter %>%
  ggplot(aes(x = library, fill = condition)) +
  geom_bar() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("Number of Cells per Library") +
  xlab("Library") +
  ylab("Number of Cells")

print(combined_plot)

  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Number_of_Cells_filter")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")
  
  image_number <- image_number+1
  
```

### Distribution of Dnajc5 per library

```{r}
# Merge all Seurat objects into one for combined plotting
filter_merged_obj <- merge(filter_seurat_objects[[1]], y = filter_seurat_objects[-1])

# Plot the expression of "Dnajc5" across all libraries
plot <- VlnPlot(filter_merged_obj, features = "Dnajc5", assay = "RNA", group.by = "library") +
  ggtitle("Expression of Dnajc5 Across Libraries") +
  theme_minimal()
print(plot)
  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Dnajc5_Expression")
  ggsave(paste0(filename, tiff_extension), plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 8, height = 6, units = "in")
  
  image_number <- image_number+1
```

### UMI counts (transcripts) per cell

The UMI counts per cell (nCount_RNA) should generally be above 500, that is the low end of what we expect.
If UMI counts are between 500-1000 counts, it is usable but the cells probably should have been sequenced more deeply.

```{r}
# Visualize the number UMIs/transcripts per cell
# Visualize the number of cells per library
combined_plot <- combined_meta_filter %>%
  ggplot(aes(x = nCount_RNA, color=library, fill = library)) +
  geom_density(alpha = 0.2) + 
  scale_x_log10() + 
  theme_classic() +
  ylab("UMI per cell") +
  geom_vline(xintercept = 500) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))


print(combined_plot)

  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_UMI_per_cell_filter")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")

    image_number <- image_number+1
```

We can see that majority of our cells in both samples have 1000 UMIs or greater, which is great.

### Genes detected per cell

We have similar expectations for gene detection as for UMI detection, although it may be a bit lower than UMIs.
For high quality data, the proportional histogram should contain a single large peak that represents cells that were encapsulated.
If we see a small shoulder to the right of the major peak (not present in our data), or a bimodal distribution of the cells, that can indicate a couple of things.
It might be that there are a set of cells that failed for some reason.
It could also be that there are biologically different types of cells (i.e. quiescent cell populations, less complex cells of interest), and/or one type is much smaller than the other (i.e. cells with high counts may be cells that are larger in size).
Therefore, this threshold should be assessed with other metrics that we describe in this lesson.

```{r}
# Visualize the distribution of genes detected per cell via histogram
combined_plot <- combined_meta_filter %>% 
  	ggplot(aes(color=library, x=nFeature_RNA, fill= library)) + 
  	geom_density(alpha = 0.2) + 
  	theme_classic() +
  	scale_x_log10() + 
  	geom_vline(xintercept = 300)+
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

print(combined_plot)

  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Genes_per_cell_denstity_filter")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")
  image_number <- image_number+1
# Visualize the distribution of genes detected per cell via boxplot
combined_plot <- combined_meta_filter %>% 
  	ggplot(aes(x=library, y=log10(nFeature_RNA), fill=library)) + 
  	geom_boxplot() + 
  	theme_classic() +
  	theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  	theme(plot.title = element_text(hjust=0.5, face="bold")) +
  	ggtitle("NCells vs NGenes")

print(combined_plot)

  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Genes_per_cell_boxplot_filter")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")
  image_number <- image_number+1
```

### UMIs vs. genes detected

Two metrics that are often evaluated together are the number of UMIs and the number of genes detected per cell.
Here, we have plotted the number of genes versus the numnber of UMIs coloured by the fraction of mitochondrial reads.
Mitochondrial read fractions are only high (light blue color) in particularly low count cells with few detected genes.
This could be indicative of damaged/dying cells whose cytoplasmic mRNA has leaked out through a broken membrane, and thus, only mRNA located in the mitochondria is still conserved.
These cells are filtered out by our count and gene number thresholds.
Jointly visualizing the count and gene thresholds shows the joint filtering effect.

Cells that are poor quality are likely to have low genes and UMIs per cell, and correspond to the data points in the bottom left quadrant of the plot.
Good cells will generally exhibit both higher number of genes per cell and higher numbers of UMIs.

With this plot we also evaluate the slope of the line, and any scatter of data points in the bottom right hand quadrant of the plot.
These cells have a high number of UMIs but only a few number of genes.
These could be dying cells, but also could represent a population of a low complexity celltype (i.e red blood cells).

```{r}

# Visualize the correlation between genes detected and number of UMIs and determine whether strong presence of cells with low numbers of genes/UMIs
combined_plot <- combined_meta_filter %>% 
  	ggplot(aes(x=nCount_RNA, y=nFeature_RNA, color=percent.mt)) + 
  	geom_point() + 
	scale_colour_gradient(low = "gray90", high = "black") +
  	stat_smooth(method=lm) +
  	scale_x_log10() + 
  	scale_y_log10() + 
  	theme_classic() +
  	geom_vline(xintercept = 500) +
  	geom_hline(yintercept = 250) +
  	facet_wrap(~library)


print(combined_plot)

  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Genes_vs_Cell_filter")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")
  image_number <- image_number+1
```

### Complexity

We can see the samples where we sequenced each cell less have a higher overall complexity, that is because we have not started saturating the sequencing for any given gene for these samples.
Outlier cells in these samples might be cells that have a less complex RNA species than other cells.
Sometimes we can detect contamination with low complexity cell types like red blood cells via this metric.
Generally, we expect the novelty score to be above 0.80.

```{r}
# Visualize the overall complexity of the gene expression by visualizing the genes detected per UMI
combined_plot <- combined_meta_filter %>%
  	ggplot(aes(x=log10GenesPerUMI, color = library, fill=library)) +
  	geom_density(alpha = 0.2) +
  	theme_classic() +
  	geom_vline(xintercept = 0.8)+
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

print(combined_plot)
  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Complexity_filter")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")
  image_number <- image_number+1

```

### Mitochondrial counts ratio

This metric can identify whether there is a large amount of mitochondrial contamination from dead or dying cells.
We define poor quality samples for mitochondrial counts as cells which surpass the 0.2 mitochondrial ratio mark, unless of course you are expecting this in your sample.

```{r}
# Visualize the distribution of mitochondrial gene expression detected per cell
combined_plot <- combined_meta_filter %>% 
  	ggplot(aes(color=library, x=percent.mt, fill=library)) + 
  	geom_density(alpha = 0.2) + 
  	scale_x_log10() + 
  	theme_classic() +
  	geom_vline(xintercept = 1)+ #Threshold of 1% of mitochondrial genes
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

print(combined_plot)

  # Set filenames with suffix
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_Density_MitoGenes_Cell_filter")
  ggsave(paste0(filename, tiff_extension), combined_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), combined_plot, width = 8, height = 6, units = "in")
  image_number <- image_number+1

```

### Percentage of mitochondrial genes

```{r}
# Generate QC plots for objects in the list
lapply(names(filter_seurat_objects), function(name) {
  obj <- seurat_objects[[name]]  # Access the object from the list
  image_number <<- generate_qc_plots(obj, name, image_number, output_path, suffix = "filter")
})
```

```{r}
# Ensure unique cell names
filter_seurat_objects <- lapply(names(filter_seurat_objects), function(name) {
  obj <- filter_seurat_objects[[name]]
  obj <- RenameCells(obj, add.cell.id = name)
  return(obj)
}) %>% setNames(names(filter_seurat_objects))
tic()
# Export results ####
  save(filter_seurat_objects, file = paste0(output_path, "RData/", "filter_seurat_objects", ".RData"))  # Save using the name
  load(paste0(output_path,"RData/filter_seurat_objects.RData"))
toc()
```

# 3. Doublets Finder

```{r}
source("code/Doublets_Finders.R")
```

# 4.Normalizing the data with SCT

```{r}
# Ensure unique cell names
doublets_seurat_objects <- lapply(names(doublets_seurat_objects), function(name) {
  obj <- doublets_seurat_objects[[name]]
  obj <- RenameCells(obj, add.cell.id = name)
  return(obj)
}) %>% setNames(names(doublets_seurat_objects))

```

## SCTransform

```{r}
#Clean enviroment and memory
# Especifica los nombres de los objetos que deseas conservar
keep_objects <- c("doublets_seurat_objects","filter_seurat_objects", "image_number", "output_path")
# Elimina todos los objetos excepto los especificados
rm(list = setdiff(ls(), keep_objects))
gc() #Clean memory
# Set up parameters
set.seed(123)
options(scipen = -1)  # Habilitar notación científica en R  
source("code/global_variables.R")
```

```{r}
# Increase memory limit for future package
options(future.globals.maxSize = 14000 * 1024^2) # 26 GB limit
#plan("multisession", workers = 2) # Use 4 workers
plan("sequential") # Use 4 workers

```

```{r}
# Normalize each dataset with SCTransform & PCA
SCT_seurat_objects <- lapply(doublets_seurat_objects, function(obj) {
  obj <- SCTransform(obj, vst.flavor = "v2", method = "glmGamPoi", vars.to.regress = "percent.mt", verbose = TRUE)
  return(obj)
})



```

## Run PCA on Each Dataset

For RPCA-based integration, PCA needs to be run on each dataset after SCTransform.

```{r}
# Run PCA for each object
SCT_seurat_objects <- lapply(SCT_seurat_objects, function(obj) {
  features = VariableFeatures(object = obj)
  obj <- RunPCA(obj, features = features, verbose = TRUE)
  return(obj)
})
```

## Integration

Select integration features:

```{r}
# Select integration features
features <- SelectIntegrationFeatures(object.list = SCT_seurat_objects, nfeatures = 5000)
# Asegúrate de incluir Dnajc5
features <- unique(c(features, "Dnajc5"))
```

Prepare the objects for integration:

```{r}
# Prepare objects for integration
SCT_seurat_objects <- PrepSCTIntegration(object.list = SCT_seurat_objects, anchor.features = features)
  # Establish default assay as integrated for future modifications

```

```{r}
tic()
# Export results ####
  save(SCT_seurat_objects, file = paste0(output_path, "RData/", "SCT_seurat_objects", ".RData"))  # Save using the name
  load(paste0(output_path,"RData/SCT_seurat_objects.RData"))
toc()
```

```{r}
DefaultAssay(SCT_seurat_objects$c1) <- "SCT"
DefaultAssay(SCT_seurat_objects$c2) <- "SCT"
DefaultAssay(SCT_seurat_objects$ko1) <- "SCT"
DefaultAssay(SCT_seurat_objects$ko1) <- "SCT"
```

Find integration anchors:

```{r}
tic()
options(future.globals.maxSize = 14000 * 1024^2) # 16 GB limit
plan("sequential") # Use 4 workers
gc()

toc()
```

```{r}
# Find integration anchors using RPCA
anchors <- FindIntegrationAnchors(object.list = SCT_seurat_objects, normalization.method = "SCT",
                                   anchor.features = features, reduction = "rpca")

```

Integrate the data:

```{r}
tic()
options(future.globals.maxSize = 14000 * 1024^2) # 16 GB limit
plan("sequential") # Use 4 workers
gc()
# Integrate datasets
data_int <- IntegrateData(anchorset = anchors, normalization.method = "SCT")
toc()
```

Run dimensional reduction and clustering on the integrated object:

```{r}
#Dimensionality Reduction. Perform PCA and UMAP to visualize clusters.
data_int <- RunPCA(data_int, verbose = TRUE)
data_int@meta.data$condition <- factor(data_int@meta.data$condition, levels = c("WT", "KO"))
```

```{r}
tic()
# Export results ####
  save(data_int, file = paste0(output_path, "RData/", "data_int", ".RData"))  # Save using the name
  load(paste0(output_path,"RData/data_int.RData"))
toc()
```

## HVG

```{r}
source("code/01_sc_functions.R")
data_int <- FindVariableFeatures(data_int, selection.method = "vst", nfeatures = 2000)
# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(data_int), 10)
# plot variable features with and without labels
plot <- LabelPoints(plot = VariableFeaturePlot(data_int), points = top10, repel = TRUE)
print(plot)
save_plot("HVGs", plot)
```

# 5. Define Clustering Resolution

## Identify significant number of dimensions

### ElbowPlot

```{r}
Elbowplot <- ElbowPlot(data_int, ndims = 50)
print(Elbowplot)
save_plot("Elblowplot", Elbowplot)
```

### Percetage of explained variance by PC

```{r}
# Extract variance explained by each PC
pca_var <- Stdev(object = data_int, reduction = "pca")^2
total_var <- sum(pca_var)
explained_variance <- (pca_var / total_var) * 100  # Percentage of variance explained

# Create a data frame with the results
variance_table <- data.frame(
  PC = seq_along(explained_variance),
  Variance = pca_var,
  ExplainedVariance = explained_variance,
  CumulativeVariance = cumsum(explained_variance)
)



# View the table
print(variance_table)

variance_table %>% write_xlsx(paste0(output_path, "tables/Variance_by_PC.xlsx"))
```

### DimHeatmaps

```{r}

plot <- DimHeatmap(data_int, dims = 1:6, cells = 500, balanced = TRUE)
print(plot)
save_plot("DimHeatmap_1_6", plot)
plot <- DimHeatmap(data_int, dims = 7:12, cells = 500, balanced = TRUE)
print(plot)
save_plot("DimHeatmap_7_12", plot)
plot <- DimHeatmap(data_int, dims = 13:18, cells = 500, balanced = TRUE)
save_plot("DimHeatmap_13_18", plot)
print(plot)
plot <- DimHeatmap(data_int, dims = 19:24, cells = 500, balanced = TRUE)
save_plot("DimHeatmap_19_24", plot)
print(plot)

```

## UMAP: Testing dimensions and resolution

```{r}
# Define the resolutions to test
resolutions <- c(0.1, 0.2, 0.4, 0.8)
data_int_list <- list()
# Loop over dimensions and resolutions
for (dims in c(10, 20, 30)) {
  print(paste("Testing with", dims, "PCs"))
  
  for (res in resolutions) {
    print(paste("Testing resolution =", res))
    
    # Neighbor graph and clustering
    data_dim_res <- FindNeighbors(data_int, dims = 1:dims)
    data_dim_res <- FindClusters(data_dim_res, resolution = res)
    col_name <- paste0("cluster_Dims_", dims, "_Res_", res)
    data_dim_res[[col_name]] <- Idents(data_dim_res)
    # Run UMAP
    data_dim_res <- RunUMAP(data_dim_res, dims = 1:dims)
    
    # Save the Seurat object with a unique name
    data_int_list[[paste0("Dims_", dims, "_Res_", res)]] <- data_dim_res
    # Guarda el objeto individualmente
    saveRDS(data_dim_res, file = paste0(output_path, "RData/data_int_Dims_", dims, "_Res_", res, ".rds"))
    
    # Visualize clusters
    plot <- DimPlot(data_dim_res, reduction = "umap", label = TRUE) + 
      ggtitle(paste("Dims =", dims, "Resolution =", res))
    print(plot)
    
    # Save plot
    plotname <- paste0("UMAP_Dims_", dims, "_Res_", res)
    save_plot(plotname, plot)
  }
}

rm(data_dim_res)
```

## Combine Clustering Results into Metadata

```{r}
# Initialize a data frame to store all clustering results
combined_metadata <- data.frame()

# Loop through each element in the list
for (name in names(data_int_list)) {
  # Extract the Seurat object
  seurat_obj <- data_int_list[[name]]
  
  # Extract cluster IDs
  clusters <- Idents(seurat_obj)
  
  # Extract cell barcodes and resolution/dimensions
  cell_barcodes <- colnames(seurat_obj)
  dims <- gsub("_Res_.*", "", gsub("Dims_", "", name))  # Extract dims
  resolution <- gsub(".*_Res_", "", name)  # Extract resolution
  
  # Create a metadata table for this combination
  metadata <- data.frame(
    cell = cell_barcodes,
    cluster = clusters,
    dims = dims,
    resolution = resolution
  )
  
  # Add this metadata to the combined table
  combined_metadata <- bind_rows(combined_metadata, metadata)
}

# View the combined metadata
head(combined_metadata)

# Reshape metadata so each resolution is a separate column
metadata_wide <- combined_metadata %>%
  pivot_wider(names_from = resolution, values_from = cluster, names_prefix = "Res_")

# Add this back to the Seurat object (if needed) or use directly for clustree

```

## Clustree

```{r}
# Generate the clustree plot
plot <- clustree(metadata_wide, prefix = "Res_")
print(plot)
save_plot("clustree", plot)
```

```{r}
# Export results ####
save(data_int, file = paste0(output_path, "RData/", "data_int", ".RData")) # Save using the name
#save(data_int_list, file = paste0(output_path, "RData/", "data_int_list", ".RData"), compress = "xz") # Save using the name
```

```{r}
#load(paste0(output_path, "RData/", "data_int", ".RData"))
#load(paste0(output_path, "RData/", "data_int_list", ".RData"))
```

Step 5: Cell Type Proportions by Genotype Compare the distribution of cell types between WT and KO.

## Cell type proportion

Calculate Cell Type Proportions Tabulate proportions for each cluster by genotype (condition).

```{r}
# Calculate proportions for each resolution and combine results
proportion_results <- list()

# Loop through each resolution in the data_int_list
for (res in names(data_int_list)) {
  # Extract the Seurat object for the current resolution
  seurat_obj <- data_int_list[[res]]
  
  # Calculate the table of counts by cluster and genotype
  cell_type_table <- table(Idents(seurat_obj), seurat_obj$condition)
  
  # Convert to proportions
  proportions <- as.data.frame(prop.table(cell_type_table, margin = 2))
  
  # Add resolution and dimensions as columns for reference
  proportions$resolution <- gsub(".*_Res_", "", res)
  proportions$dims <- gsub("_Res_.*", "", gsub("Dims_", "", res))
  
  # Store the result in the list
  proportion_results[[res]] <- proportions
}

# Combine all results into a single data frame
proportions_all <- do.call(rbind, proportion_results)

# Print the combined proportions table
print(proportions_all)

# Save the proportions table as a CSV for future reference
proportions_all %>% write_xlsx(paste0(output_path,"tables/Cell_type_genotype_proportion.xlsx"))

```

```{r}
# Ensure 'dims' is treated as a factor for better plotting
proportions_all$dims <- factor(proportions_all$dims, levels = unique(proportions_all$dims))

# Loop through each dimension and generate separate plots for resolutions
for (dim in levels(proportions_all$dims)) {
  # Filter the data for the current dimension
  filtered_data <- proportions_all %>% filter(dims == dim)
  
  # Generate the plot
  plot <- ggplot(filtered_data, aes(x = Var1, y = Freq, fill = Var2)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~ resolution, ncol = 2) +  # Facet by resolution
    labs(
      x = "Cluster",
      y = "Proportion",
      fill = "Condition",
      title = paste("Cluster Proportions for Dimensions =", dim)
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.major = element_line(color = "gray", linetype = "dotted"),
      legend.position = "bottom"
    )
  
  # Save the plot
  filename <- paste0("Cluster_Proportions_Dims_", dim)
 save_plot(filename, plot)
  
  # Optionally print the plot to the console
  print(plot)
}


```

# Select one specific resolution and number of dimensions

```{r}
dims <- 30
res <- 0.8

data_select <- readRDS(paste0(output_path, "RData/data_int_Dims_", dims, "_Res_", res, ".rds"))
data_select@meta.data$condition <- factor(data_select@meta.data$condition, levels = c("WT", "KO"))
rm("data_int_list")
```

# 6. Plot UMAP & UMAP Features

```{r}
tic()

## UMAP Visualize clusters
p1 <- DimPlot(data_select, reduction = "umap", group.by = "condition")
p2 <- DimPlot(data_select, reduction = "umap", split.by="condition", ncol = 2, label = TRUE) + NoLegend()

  plot <- p1 | p2
  print(plot)
  save_plot("clusters_integrated", plot)
  #UMAP
 # note that you can set `label = TRUE` or use the LabelClusters function to help label individual clusters
  plot <- DimPlot(data_select, reduction = "umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + NoLegend()
  print(plot)
  save_plot("UMAP", plot)
#UMAP by condition
  plot <- DimPlot(data_select, reduction = "umap", split.by="condition", ncol = 2, label = TRUE)
  print(plot)
  save_plot("UMAP_ident", plot)
  
  #UMAP by features
  features <- c("nFeature_RNA", "nCount_RNA")
  feature_plot <- FeaturePlot(data_select, features = features)
  print(feature_plot)
  save_plot("Feature_plot", feature_plot)
  
toc()

```

## UMAP HVGs

```{r}
top <-head(VariableFeatures(data_select),10)

  #UMAP by features
  feature_plot <- FeaturePlot(data_select, features = top, ncol = 2)
    print(feature_plot)
  save_plot("Feature_plot_HVGs", feature_plot,width = 8, height = 18)
  
```

## UMAP Dnajc5

```{r}
DefaultAssay(data_select) <- "integrated"

  #UMAP by features
  feature_plot <- FeaturePlot(data_select, 
                              features="Dnajc5",
                              split.by="condition", 
                              ncol = 2,
                              slot = "data")
    print(feature_plot)
    save_plot("Feature_plot", feature_plot)
  
  # Cambia el assay a RNA
DefaultAssay(data_select) <- "RNA"

# Visualiza la expresión de Dnajc5 en los clústeres integrados
plot <- VlnPlot(data_select, features = "Dnajc5", group.by = "seurat_clusters", split.by = "condition") +
  ggtitle("Dnajc5 Expression in clusters (RNA Assay)")
print(plot)
save_plot("Vln_plot_Dnajc5", plot)

  
```

# 7. Finding differentially expressed features (cluster biomarkers)

Seurat can help you find markers that define clusters via differential expression (DE).
By default, it identifies positive and negative markers of a single cluster (specified in ident.1), compared to all other cells.
FindAllMarkers() automates this process for all clusters, but you can also test groups of clusters vs. each other, or against all cells.

In Seurat v5, we use the presto package (as described here and available for installation here), to dramatically improve the speed of DE analysis, particularly for large datasets.
For users who are not using presto, you can examine the documentation for this function (?FindMarkers) to explore the min.pct and logfc.threshold parameters, which can be increased in order to increase the speed of DE testing.

## Subset Clusters and Compare Genotypes

```{r}
options(future.globals.maxSize = 28000 * 1024^2) # 16 GB limit
plan("sequential") # Use 4 workers
gc()

# Normalizar RNA
DefaultAssay(data_select) <- "RNA"
data_select <- NormalizeData(data_select, normalization.method = "LogNormalize", scale.factor = 10000)

# Unir las capas del assay RNA antes de ejecutar FindMarkers
data_select <- JoinLayers(data_select, assay = "RNA")

```

```{r}
# Verificar que 'condition' está en metadata
if (!"condition" %in% colnames(data_select@meta.data)) {
  stop("Error: Column 'condition' does not exist in metadata. Verify that 'WT' & 'KO' are correctly assigned.")
}
# Establecer los clústeres como identidades antes de la iteración
Idents(data_select) <- "seurat_clusters"  # Set cluster identities

# List to store results
cluster_list <- list()

# Loop through each cluster
cluster_ids <- levels(data_select$seurat_clusters)

for (cluster in cluster_ids) {
  # Subset cells belonging to the current cluster
  cluster_cells <- subset(data_select, idents = cluster)
  
  Idents(cluster_cells) <- cluster_cells$condition  # Asigna 'WT' y 'KO' directamente como identidades                                    # Verificar que el clúster tiene ambas condiciones antes de correr FindMarkers
  if (length(unique(Idents(cluster_cells))) < 2) {
    warning(paste("Skipping cluster", cluster, "because does not have cells in both conditions (WT & KO)."))
    next  # Salta al siguiente clúster
  }                                 
  # Compare WT vs KO within the cluster
  res <- FindMarkers(cluster_cells, 
                     ident.1 = "WT", 
                     ident.2 = "KO", 
                     test.use = "MAST", # Cambia si prefieres otro test
                     assay = "RNA",          # Use RNA assay
                     min.pct = 0.1,         # Filter genes expressed in at least 25% of cells
                     logfc.threshold = 0.1) # Minimum log fold-change threshold
  
  # Add cluster info to results
  res$gene <- rownames(res)
  res$cluster <- cluster
  cluster_list[[cluster]] <- res
}

# Save using the name
save(cluster_list, file = paste0(output_path, "RData/", "cluster_list", ".RData"))  
cluster_list %>% 
write_xlsx(paste0(output_path,"tables/","cluster_list.xlsx"))
# Load results ####
load(paste0(output_path,"RData/cluster_list.RData"))

# Convertir la lista en un data.frame con una columna adicional 'cluster' (valor numérico)
cluster_markers_df <- do.call(rbind, lapply(names(cluster_list), function(cluster) {
  # Extraer el dataframe del cluster actual y añadir una columna con el número del cluster
  markers <- cluster_list[[cluster]]
  markers$cluster <- as.numeric(gsub("Cluster_", "", cluster)) # Eliminar el prefijo y convertir a numérico
  return(markers)
}))

# Reiniciar los nombres de las filas
rownames(cluster_markers_df) <- NULL




cluster_markers_df <- cluster_markers_df %>% 
  dplyr::select(gene, everything())

# Ver el resultado
head(cluster_markers_df)

#Generate a list 

cluster_markers_df %>% 
write_xlsx(paste0(output_path,"tables/","cluster_markers_df.xlsx"))
# Save using the name
save(cluster_markers_df, file = paste0(output_path, "RData/", "cluster_markers_df", ".RData"))  

### Load RDatas

# Load results ####
load(paste0(output_path,"RData/cluster_markers_df.RData"))



```

## FindAllMarkers

```{r}
plan("sequential")  # Usa 6 núcleos (ajusta según tu PC)
options(future.globals.maxSize = 14000 * 1024^2) # 16 GB limit
gc()
# find markers for every cluster compared to all remaining cells, report only the positive
# ones
DefaultAssay(data_select) <- "integrated" #SCT
#data_select <- PrepSCTFindMarkers(data_select)
Idents(data_select) <- "seurat_clusters"  # Set cluster identities. Look for differences between cluster. Dont look for differences between genotype.
#Idents(data_select) <- "condition"  # Set condition identities. Look for differences between genotype and dont look for differenes between clusters

# Encuentra marcadores de todos los clusters
cluster.all.markers <- FindAllMarkers(data_select, 
                                      test.use = "wilcox",  # Test estadístico Wilcox,wilcox_limma, LR, MAST, DESeq2
                                      only.pos = FALSE, 
                                      min.pct = 0.1, 
                                      logfc.threshold = 0.1)


cluster.all.markers %>% 
write_xlsx(paste0(output_path,"tables/","cluster.all.markers.xlsx"))

save(cluster.all.markers, file = paste0(output_path, "RData/", "cluster.all.markers", ".RData")) # Save using the name
load(paste0(output_path, "RData/", "cluster.all.markers", ".RData"))
plan("sequential")  # Usa 6 núcleos (ajusta según tu PC)
```

## HeatMap

```{r}
source("code/01_sc_functions.R")
DefaultAssay(data_select) <- "SCT" #integrated
top10_all <- cluster.all.markers %>%
  group_by(cluster) %>%
  dplyr::filter(avg_log2FC > FC, p_val_adj < 0.05) %>%
  slice_head(n = 10) %>%
  ungroup()

top10 <- cluster_markers_df %>%
  group_by(cluster) %>%
  dplyr::filter(avg_log2FC > FC) %>%
  slice_head(n = 10) %>%
  ungroup()



plot <- DoHeatmap(data_select, features = top10_all$gene) + NoLegend()
print(plot)
save_plot("HeatMap_cluster_plot", plot, height = 18)


```

# 8. Manual Clustering Association

## Brain cell types (Rubio-Pastor F. PhD Thsis 2025)

We used Uniform Manifold Approximation and Projection (UMAP) to visualize all cell clusters and classified and annotated distinct cell types based on known marker genes for each major brain cell type.
Following the initial cell clustering, we identified 7 major populations based on the expression of common cell-type markers (Haney et al., 2024): excitatory neurons (Slc17a7), inhibitory neurons (Gad1), oligodendrocytes (Mbp), oligodendrocyte precursor cells (OPCs, Cacng4), microglia (Siglech), astrocytes (Aqp4) and vascular cells (Cldn5) (Figure 67).
We then subdivided both excitatory and inhibitory neurons into distinct cell subtypes, annotating them further based on known markers and cortical cytoarchitecture, resulting in 14 clusters (Figure 68).
For excitatory neurons, clustering was performed using their gene expression patterns and known markers (Allen Institute, n.d.; Matho et al., 2021), as classification by cortical layer location proved challenging and produced a large number of distinct subtypes.
We identified 5 different clusters (Figure 69): ExN1 (Lamp5+ Cux2+) enriched in L2/3, ExN2 (Ndst4+) enriched in L2/3, ExN3 (Rorb+ Cux1+) enriched in L4, ExN4 (Tox+ Hs3st2+) enriched in L4/5, and ExN5 (Tle4+ Fexf2+) enriched in L5/6.
We classified inhibitory neurons into subtypes based on previous studies (Bugeon et al., 2022; Tasic et al., 2018; Yao et al., 2021), resulting in four groups: InN1 (Pvalb+), InN2 (Sst+), InN3 (Vip+), and InN4 (Lamp5+).

-   Excitatory Neurons: Slc17a7
    -   ExN1 enriched L2/3: Lamp5, Cux2
    -   ExN2 enriched L2/3: Ndst4
    -   ExN3 enriched L4: Rorb, Cux2
    -   ExN4 enriched L4/5: Tox, Hs3st2
    -   ExN5 enriched L5/6: Tle4, Fexf2
-   Inhibitory Neurons: Gad1
    -   InN1: Pvalb
    -   InN2: Sst
    -   InN3: Vip
    -   InN4: Lamp5
-   Oligodendrocytes: Mbp
-   OPC (oligodendrocyte precursor cells): Cacng4
-   Microglia: Siglech
-   Astrocytes: Aqp4
-   Vascular cells: Cldn5

## Brain cell types ChatGPT

Neuronales:

-   Neuronas de Purkinje: Calbindina (CALB1)
-   Neuronas granulares: GABRA6 (subunidad α6 del receptor GABA-A)
-   Neuronas inhibitorias(interneuronas): Parvalbúmina (PV), Somatostatina (SST), Calretinina(CALB2)
-   Neuronas excitatorias (proyección): Vesicular Glutamate Transporter 1 (VGLUT1), Vesicular Glutamate Transporter 2 (VGLUT2)
-   Neuronas dopaminérgicas: Tirosina hidroxilasa (TH)
-   Neuronas serotoninérgicas: Tryptophan Hydroxylase 2 (TPH2)
-   Neuronas colinérgicas: Colina acetiltransferasa (ChAT)

No neuronales:

-   Astrocitos: Proteína ácida fibrilar glial (GFAP), S100β
-   Oligodendrocitos: Proteína proteolipídica (PLP1), Mielina básica proteína (MBP)
-   Células microgliales: Ionized Calcium Binding Adapter Molecule 1 (IBA1, Aif1), CD11b (Cdk11b),Cx3cr1
-   Células endoteliales: Cluster of Differentiation 31 (CD31), Claudina-5 (CLDN5)

## Cerebellum cell types ()

-   Purkinje cells: Ppp1r17
-   Granule: Gabra6
-   UBC: Eomes
-   MLI1: Lypd6
-   MLI2: Prkcd
-   PLI: Klhl1
-   Golgi: Lgi2
-   Bergmannn: Gdf10
-   Astrocyte: Aqp4
-   ODC: Mobp
-   OPC: Ppfibp1
-   Fibroblast: Dcn
-   Endothelial Mural: Kcnj8
-   Choroid: Ttr
-   Macrophage: Mrc1
-   Microglia: C1qa
-   Endothelial: Flt1
-   Ependymal: Foxj1

### Violin Plot & Features plot

```{r}

top1 <- cluster.all.markers %>%
  group_by(cluster) %>%
  dplyr::filter(avg_log2FC > FC) %>%
  slice_head(n = 1) %>%
  ungroup()
# Cerebellum cell types
features = c("Ppp1r17", "Gabra6", "Eomes", "Lypd6", "Prkcd", "Klhl1", "Lgi2", "Gdf10", "Aqp4", "Mobp", "Ppfibp1", "Dcn", "Kcnj8", "Ttr", "Mrc1", "C1qa", "Flt1", "Foxj1")
# Brain cell types ChatGPT
features = c("Calb1", "Gabra6", "Pvalb", "Sst", "Calb2", "Tph2",  "Gfap", "Plp1","Mbp","Aif1", "Cdk11b", "Cx3cr1", "Pecam1", "Cldn5")

# Brain Cell types FRP
features = c("Slc17a7", "Lamp5", "Cux2", "Ndst4", "Rorb", "Tox", "Hs3st2", "Tle4", "Fexf2", "Gad1", "Pvalb", "Sst", "Vip", "Cacng4", "Siglech", "Cx3cr1", "Aqp4", "Cldn5")

features = c("Ppp1r17", "Gabra6", "Eomes", "Lypd6", "Prkcd", "Klhl1", "Lgi2", "Gdf10", "Aqp4", "Mobp", "Ppfibp1", "Dcn", "Kcnj8", "Ttr", "Mrc1", "C1qa", "Flt1", "Foxj1", "Calb1", "Pvalb", "Sst", "Calb2", "Tph2",  "Gfap", "Plp1","Mbp","Aif1", "Cdk11b", "Cx3cr1", "Pecam1", "Cldn5", "Slc17a7", "Lamp5", "Cux2", "Ndst4", "Rorb", "Tox", "Hs3st2", "Tle4", "Fexf2", "Gad1",  "Vip", "Cacng4", "Siglech")

features_presents <- features[features %in% rownames(data_select)]

for(feature in features_presents){
  #VlnPlot
  plot<- VlnPlot(data_select, features = feature, split.by = "condition")
  print(plot)
  save_plot(paste0("Vln_plot_",feature), plot)
  #UMAP by features
  feature_plot <- FeaturePlot(data_select, features = feature)
  print(feature_plot)
  save_plot("Feature_plot_Markers", feature_plot,width = 8, height = 6)
}
```

## DotPlot

```{r}

dotplot <- DotPlot(data_select, features = features_presents, cols = c("blue", "red"), dot.scale = 8, split.by = "condition") +
    RotatedAxis()
print(dotplot)
save_plot("DotPlot_features", dotplot, width = 16 ,height = 18)
```

# 8. Saving

```{r}
#Save Normalized and dimentionally reduced data
tic()
# Export results ####
# Save using the name
  save(data_select, file = paste0(output_path, "RData/", "data_select", ".RData"))  

### Load RDatas

# Load results ####
load(paste0(output_path,"RData/data_select.RData"))
toc()

gc()
```

# 9. Automatic Clustering association

1.  Use 'SingleR' is a widely used tool that compares gene expression profiles of your clusters to predefined reference databases. Load an appropriate reference database. For example:

-   MouseRNAseqData: Reference for murine tissues.
-   ImmGenData: Immunological reference.
-   HumanPrimaryCellAtlasData: For human cells.

2.  Use scType for fast annotation 'scType' is a simpler, more customizable tool that uses user-defined or database-available marker lists.

    Comparison between 'SingleR' and 'scType'

    | Feature           | SingleR                        | scType            |
    |-------------------|--------------------------------|-------------------|
    | **Ease of use**   | More robust but more technical | Simpler           |
    | **Database**      | Use predefined references      | Use marker lists  |
    | **Customization** | Limited                        | Very customizable |
    | **Speed**         | Moderate                       | Very fast         |

    : Recommendation If you have a well-characterized dataset (such as mouse brain transcriptomics), use SingleR with a database such as MouseRNAseqData. If you prefer greater flexibility or speed, try scType with specific marker lists.

## Allen Brain Atlas clustering

Download matrix Gene expression matrix and Table cell metadata from Allen Brain Atlas Cell Types Database: RNA-Seq Data (<https://portal.brain-map.org/atlases-and-data/rnaseq>) Whole Cortex & Hippocampus - 10x genomics (2020) with 10x-SMART-seq taxonomy (2021)

## Generate Matrix from FindAllMarkers

```{r}
# Extraer la matriz de expresión y las identidades de los clusters
#cluster.markers <- GetAssayData(data_select, slot = "data")
#cluster.ids <- Idents(data_select)
```

```{r}

# Asegúrate de que el assay está configurado correctamente
DefaultAssay(data_select) <- "integrated"

# Extraer los genes significativos encontrados por FindAllMarkers
significant_genes <- cluster.all.markers %>% 
  dplyr::filter(p_val_adj < 0.05) %>% 
  dplyr::select(gene) %>% 
  dplyr::distinct()%>% 
  dplyr::pull(gene)
  
# Extraer la matriz de expresión para esos genes (utilizando 'layer' en lugar de 'slot')
expression_matrix <- GetAssayData(data_select, layer = "data")[significant_genes, ]

# Extraer las identidades de los clústeres
cluster.ids <- Idents(data_select)

# Crear un objeto SummarizedExperiment con la matriz de expresión
cluster.markers <- SummarizedExperiment(
  assays = list(logcounts = as.matrix(expression_matrix)),
  rowData = DataFrame(gene = rownames(expression_matrix)),
  colData = DataFrame(cluster = cluster.ids)
)



```

### Medians

```{r}
# Cargar los archivos CSV
median_data <- fread("rawdata/medians.csv")
trimmed_means_data <- fread("rawdata/trimmed_means.csv")

# Verificar las primeras filas
head(median_data)
head(trimmed_means_data)

# Revisar dimensiones
dim(median_data)
dim(trimmed_means_data)

# Convertir a matriz
median_matrix <- as.matrix(median_data[ , -1, with = FALSE])  # Asumiendo que la primera columna son los nombres de genes
rownames(median_matrix) <- median_data[[1]]  # Asignar nombres de genes a las filas

trimmed_means_matrix <- as.matrix(trimmed_means_data[ , -1, with = FALSE])  # Similar al anterior
rownames(trimmed_means_matrix) <- trimmed_means_data[[1]]

# Crear referencia de SingleR
# Crear un nuevo objeto SummarizedExperiment con logcounts
reference_median <- SummarizedExperiment(
    assays = list(logcounts = log1p(median_matrix)), # Aplicar log-transformation
    rowData = DataFrame(gene = rownames(median_matrix)),
    colData = DataFrame(label.main = colnames(median_matrix))
)
# Revisar las etiquetas
colData(reference_median)
# Confirmar que `label.main` contiene los nombres de los tipos celulares
reference_median$label.main <- colnames(median_matrix)

```

```{r}
# Comparar con la base de datos de referencia
annotations_median <- SingleR(test = cluster.markers, 
                       ref = reference_median, 
                       labels = reference_median$label.main,
                       clusters = cluster.ids)


# Revisar resultados
head(annotations_median)


```

#### Plot Median

```{r}
new.cluster.median <- annotations_median$labels
names(new.cluster.median) <- levels(data_select)
data_select_median <- RenameIdents(data_select, new.cluster.median)

# Convertir las anotaciones a un data.frame para unirlas con cluster_markers_df
annotations_median_df <- as.data.frame(annotations_median)
annotations_median_df$cluster <- rownames(annotations_median_df)  # Añadir los IDs de los clusters como columna

# Añadir columna 'labels'
cell_types_median <- annotations_median_df$labels
cluster_markers_df$labels_median <- cell_types_median[cluster_markers_df$cluster + 1]

```

```{r}
# Visualiza con las nuevas identidades
plot <- DimPlot(data_select_median, reduction = "umap", label = TRUE, repel = TRUE)
print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_UMAP_median_cluster_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  image_number <<- image_number + 1  # Use <<- to update the external variable

```

### Trimmed_Means

```{r}

reference_trimmed <- SummarizedExperiment(
    assays = list(logcounts = log1p(trimmed_means_matrix)), # Aplicar log-transformation
    rowData = DataFrame(gene = rownames(trimmed_means_matrix)),
    colData = DataFrame(label.main = colnames(trimmed_means_matrix))
)
# Revisar las etiquetas
colData(reference_trimmed)

# Confirmar que `label.main` contiene los nombres de los tipos celulares
reference_trimmed$label.main <- colnames(trimmed_means_matrix)
```

```{r}
# Comparar con la base de datos de referencia
annotations_trimmed <- SingleR(test = cluster.markers, 
                       ref = reference_trimmed, 
                       labels = reference_trimmed$label.main,
                       clusters = cluster.ids)


# Revisar resultados
head(annotations_trimmed)
```

#### Plot Trimmed

```{r}
new.cluster.trimmed <- annotations_trimmed$labels
names(new.cluster.trimmed) <- levels(data_select)
data_select_trimmed <- RenameIdents(data_select, new.cluster.trimmed)
# Convertir las anotaciones a un data.frame para unirlas con cluster_markers_df
annotations_trimmed_df <- as.data.frame(annotations_trimmed)
annotations_trimmed_df$cluster <- rownames(annotations_trimmed_df)  # Añadir los IDs de los clusters como columna

# Añadir columna 'labels'
cell_types_trimmed <- annotations_trimmed_df$labels
cluster_markers_df$labels_trimmed <- cell_types_trimmed[cluster_markers_df$cluster + 1]
```

```{r}
# Visualiza con las nuevas identidades
plot <- DimPlot(data_select_trimmed, reduction = "umap", label = TRUE, repel = TRUE)
print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_UMAP_trimmed_cluster_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  image_number <<- image_number + 1  # Use <<- to update the external variable

```

## MouseRNAseqData

```{r}


ref <- MouseRNAseqData()  # Change according needs
# Extraer la matriz de expresión y las identidades de los clusters
cluster.markers <- GetAssayData(data_select, slot = "data")
cluster.ids <- Idents(data_select)

# Comparar con la base de datos de referencia
annotations_ms <- SingleR(test = cluster.markers, 
                       ref = ref, 
                       labels = ref$label.main,
                       clusters = cluster.ids)

# Ver los resultados
head(annotations_ms)
```

```{r}
new.cluster.ids <- annotations_ms$labels
names(new.cluster.ids) <- levels(data_select)
data_select_ms <- RenameIdents(data_select, new.cluster.ids)
# Convertir las anotaciones a un data.frame para unirlas con cluster_markers_df
annotations_ms_df <- as.data.frame(annotations_ms)
annotations_ms_df$cluster <- rownames(annotations_ms_df)  # Añadir los IDs de los clusters como columna
# Añadir columna 'labels'
cell_types_ms <- annotations_ms_df$labels
cluster_markers_df$labels_ms <- cell_types_ms[cluster_markers_df$cluster + 1]
```

```{r}
# Visualiza con las nuevas identidades
plot <- DimPlot(data_select_ms, reduction = "umap", label = TRUE, repel = TRUE)
print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "UMAP_simple_cluster_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  image_number <<- image_number + 1  # Use <<- to update the external variable

```

## Cerebellum types

Taken from supplementary material from: Kozareva, V., Martin, C., Osorno, T.
et al.
A transcriptomic atlas of mouse cerebellar cortex comprehensively defines cell types.
Nature 598, 214–219 (2021).
<https://doi.org/10.1038/s41586-021-03220-z>

Supplementary Table 2: Highly differentially expressed genes across all clusters.
Genes are limited to top 500 differentially expressed genes per subtype.
Column names represent the following: “avgExpr” is the mean gene expression in that subtype; “logFC” is the log fold change between cells within the subtype compared to remaining cells; “pval” is the p-value, calculated with the two-tailed Gaussian approximation of U statistic; “padj” is the Benjamini-Hochberg adjusted p-value; “pct_in” is the percent of cells in the subtype expressing the gene; “pct_out” is the percent of cells outside the subtype expressing the gene.

```{r}
# Leer el archivo Excel (asumiendo que la primera hoja contiene los datos)
file_path <- "rawdata/cb_cell_types.xlsx"
data <- read_excel(file_path)

# Transformar los datos: agrupar subtipos similares y pivotar
cerebellum_data <- data %>%
  dplyr::select(gene, subtype, avgExpr) %>%  # Seleccionar solo las columnas necesarias
  dplyr::mutate(subtype = str_replace(subtype, "_\\d+$", "")) %>%  # Eliminar sufijos _1, _2, etc.
  dplyr::group_by(gene, subtype) %>%  # Agrupar por gen y subtipo
  dplyr::summarise(avgExpr = sum(avgExpr, na.rm = TRUE), .groups = "drop") %>%  # Sumar valores de expresión
  pivot_wider(
    names_from = subtype,  # Usar la columna "subtype" como nombres de columnas
    values_from = avgExpr,  # Usar los valores de la columna "avgExpr" como contenido de las celdas
    values_fill = list(avgExpr = 0)  # Asignar 0 a los valores NA
  )%>%
  dplyr::mutate(Purkinje = coalesce(Purkinje_Aldoc, 0) + coalesce(Purkinje_Anti_Aldoc, 0)) %>%  # Combinar columnas Purkinje_Aldoc y Purkinje_Anti_Aldoc
  dplyr::select(-Purkinje_Aldoc, -Purkinje_Anti_Aldoc)  # Eliminar las columnas originales
# Verificar las primeras filas
head(cerebellum_data)
# Revisar dimensiones
dim(cerebellum_data)

# Convertir a matriz
cerebellum_matrix <- as.matrix(cerebellum_data[ , -1, with = FALSE])  # Asumiendo que la primera columna son los nombres de genes
rownames(cerebellum_matrix) <- cerebellum_data[[1]]  # Asignar nombres de genes a las filas

# Crear referencia de SingleR
# Crear un nuevo objeto SummarizedExperiment con logcounts
reference_cerebellum <- SummarizedExperiment(
    assays = list(logcounts = log1p(cerebellum_matrix)), # Aplicar log-transformation
    rowData = DataFrame(gene = rownames(cerebellum_matrix)),
    colData = DataFrame(label.main = colnames(cerebellum_matrix))
)
# Revisar las etiquetas
colData(reference_cerebellum)

# Confirmar que `label.main` contiene los nombres de los tipos celulares
reference_cerebellum$label.main <- colnames(cerebellum_matrix)

```

```{r}
# Comparar con la base de datos de referencia
annotations_cerebellum <- SingleR(test = cluster.markers, 
                       ref = reference_cerebellum, 
                       labels = reference_cerebellum$label.main,
                       clusters = cluster.ids)


# Revisar resultados
head(annotations_cerebellum)


plot <- plotScoreHeatmap(annotations_cerebellum)
print(plot)
save_plot("Cerebellum_annotation", plot)
```

```{r}
new.cluster.cerebellum <- annotations_cerebellum$labels
names(new.cluster.cerebellum) <- levels(data_select)
data_select_cerebellum <- RenameIdents(data_select, new.cluster.cerebellum)

# Convertir las anotaciones a un data.frame para unirlas con cluster_markers_df
annotations_cerebellum_df <- as.data.frame(annotations_cerebellum)
annotations_cerebellum_df$cluster <- rownames(annotations_cerebellum_df)  # Añadir los IDs de los clusters como columna

cell_types_cerebellum <- annotations_cerebellum_df$labels
cluster_markers_df$labels_cerebellum <- cell_types_cerebellum[cluster_markers_df$cluster + 1]

```

```{r}
# Visualiza con las nuevas identidades
plot <- DimPlot(data_select_cerebellum, reduction = "umap", label = TRUE, repel = TRUE)
print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_UMAP_cerebellum_cluster_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  image_number <<- image_number + 1  # Use <<- to update the external variable

```

## Save Clusters

```{r}
#Generate a list 
cluster_list <- cluster_markers_df %>%
  group_split(cluster) %>%
  set_names(paste0("cluster_", unique(cluster_markers_df$cluster)))
#Save xlsx
cluster_list %>% 
write_xlsx(paste0(output_path,"tables/","cluster_list.xlsx"))
#Save RData

# Save using the name
save(cluster_list, file = paste0(output_path, "RData/", "cluster_list", ".RData"))
```

```{r}
#Clean enviroment and memory
# Especifica los nombres de los objetos que deseas conservar
keep_objects <- c("cluster_list", "image_number", "output_path", "data_select", "cluster.all.markers")
# Elimina todos los objetos excepto los especificados
rm(list = setdiff(ls(), keep_objects))
gc() #Clean memory
# Set up parameters
set.seed(123)
options(scipen = -1)  # Habilitar notación científica en R  
source("code/global_variables.R")  
### Load RDatas
# Load results ####
load(paste0(output_path,"RData/cluster_list.RData"))

```

Now, cluster_list will be used as input for the next steps.

# 9. Vulcano Plots

```{r}
filtration <- 0
source("code/02_vulcano_plots.R")

```

# 10. GO

```{r}

source("code/001_GO.R")
```

# 11. Strings

```{r}
source("code/003_strings.R")
```

# 12. GSE

```{r}
source("code/004_gse.R")
```

#13.
KEGG

```{r}
source("code/09_gseKEGG.R")
```

# 14. HeatMaps

```{r}

# Gene list
gene_list <- c("Hmgcr", "Srebf2", "Apoe","Lrp1", "Lrp1b", "Abca1")

source("code/06_Heatmap.R")
```

# 15. HeatMaps GO Types

```{r}
#source("code/07_HeatMap_GO_types.R")
```

# 16. EnrichR

```{r}
source("code/08_EnrichR.R")
```
