

# 0. Loading Libraries

## Loading functions
```{r}
output_path <- "output/SC_UbCRE/SCTransform/"
source("code/global_variables.R")
```


```{r}
source("code/00_packages.R")
```

# 1. Importing data
We start by reading in the data. The Read10X() function reads in the output of the cellranger pipeline from 10X, returning a unique molecular identified (UMI) count matrix. The values in this matrix represent the number of molecules for each feature (i.e. gene; row) that are detected in each cell (column). Note that more recent versions of cellranger now also output using the h5 file format, which can be read in using the Read10X_h5() function in Seurat.
```{r}
tic()
#import data
# Load the PBMC dataset
ubc_c1 <- Seurat::Read10X(data.dir = "rawdata/SC/AWGC-1_Ctrl-1_results")
ubc_c2 <- Seurat::Read10X(data.dir = "rawdata/SC/AWGC-2_Ctrl-2_results")
ubc_ko1 <- Seurat::Read10X(data.dir = "rawdata/SC/AWGC-3_KOut-1_results")
ubc_ko2 <- Seurat::Read10X(data.dir = "rawdata/SC/AWGC-4_KOut-2_results")

# Initialize the Seurat object with the raw (non-normalized data).
ubc_c1 <- CreateSeuratObject(counts = ubc_c1, project = "wt_1", min.cells = 3, min.features = 200)
ubc_c2 <- CreateSeuratObject(counts = ubc_c2, project = "wt_2", min.cells = 3, min.features = 200)
ubc_ko1 <- CreateSeuratObject(counts = ubc_ko1, project = "ko_1", min.cells = 3, min.features = 200)
ubc_ko2 <- CreateSeuratObject(counts = ubc_ko2, project = "ko_2", min.cells = 3, min.features = 200)

toc()
```

Create vector with SeurtalObjects names for automatization

```{r}
seurat_names <- c("ubc_c1", "ubc_c2", "ubc_ko1", "ubc_ko2")
```

```{r}
tic()
# Export results ####
lapply(seurat_names, function(x) {
  obj <- get(x)  # Retrieve the object
  save(obj, file = paste0(output_path, "RData/", x, ".RData"))
})
toc()
```

# 2. Data preprocessing for each library

## Percentage Features
```{r}
# The [[ operator can add columns to object metadata. This is a great place to stash QC stats
lapply(seurat_names, function(x) {
  obj <- get(x)  # Retrieve the Seurat object
  # store mitochondrial percentage in object meta data
  obj <- PercentageFeatureSet(obj, pattern = "^MT-", col.name = "percent.mt")
  assign(x, obj, envir = .GlobalEnv)  # Reassign the modified object back to the global environment
})
```

# Increase memory limit for future package
```{r}
# Increase memory limit for future package
options(future.globals.maxSize = 16000 * 1024^2) # 16 GB limit

library(future)
plan("multisession", workers = 4) # Use 4 workers
```

## Perform SCTransform
```{r}
# The [[ operator can add columns to object metadata. This is a great place to stash QC stats
lapply(seurat_names, function(x) {
  obj <- get(x)  # Retrieve the Seurat object
  # run sctransform
  obj <- SCTransform(obj, vars.to.regress = "percent.mt", verbose = FALSE)
  assign(x, obj, envir = .GlobalEnv)  # Reassign the modified object back to the global environment
})
```

# 3. Find Integration Anchors
After transforming the data, identify integration anchors. Use the transformed Seurat objects for this step.

```{r}
tic()
gc()
# Paso 1: Seleccionar características de integración comunes
integration_features <- SelectIntegrationFeatures(
  object.list = list(ubc_c1, ubc_c2, ubc_ko1, ubc_ko2),
  nfeatures = 3000 # Ajusta según sea necesario
)

# Paso 2: Preparar los objetos para la integración
object.list <- PrepSCTIntegration(
  object.list = list(ubc_c1, ubc_c2, ubc_ko1, ubc_ko2),
  anchor.features = integration_features
)
gc()
# Paso 3: Encontrar los anclajes de integración
integration_anchors <- FindIntegrationAnchors(
  object.list = object.list,
  normalization.method = "SCT",
  anchor.features = integration_features,
  dims = 1:20,
  verbose = FALSE
)
toc()
```

Use the identified anchors to integrate the datasets.
```{r}
# Increase memory limit for future package
options(future.globals.maxSize = 28000 * 1024^2) # 16 GB limit
tic()
sct_data <- IntegrateData(
  anchorset = integration_anchors,
  normalization.method = "SCT",
  verbose = TRUE,
  dims = 1:20, # Reduce el rango de dimensiones
)
toc()

```

# 4.Linear dimensional reduction

Next we perform PCA on the scaled data. By default, only the previously
determined variable features are used as input, but can be defined using
features argument if you wish to choose a different subset (if you do
want to use a custom subset of features, make sure you pass these to
ScaleData first).

For the first principal components, Seurat outputs a list of genes with
the most positive and negative loadings, representing modules of genes
that exhibit either correlation (or anti-correlation) across
single-cells in the dataset.

```{r}
tic()

#Add PCA 

  sct_data <- RunPCA(sct_data, features = VariableFeatures(object = sct_data))
  
  # Examine and visualize PCA results a few different ways
  print(sct_data[["pca"]], dims = 1:5, nfeatures = 5)
  
  plot <- VizDimLoadings(sct_data, dims = 1:2, reduction = "pca")
  print(plot)
 # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_VimPCA_Loadings")
  
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 8, height = 6, units = "in")
  
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
  
toc()

```

## Elbow Plot

Determine the ‘dimensionality’ of the dataset To overcome the extensive
technical noise in any single feature for scRNA-seq data, Seurat
clusters cells based on their PCA scores, with each PC essentially
representing a ‘metafeature’ that combines information across a
correlated feature set. The top principal components therefore represent
a robust compression of the dataset. However, how many components should
we choose to include? 10? 20? 100?

In Macosko et al, we implemented a resampling test inspired by the
JackStraw procedure. While still available in Seurat (see previous
vignette), this is a slow and computationally expensive procedure, and
we is no longer routinely used in single cell analysis.

An alternative heuristic method generates an ‘Elbow plot’: a ranking of
principle components based on the percentage of variance explained by
each one (ElbowPlot() function). In this example, we can observe an
‘elbow’ around PC9-10, suggesting that the majority of true signal is
captured in the first 10 PCs.

```{r}
tic()

  #ElbowPlot
  plot <- ElbowPlot(sct_data)
  print(plot)
 # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_ElbowPlot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 8, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
  
toc()


```

## Cluster cells

Seurat applies a graph-based clustering approach, building upon initial
strategies in (Macosko et al). Importantly, the distance metric which
drives the clustering analysis (based on previously identified PCs)
remains the same. However, our approach to partitioning the cellular
distance matrix into clusters has dramatically improved. Our approach
was heavily inspired by recent manuscripts which applied graph-based
clustering approaches to scRNA-seq data [SNN-Cliq, Xu and Su,
Bioinformatics, 2015] and CyTOF data [PhenoGraph, Levine et al., Cell,
2015]. Briefly, these methods embed cells in a graph structure - for
example a K-nearest neighbor (KNN) graph, with edges drawn between cells
with similar feature expression patterns, and then attempt to partition
this graph into highly interconnected ‘quasi-cliques’ or ‘communities’.

As in PhenoGraph, we first construct a KNN graph based on the euclidean
distance in PCA space, and refine the edge weights between any two cells
based on the shared overlap in their local neighborhoods (Jaccard
similarity). This step is performed using the FindNeighbors() function,
and takes as input the previously defined dimensionality of the dataset
(first 10 PCs).

To cluster the cells, we next apply modularity optimization techniques
such as the Louvain algorithm (default) or SLM [SLM, Blondel et al.,
Journal of Statistical Mechanics], to iteratively group cells together,
with the goal of optimizing the standard modularity function. The
FindClusters() function implements this procedure, and contains a
resolution parameter that sets the ‘granularity’ of the downstream
clustering, with increased values leading to a greater number of
clusters. We find that setting this parameter between 0.4-1.2 typically
returns good results for single-cell datasets of around 3K cells.
Optimal resolution often increases for larger datasets. The clusters can
be found using the Idents() function.

```{r}
sct_data <- FindNeighbors(sct_data, dims = 1:11)
sct_data <- FindClusters(sct_data, resolution = 0.5)
```

## UMAP dimensions

```{r}
tic()
  #UMAP
  sct_data <- RunUMAP(sct_data, dims = 1:11)
toc()
```

## Plot UMAP & UMAP Features

```{r}
tic()
 # note that you can set `label = TRUE` or use the LabelClusters function to help label individual clusters
  plot <- DimPlot(sct_data, reduction = "umap", label = TRUE)

  print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_UMAP")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 8, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
    


# note that you can set `label = TRUE` or use the LabelClusters function to help label individual clusters
  plot <- DimPlot(sct_data, reduction = "umap", label = TRUE)
  plot <- DimPlot(sct_data, reduction = "umap", split.by="orig.ident", ncol = 2)
  
  print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_UMAP_ident")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 8, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
  
  
  #UMAP by features
  features <- c("nFeature_RNA", "nCount_RNA")
  feature_plot <- FeaturePlot(sct_data, features = features)
    print(feature_plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_Feature_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), feature_plot, width = 8, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), feature_plot, width = 8, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
toc()

```

## UMAP HVGs

```{r}
top <-head(VariableFeatures(sct_data),10)

  #UMAP by features
  feature_plot <- FeaturePlot(sct_data, features = top, ncol = 5)
    print(feature_plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_Feature_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), feature_plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), feature_plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
  
```

## UMAP Dnajc5

```{r}
  #UMAP by features
  feature_plot <- FeaturePlot(sct_data, features = "Dnajc5",split.by="orig.ident", ncol = 2 )
    print(feature_plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "_Feature_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), feature_plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), feature_plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
  
```

```{r}
tic()
SaveH5Seurat(sct_data,
             filename = paste0(output_path, "RData/","UBC_experiment.h5Seurat"),
             overwrite = TRUE, 
             verbose = TRUE)
toc()
```

#5. Finding differentially expressed features (cluster biomarkers)
Seurat can help you find markers that define clusters via differential
expression (DE). By default, it identifies positive and negative markers
of a single cluster (specified in ident.1), compared to all other cells.
FindAllMarkers() automates this process for all clusters, but you can
also test groups of clusters vs. each other, or against all cells.

In Seurat v5, we use the presto package (as described here and available
for installation here), to dramatically improve the speed of DE
analysis, particularly for large datasets. For users who are not using
presto, you can examine the documentation for this function
(?FindMarkers) to explore the min.pct and logfc.threshold parameters,
which can be increased in order to increase the speed of DE testing.

```{r}
# find markers for every cluster compared to all remaining cells, report only the positive
# ones
# Encuentra marcadores de todos los clusters
cluster.markers <- FindAllMarkers(sct_data, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)

sct_data.markers <- FindAllMarkers(sct_data, 
                              logfc.threshold = 0.1,
                              min.pct = 0.1,
                              only.pos = TRUE)
```

```{r}
#Generate a list 
cluster_list <- sct_data.markers %>%
  group_split(cluster) %>%
  set_names(paste0("cluster_", unique(sct_data.markers$cluster)))

cluster_list %>% 
write_xlsx(paste0(output_path,"tables/","cluster_list.xlsx"))
```

## HeatMap

```{r}
sct_data.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10


plot <- DoHeatmap(sct_data, features = top10$gene) 
print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "HeatMap_cluster_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable
```

# 6. Clustering association

1.  Use 'SingleR' is a widely used tool that compares gene expression
    profiles of your clusters to predefined reference databases. Load an
    appropriate reference database. For example:

-   MouseRNAseqData: Reference for murine tissues.
-   ImmGenData: Immunological reference.
-   HumanPrimaryCellAtlasData: For human cells.

2.  Use scType for fast annotation 'scType' is a simpler, more
    customizable tool that uses user-defined or database-available
    marker lists.

    Comparison between 'SingleR' and 'scType'
    
    | Feature           | SingleR                         | scType            |
    |-------------------|---------------------------------|-------------------|
    | **Ease of use**   | More robust but more technical  |  Simpler          |
    | **Database**      | Use predefined references       | Use marker lists  |
    | **Customization** | Limited                         | Very customizable |
    | **Speed**         | Moderate                        |  Very fast        |

    : Recommendation If you have a well-characterized dataset (such as
    mouse brain transcriptomics), use SingleR with a database such as
    MouseRNAseqData. If you prefer greater flexibility or speed, try
    scType with specific marker lists.


## Allen Brain Atlas clustering
Download matrix Gene expression matrix and Table cell metadata from Allen Brain Atlas Cell Types Database: RNA-Seq Data (https://portal.brain-map.org/atlases-and-data/rnaseq)
Whole Cortex & Hippocampus - 10x genomics (2020)
with 10x-SMART-seq taxonomy (2021)
```{r}

# Extraer la matriz de expresión y las identidades de los clusters
cluster.markers <- GetAssayData(data_merge, slot = "data")
cluster.ids <- Idents(data_merge)

# Cargar los archivos CSV
median_data <- fread("rawdata/medians.csv")
trimmed_means_data <- fread("rawdata/trimmed_means.csv")

# Verificar las primeras filas
head(median_data)
head(trimmed_means_data)

# Revisar dimensiones
dim(median_data)
dim(trimmed_means_data)

# Convertir a matriz
median_matrix <- as.matrix(median_data[ , -1, with = FALSE])  # Asumiendo que la primera columna son los nombres de genes
rownames(median_matrix) <- median_data[[1]]  # Asignar nombres de genes a las filas

trimmed_means_matrix <- as.matrix(trimmed_means_data[ , -1, with = FALSE])  # Similar al anterior
rownames(trimmed_means_matrix) <- trimmed_means_data[[1]]

# Crear referencia de SingleR
# Crear un nuevo objeto SummarizedExperiment con logcounts
reference_median <- SummarizedExperiment(
    assays = list(logcounts = log1p(median_matrix)), # Aplicar log-transformation
    rowData = DataFrame(gene = rownames(median_matrix)),
    colData = DataFrame(label.main = colnames(median_matrix))
)
# Revisar las etiquetas
colData(reference_median)

# Confirmar que `label.main` contiene los nombres de los tipos celulares
reference_median$label.main <- colnames(median_matrix)

reference_trimmed <- SummarizedExperiment(
    assays = list(logcounts = log1p(trimmed_means_matrix)), # Aplicar log-transformation
    rowData = DataFrame(gene = rownames(trimmed_means_matrix)),
    colData = DataFrame(label.main = colnames(trimmed_means_matrix))
)
# Revisar las etiquetas
colData(reference_trimmed)

# Confirmar que `label.main` contiene los nombres de los tipos celulares
reference_trimmed$label.main <- colnames(trimmed_means_matrix)
```


```{r}
# Comparar con la base de datos de referencia
annotations_median <- SingleR(test = cluster.markers, 
                       ref = reference_median, 
                       labels = reference_median$label.main,
                       clusters = cluster.ids)


# Revisar resultados
head(annotations_median)
```


```{r}
# Comparar con la base de datos de referencia
annotations_trimmed <- SingleR(test = cluster.markers, 
                       ref = reference_trimmed, 
                       labels = reference_trimmed$label.main,
                       clusters = cluster.ids)


# Revisar resultados
head(annotations_trimmed)
```

## Plot Median
```{r}
new.cluster.median <- annotations_median$labels
names(new.cluster.median) <- levels(data_merge)
data_merge_median <- RenameIdents(data_merge, new.cluster.median)
```


```{r}
# Visualiza con las nuevas identidades
plot <- DimPlot(data_merge_median, reduction = "umap", label = TRUE, repel = TRUE)
print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "UMAP_simple_cluster_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable

```

## Plot Trimmed
```{r}
new.cluster.trimmed <- annotations_trimmed$labels
names(new.cluster.trimmed) <- levels(data_merge)
data_merge_trimmed <- RenameIdents(data_merge, new.cluster.median)
```


```{r}
# Visualiza con las nuevas identidades
plot <- DimPlot(data_merge_trimmed, reduction = "umap", label = TRUE, repel = TRUE)
print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "UMAP_simple_cluster_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable

```


## MouseRNAseqData
```{r}


ref <- MouseRNAseqData()  # Change according needs
# Extraer la matriz de expresión y las identidades de los clusters
cluster.markers <- GetAssayData(data_merge, slot = "data")
cluster.ids <- Idents(data_merge)

# Comparar con la base de datos de referencia
annotations_ms <- SingleR(test = cluster.markers, 
                       ref = ref, 
                       labels = ref$label.main,
                       clusters = cluster.ids)

# Ver los resultados
head(annotations)
```


```{r}
new.cluster.ids <- annotations_ms$labels
names(new.cluster.ids) <- levels(data_merge)
data_merge_ms <- RenameIdents(data_merge, new.cluster.ids)
```


```{r}
# Visualiza con las nuevas identidades
plot <- DimPlot(data_merge_ms, reduction = "umap", label = TRUE, repel = TRUE)
print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", figure_counter), "UMAP_simple_cluster_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  figure_counter <<- figure_counter + 1  # Use <<- to update the external variable

```

