---
title: "SingleCell RNA seq general pipeline"
author: "Santiago López Begines"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: tango
    fig_width: 10
    fig_height: 8
    fig_caption: true
    code_folding: hide
    number_sections: true
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
editor_options:
  markdown:
    wrap: sentence
---

# 0. Setup

## Environment and global parameters

```{r env_setup}
rm(list = ls())   # Clean environment
gc()              # Release memory

# --- Global parameters --------------------------------------------------------
image_number <- 1
#output_path  <- "output/GSE262881/SCT"
output_path <- "output/GSE262881/LogNorm"
# Data paths
DATA_DIR  <- "../rawdata/GSE262881_RAW"

# Random seed for reproducibility
set.seed(123)
options(scipen = 3, digits = 3)

# Increase memory limit for future package
options(future.globals.maxSize = 14000 * 1024^2) # 16 GB limit
# Memory: sequential plan (i7-7560U: 2 physical cores, 16 GB RAM)
# Parallel overhead outweighs benefits on 2-core hardware; protects RAM
library(future)
#plan("multisession", workers = 4) # Use 4 workers
plan("sequential") # Sequential to save RAM
```

## Packages

```{r packages}
#| message: false
#| warning: false  
# renv::restore()  # Uncomment to restore packages from renv.lock
source("../code/00_packages.R")
# renv::snapshot()  # Snapshot current package state to renv.lock (uncomment if changes made)
```


```{r load_globals}
source("../code/global_variables.R")
source("../code/01_aux_functions.R")
source("../code/02_sc_functions.R")
```

Setup Logger
```{r}
# --- Logging ------------------------------------------------------------------
logger   <- setup_logging("Logger")
log_msg  <- logger$log_msg
log_file <- logger$log_file
on.exit(logger$close_log(), add = TRUE)

log_msg("Starting Merging & Normalization — RAM: ", ram_mb(), " MB")
```


# 1. Input data

**Dataset:** GSE262881 — Wang et al., BMC Biology (2025) 23:124  
**Paper:** A single-cell transcriptomic atlas of all cell types in the brain of 5xFAD Alzheimer mice in response to dietary inulin supplementation

---

## Full Processing Workflow (per Methods, p. 19)

```
FASTQ  →  CellRanger 7.1.0  →  raw_feature_bc_matrix/
                                        ↓
                              CellBender 0.2.2
                                        ↓
                    *_CellBender_feature_bc_matrix_filtered.h5   ← .h5 files on GEO

FASTQ  →  CellRanger 7.1.0 (BAM)  →  velocyto 0.17.17
                                        ↓
                                    *.loom.gz   ← .loom files on GEO
```

We start by reading in the data. The Read10X() function reads in the
output of the cellranger pipeline from 10X, returning a unique molecular
identified (UMI) count matrix. The values in this matrix represent the
number of molecules for each feature (i.e. gene; row) that are detected
in each cell (column). Note that more recent versions of cellranger now
also output using the h5 file format, which can be read in using the
Read10X_h5() function in Seurat.

Data from Wang et al., BMC Biology (2025) 23:124 comes from CellBender filtering. The paper check 2 conditions, Control vs. Insulin; in 4 different parts of brain: we will center just in one region, Forebrain, as example. More regions (translated in more libraries demand more computational power).


## Read 10X matrices (with checkpoint)

Each library is read and saved individuallyto avoid holding all count matrices in RAM simultaneously.

```{r import_raw, message=TRUE}
# define output_path y CHECKPOINT_PREFIX in global_variables                                                                                                    

# Checkpoint: all raw Seurat objects
# With checkpoint (recomended — 24 h5 file would take longer)    
  #seurat_list <- load_h5_samples(data_path, checkpoint = "01_raw_data")      

# Only desired region, with checkpoint
 # seurat_list <- load_h5_samples(DATA_DIR, regions = "forebrain", checkpoint = "01_raw_data")
# Use BPCells library for high efficiency, with checkpoint
seurat_list <- load_h5_samples_bpcells(data_path = DATA_DIR, regions = "forebrain", checkpoint = "01_raw_data_bp", specie = "mouse")

gc()
```


## Import summary

```{r import_summary}
# Quick check of matching quality (cells found in metadata)
import_tbl <- do.call(rbind, lapply(names(seurat_list), function(nm) {
  obj <- seurat_list[[nm]]
  n_total     <- ncol(obj)
  data.frame(Library = nm, N_Cells = n_total,
             row.names = NULL)
}))

print(import_tbl)
cat("\nTotal barcodes imported:", sum(import_tbl$N_Cells), "\n")
```
# 2. QC Metrics Calculation

Add human-specific QC metrics.
Key difference from mouse data: mitochondrial genes use **`^MT-`** (uppercase).

```{r add_qc_metrics}
if (check_checkpoint("02_seurat_qc")) {
  seurat_qc <- load_checkpoint("02_seurat_qc")
} else {
    if (!exists("seurat_list"))
      seurat_list <- load_checkpoint("01_raw_data_bp")
    seurat_qc <- lapply(seurat_list, function(x) {
      add_qc_metrics(x, pattern_mito = mito_pattern, pattern_ribo = ribo_pattern)
    })
    rm(seurat_list)
    gc()
    save_checkpoint(seurat_qc, "02_seurat_qc")
  }


```
## Pre-filter QC overview (combined across all libraries)
```{r qc_combined_meta}
combined_meta_prefilter <- build_combined_meta(seurat_qc)
cat("Total barcodes with QC metrics:", nrow(combined_meta_prefilter), "\n")
cat("Libraries:", length(unique(combined_meta_prefilter$library)), "\n")

# Save metadata_prefilter for downstream use
save_checkpoint(combined_meta_prefilter, "qc_combined_meta_prefilter")
#load_checkpoint("03_qc_combined_meta_prefilter") # Load to free up RAM from seurat_qc)
plots_prefilter <- list()  # Initialize list to store pre-filter QC plots
plots_filter <- list()     # Initialize list to store post-filter QC plots
```

### Distribution of features
#### nCount_RNA distribution per library (pre-filter)
```{r}
ncount_rna_prefilter <- ggplot(combined_meta_prefilter) +
  aes(x = "", y = nCount_RNA, fill = library) +
  geom_violin(adjust = 1L) +
  geom_jitter(alpha = 0.01, size=0.02) +
  scale_fill_hue(direction = 1) +
  ggthemes::theme_few() +
  facet_wrap(vars(library))+
  #remove legend
  theme(legend.position = "none") +
  labs(x = "", y = "nCount_RNA")

print(ncount_rna_prefilter)
plots_prefilter[["ncount_rna_prefilter"]] <- ncount_rna_prefilter

rm(ncount_rna_prefilter)

```
#### nFeature_RNA distribution per library (pre-filter)
```{r}
nFeature_RNA_prefilter <- ggplot(combined_meta_prefilter) +
  aes(x = "", y = nFeature_RNA, fill = library) +
  geom_violin(adjust = 1L) +
  geom_jitter(alpha = 0.01, size=0.02) +
  scale_fill_hue(direction = 1) +
  ggthemes::theme_few() +
  facet_wrap(vars(library))+
  #remove legend
  theme(legend.position = "none") +
  labs(x = "", y = "nFeature_RNA")

print(nFeature_RNA_prefilter)
plots_prefilter[["nFeature_RNA_prefilter"]] <- nFeature_RNA_prefilter

rm(nFeature_RNA_prefilter)

```
#### percent.MT distribution per library (pre-filter)
```{r}
percent.MT_prefilter <- ggplot(combined_meta_prefilter) +
  aes(x = "", y = percent.MT, fill = library) +
  geom_violin(adjust = 1L) +
  geom_jitter(alpha = 0.01, size=0.02) +
  scale_fill_hue(direction = 1) +
  ggthemes::theme_few() +
  facet_wrap(vars(library))+
  #remove legend
  theme(legend.position = "none") +
  labs(x = "", y = "percent.MT")

print(percent.MT_prefilter)
plots_prefilter[["percent.MT_prefilter"]] <- percent.MT_prefilter

rm(percent.MT_prefilter)

```
#### log10GenesPerUMI distribution per library (pre-filter)
```{r}
log10GenesPerUMI_prefilter <- ggplot(combined_meta_prefilter) +
  aes(x = "", y = log10GenesPerUMI, fill = library) +
  geom_violin(adjust = 1L) +
  geom_jitter(alpha = 0.01, size=0.02) +
  scale_fill_hue(direction = 1) +
  ggthemes::theme_few() +
  facet_wrap(vars(library))+
  #remove legend
  theme(legend.position = "none") +
  labs(x = "", y = "log10GenesPerUMI")

print(log10GenesPerUMI_prefilter)
plots_prefilter[["log10GenesPerUMI_prefilter"]] <- log10GenesPerUMI_prefilter

rm(log10GenesPerUMI_prefilter)

```
#### percent.Ribo distribution per library (pre-filter)
```{r}
percent.ribosomal_prefilter <- ggplot(combined_meta_prefilter) +
  aes(x = "", y = percent.ribosomal, fill = library) +
  geom_violin(adjust = 1L) +
  geom_jitter(alpha = 0.01, size=0.02) +
  scale_fill_hue(direction = 1) +
  ggthemes::theme_few() +
  facet_wrap(vars(library))+
  #remove legend
  theme(legend.position = "none") +
  labs(x = "", y = "percent.ribosomal")

print(percent.ribosomal_prefilter)
plots_prefilter[["percent.ribosomal_prefilter"]] <- percent.ribosomal_prefilter

rm(percent.ribosomal_prefilter)

```

### Cells per library

```{r plot_ncells_prelim}
p_cells_library <- combined_meta_prefilter %>%
  ggplot(aes(x = library, fill = condition)) +
  geom_bar() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("Number of Cells per Library") +
  xlab("Library") +
  ylab("Number of Cells") 
  # + facet_wrap(~region)


print(p_cells_library)
plots_prefilter[["p_cells_library"]] <- p_cells_library

rm(p_cells_library)
```
### UMI counts distribution

The UMI counts per cell (nCount_RNA) should generally be above 500, that
is the low end of what we expect. If UMI counts are between 500-1000
counts, it is usable but the cells probably should have been sequenced
more deeply.

```{r plot_umi_dist}
p_UMI_dist <- combined_meta_prefilter %>%
  ggplot(aes(x = nCount_RNA, color = library, fill = library)) +
  geom_density(alpha = 0.15) +
  scale_x_log10() +
  geom_vline(xintercept = QC_MIN_COUNTS, linetype = "dashed", color = "red") +
  geom_vline(xintercept = QC_MAX_COUNTS, linetype = "dashed", color = "red") +
  theme_classic(base_size = 10) +
  theme(legend.position = "none") +
  facet_wrap(~region, ncol = 2) +
  labs(title = "UMI distribution per library (pre-filter)",
       x = "nCount_RNA (log10)", y = "Density")

print(p_UMI_dist)
plots_prefilter[["p_UMI_dist"]] <- p_UMI_dist

rm(p_UMI_dist)
```

### Genes detected per cell
We have similar expectations for gene detection as for UMI detection,
although it may be a bit lower than UMIs. For high quality data, the
proportional histogram should contain a single large peak that
represents cells that were encapsulated. If we see a small shoulder to
the right of the major peak (not present in our data), or a bimodal
distribution of the cells, that can indicate a couple of things. It
might be that there are a set of cells that failed for some reason. It
could also be that there are biologically different types of cells (i.e.
quiescent cell populations, less complex cells of interest), and/or one
type is much smaller than the other (i.e. cells with high counts may be
cells that are larger in size). Therefore, this threshold should be
assessed with other metrics that we describe in this lesson.

```{r plot_genes_dis}
p_genes_dist <-  combined_meta_prefilter %>%
  ggplot(aes(x = nFeature_RNA, color = library, fill = library)) +
  geom_density(alpha = 0.15) +
  scale_x_log10() +
  geom_vline(xintercept = QC_MIN_COUNTS, linetype = "dashed", color = "red") +
  geom_vline(xintercept = QC_MAX_COUNTS, linetype = "dashed", color = "red") +
  theme_classic(base_size = 10) +
  theme(legend.position = "none") +
  facet_wrap(~region, ncol = 2) +
  labs(title = "Distribution of number of genes (pre-filter)",
       x = "nFeature_RNA", y = "Density")


print(p_genes_dist)
plots_prefilter[["p_genes_dist"]] <- p_genes_dist

rm(p_genes_dist)
```


### Mitochondrial percentage

This metric can identify whether there is a large amount of
mitochondrial contamination from dead or dying cells. We define poor
quality samples for mitochondrial counts as cells which surpass the 0.2
mitochondrial ratio mark, unless of course you are expecting this in
your sample.


```{r plot_mt_dist}
p_mt_dist <- combined_meta_prefilter %>%
  dplyr::filter(percent.MT<=5) %>%  # Filter out extreme outliers for better visualization)
  ggplot(aes(x = percent.MT, color = library, fill = library)) +
  geom_density(alpha = 0.15) +
  geom_vline(xintercept = QC_MAX_MT, linetype = "dashed", color = "red") +
  theme_classic(base_size = 10) +
  theme(legend.position = "none") +
  facet_wrap(~region, ncol = 2) +
  labs(title = paste0("Mitochondrial % (pre-filter) | threshold = ", QC_MAX_MT, "%"),
       x = "% Mitochondrial reads", y = "Density")
print(p_mt_dist)
plots_prefilter[["p_mt_dist"]] <- p_mt_dist

rm(p_mt_dist)
```


### UMIs vs. genes detected (coloured by MT%)

Two metrics that are often evaluated together are the number of UMIs and
the number of genes detected per cell. Here, we have plotted the number
of genes versus the numnber of UMIs coloured by the fraction of
mitochondrial reads. Mitochondrial read fractions are only high (light
blue color) in particularly low count cells with few detected genes.
This could be indicative of damaged/dying cells whose cytoplasmic mRNA
has leaked out through a broken membrane, and thus, only mRNA located in
the mitochondria is still conserved. These cells are filtered out by our
count and gene number thresholds. Jointly visualizing the count and gene
thresholds shows the joint filtering effect.

Cells that are poor quality are likely to have low genes and UMIs per
cell, and correspond to the data points in the bottom left quadrant of
the plot. Good cells will generally exhibit both higher number of genes
per cell and higher numbers of UMIs.

With this plot we also evaluate the slope of the line, and any scatter
of data points in the bottom right hand quadrant of the plot. These
cells have a high number of UMIs but only a few number of genes. These
could be dying cells, but also could represent a population of a low
complexity celltype (i.e red blood cells).

```{r plot_umi_vs_genes}
p_umi_vs_genes <- combined_meta_prefilter %>%
  ggplot(aes(x = nCount_RNA, y = nFeature_RNA, color = percent.MT)) +
  geom_point(size = 0.3, alpha = 0.5) +
  scale_colour_gradient(low = "grey90", high = "red") +
  scale_x_log10() + scale_y_log10() +
  geom_vline(xintercept = QC_MIN_COUNTS, linetype = "dashed") +
  geom_hline(yintercept = QC_MIN_FEATURES, linetype = "dashed") +
  facet_wrap(~region, ncol = 2) +
  theme_classic(base_size = 9) +
  labs(title = "UMIs vs Genes per cell (pre-filter, 50k sample)",
       x = "nCount_RNA (log10)", y = "nFeature_RNA (log10)")
print(p_umi_vs_genes)
plots_prefilter[["p_umi_vs_genes"]] <- p_umi_vs_genes

rm(p_umi_vs_genes)
```


### Complexity (log10 Genes per UMI)


We can see the samples where we sequenced each cell less have a higher
overall complexity, that is because we have not started saturating the
sequencing for any given gene for these samples. Outlier cells in these
samples might be cells that have a less complex RNA species than other
cells. Sometimes we can detect contamination with low complexity cell
types like red blood cells via this metric. Generally, we expect the
novelty score to be above 0.80.

```{r plot_complexity}
p_complexity <- combined_meta_prefilter %>%
  ggplot(aes(x = log10GenesPerUMI, color = library, fill = library)) +
  geom_density(alpha = 0.15) +
  geom_vline(xintercept = QC_MIN_COMPLEXITY, linetype = "dashed", color = "red") +
  theme_classic(base_size = 10) +
  theme(legend.position = "none") +
  facet_wrap(~region, ncol = 2) +
  labs(title = "Transcriptome complexity (pre-filter)",
       x = "log10(Genes / UMI)", y = "Density")

print(p_complexity)

plots_prefilter[["p_complexity"]] <- p_complexity

rm(p_complexity)
```

```{r}
base::saveRDS(plots_prefilter, file.path(output_path, "RData", "QC_plots_prefilter.rds"))
plots_prefilter <- readRDS(file.path(output_path,"RData/QC_plots_prefilter.rds"))
rm(combined_meta_prefilter)
gc()
```


# 3. Cell Filtering

Three-layer filtering strategy:

1. **Demuxlet singlet filter** — Remove barcodes classified as doublets
   by Demuxlet (included in paper metadata: `DemuxletType == "SNG"`)
2. **Study inclusion filter** — Keep only cells the authors included
   (`IncludedInStudy == TRUE`)
3. **QC thresholds** — Standard human PBMC thresholds

```{r}
cat("Filtering Parameters:\n")
cat("===========================\n")
cat("Genes per cell: Minimun: ", QC_MIN_FEATURES,  ", Maximum: ", QC_MAX_FEATURES, "\n")
cat("Genes Counts: Minimun: ", QC_MIN_COUNTS,  ", Maximum: ", QC_MAX_COUNTS, "\n")
cat("Ribosomal % Counts: Minimun: ", QC_MIN_RIBO,  ", Maximum: ", QC_MAX_RIBO, "\n")
cat("Maximum % mitochondrial reads:", QC_MAX_MT,"\n\n" )
```

```{r filter_cells}
if (check_checkpoint("03_seurat_filtered")) {
  seurat_filtered <- load_checkpoint("03_seurat_filtered")
} else {
  libs_dir <- file.path(output_path, "RData/filtered_libs/")
  dir.create(libs_dir, recursive = TRUE, showWarnings = FALSE)

  seurat_qc  <- load_checkpoint("02_seurat_qc")
  lib_names  <- names(seurat_qc)

  # --- Sequential loop: filter one library at a time --------------------------
  # After filtering each library it is saved individually and its slot in
  # seurat_qc is nulled so peak RAM stays bounded (~size of seurat_qc alone).
  for (lib_name in lib_names) {
    lib_rds <- paste0(libs_dir, "filtered_", lib_name, ".rds")

    if (file.exists(lib_rds)) {
      message(sprintf("[SKIP] Already filtered: %s", lib_name))
      seurat_qc[[lib_name]] <- NULL   # free slot even when skipping
      gc()
      next
    }

    obj <- seurat_qc[[lib_name]]
    n0  <- ncol(obj)
    
    # Layer 1: Demuxlet singlet filter
    

    # Layer 2: Paper inclusion filter
    # NOTE: cells with IncludedInStudy == FALSE may still pass QC;
    #        we include them here but flag them for optional downstream exclusion
    # (keep all singlets regardless of IncludedInStudy for our own QC analysis)

    # Layer 3: Standard human PBMC QC thresholds
    obj <- subset(obj,
                  subset = nFeature_RNA     >= QC_MIN_FEATURES   &
                           nFeature_RNA     <= QC_MAX_FEATURES   &
                           nCount_RNA       >= QC_MIN_COUNTS     &
                           nCount_RNA       <= QC_MAX_COUNTS     &
                           percent.MT       <= QC_MAX_MT         &
                           log10GenesPerUMI >= QC_MIN_COMPLEXITY)
    n1 <- ncol(obj)

    message(sprintf("[FILTER] %s: %d → filter %d (kept %.1f%%)",
                    lib_name, n0, n1,  n1 / n0 * 100))

    saveRDS(obj, file = lib_rds)
    message(sprintf("[SAVED]  %s", lib_rds))

    # Free this library's slot before loading the next one
    seurat_qc[[lib_name]] <- NULL
    rm(obj)
    gc()
  }
  rm(seurat_qc)
  gc()

  # --- Assemble final list from individual checkpoints ------------------------
  message("\n[ASSEMBLE] Loading individual filtered libraries...")
  seurat_filtered <- lapply(lib_names, function(lib_name) {
    readRDS(paste0(libs_dir, "filtered_", lib_name, ".rds"))
  })
  names(seurat_filtered) <- lib_names

  save_checkpoint(seurat_filtered, "04_seurat_filtered")

  # To remove individual per-library files after the combined checkpoint is
  # confirmed good, uncomment and run manually:
  # unlink(paste0(libs_dir, "filtered_", lib_names, ".rds"))
}
```

## Post-filter QC summary


```{r}
# Quick check of matching quality (cells found in metadata)
import_filtered <- do.call(rbind, lapply(names(seurat_filtered), function(nm) {
  obj <- seurat_filtered[[nm]]
  n_total     <- ncol(obj)
  data.frame(Library = nm, N_Cells = n_total,
             row.names = NULL)
}))

resume_filtered <- left_join(import_tbl, import_filtered, by = "Library") %>%
  dplyr::rename(N_Cells_PreFilter = N_Cells.x,
                N_Cells_PostFilter = N_Cells.y) %>%
  mutate(Retained.pct = N_Cells_PostFilter/N_Cells_PreFilter*100)

print(resume_filtered)
# Save table
save_table(resume_filtered, "QC_summary_cell_retention", subdir = "QC/")
rm(import_filtered, import_tbl, resume_filtered)
```

```{r qc_combined_meta}
combined_meta_filter <- build_combined_meta(seurat_filtered)
cat("Total barcodes with QC metrics:", nrow(combined_meta_filter), "\n")
cat("Libraries:", length(unique(combined_meta_filter$library)), "\n")

# Save metadata_prefilter for downstream use
save_checkpoint(combined_meta_filter, "qc_combined_meta_filter")
```

### Distribution of features
#### nCount_RNA distribution per library (filter)
```{r}
ncount_rna_filter <- ggplot(combined_meta_filter) +
  aes(x = "", y = nCount_RNA, fill = library) +
  geom_violin(adjust = 1L) +
  geom_jitter(alpha = 0.01, size=0.02) +
  scale_fill_hue(direction = 1) +
  ggthemes::theme_few() +
  facet_wrap(vars(library))+
  #remove legend
  theme(legend.position = "none") +
  labs(x = "", y = "nCount_RNA")

print(ncount_rna_filter)
plots_filter[["ncount_rna_filter"]] <- ncount_rna_filter

rm(ncount_rna_filter)

```
#### nFeature_RNA distribution per library (filter)
```{r}
nFeature_RNA_filter <- ggplot(combined_meta_filter) +
  aes(x = "", y = nFeature_RNA, fill = library) +
  geom_violin(adjust = 1L) +
  geom_jitter(alpha = 0.01, size=0.02) +
  scale_fill_hue(direction = 1) +
  ggthemes::theme_few() +
  facet_wrap(vars(library))+
  #remove legend
  theme(legend.position = "none") +
  labs(x = "", y = "nFeature_RNA")

print(nFeature_RNA_filter)
plots_filter[["nFeature_RNA_filter"]] <- nFeature_RNA_filter

rm(nFeature_RNA_filter)

```
#### percent.MT distribution per library (filter)
```{r}
percent.MT_filter <- ggplot(combined_meta_filter) +
  aes(x = "", y = percent.MT, fill = library) +
  geom_violin(adjust = 1L) +
  geom_jitter(alpha = 0.01, size=0.02) +
  scale_fill_hue(direction = 1) +
  ggthemes::theme_few() +
  facet_wrap(vars(library))+
  #remove legend
  theme(legend.position = "none") +
  labs(x = "", y = "percent.MT")

print(percent.MT_filter)
plots_filter[["percent.MT_filter"]] <- percent.MT_filter

rm(percent.MT_filter)

```
#### log10GenesPerUMI distribution per library (filter)
```{r}
log10GenesPerUMI_filter <- ggplot(combined_meta_filter) +
  aes(x = "", y = log10GenesPerUMI, fill = library) +
  geom_violin(adjust = 1L) +
  geom_jitter(alpha = 0.01, size=0.02) +
  scale_fill_hue(direction = 1) +
  ggthemes::theme_few() +
  facet_wrap(vars(library))+
  #remove legend
  theme(legend.position = "none") +
  labs(x = "", y = "log10GenesPerUMI")

print(log10GenesPerUMI_filter)
plots_filter[["log10GenesPerUMI_filter"]] <- log10GenesPerUMI_filter

rm(log10GenesPerUMI_filter)

```
#### percent.Ribo distribution per library (filter)
```{r}
percent.ribosomal_filter <- ggplot(combined_meta_filter) +
  aes(x = "", y = percent.ribosomal, fill = library) +
  geom_violin(adjust = 1L) +
  geom_jitter(alpha = 0.01, size=0.02) +
  scale_fill_hue(direction = 1) +
  ggthemes::theme_few() +
  facet_wrap(vars(library))+
  #remove legend
  theme(legend.position = "none") +
  labs(x = "", y = "percent.ribosomal")

print(percent.ribosomal_filter)
plots_filter[["percent.ribosomal_filter"]] <- percent.ribosomal_filter

rm(percent.ribosomal_filter)

```


### Cells per library

```{r plot_ncells_prelim}
p_cells_library <- combined_meta_filter %>%
  ggplot(aes(x = library, fill = condition)) +
  geom_bar() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("Number of Cells per Library") +
  xlab("Library") +
  ylab("Number of Cells") 
  # + facet_wrap(~region)


print(p_cells_library)
plots_filter[["p_cells_library"]] <- p_cells_library

rm(p_cells_library)
```
### UMI counts distribution

The UMI counts per cell (nCount_RNA) should generally be above 500, that
is the low end of what we expect. If UMI counts are between 500-1000
counts, it is usable but the cells probably should have been sequenced
more deeply.

```{r plot_umi_dist}
p_UMI_dist <- combined_meta_filter %>%
  ggplot(aes(x = nCount_RNA, color = library, fill = library)) +
  geom_density(alpha = 0.15) +
  scale_x_log10() +
  geom_vline(xintercept = QC_MIN_COUNTS, linetype = "dashed", color = "red") +
  geom_vline(xintercept = QC_MAX_COUNTS, linetype = "dashed", color = "red") +
  theme_classic(base_size = 10) +
  theme(legend.position = "none") +
  facet_wrap(~region, ncol = 2) +
  labs(title = "UMI distribution per library (filter)",
       x = "nCount_RNA (log10)", y = "Density")

print(p_UMI_dist)
plots_filter[["p_UMI_dist"]] <- p_UMI_dist

rm(p_UMI_dist)
```

### Genes detected per cell
We have similar expectations for gene detection as for UMI detection,
although it may be a bit lower than UMIs. For high quality data, the
proportional histogram should contain a single large peak that
represents cells that were encapsulated. If we see a small shoulder to
the right of the major peak (not present in our data), or a bimodal
distribution of the cells, that can indicate a couple of things. It
might be that there are a set of cells that failed for some reason. It
could also be that there are biologically different types of cells (i.e.
quiescent cell populations, less complex cells of interest), and/or one
type is much smaller than the other (i.e. cells with high counts may be
cells that are larger in size). Therefore, this threshold should be
assessed with other metrics that we describe in this lesson.

```{r plot_genes_dis}
p_genes_dist <-  combined_meta_filter %>%
  ggplot(aes(x = nFeature_RNA, color = library, fill = library)) +
  geom_density(alpha = 0.15) +
  scale_x_log10() +
  geom_vline(xintercept = QC_MIN_COUNTS, linetype = "dashed", color = "red") +
  geom_vline(xintercept = QC_MAX_COUNTS, linetype = "dashed", color = "red") +
  theme_classic(base_size = 10) +
  theme(legend.position = "none") +
  facet_wrap(~region, ncol = 2) +
  labs(title = "Distribution of number of genes (filter)",
       x = "nFeature_RNA", y = "Density")


print(p_genes_dist)
plots_filter[["p_genes_dist"]] <- p_genes_dist

rm(p_genes_dist)
```


### Mitochondrial percentage

This metric can identify whether there is a large amount of
mitochondrial contamination from dead or dying cells. We define poor
quality samples for mitochondrial counts as cells which surpass the 0.2
mitochondrial ratio mark, unless of course you are expecting this in
your sample.


```{r plot_mt_dist}
p_mt_dist <- combined_meta_filter %>%
  dplyr::filter(percent.MT<=5) %>%  # Filter out extreme outliers for better visualization)
  ggplot(aes(x = percent.MT, color = library, fill = library)) +
  geom_density(alpha = 0.15) +
  geom_vline(xintercept = QC_MAX_MT, linetype = "dashed", color = "red") +
  theme_classic(base_size = 10) +
  theme(legend.position = "none") +
  facet_wrap(~region, ncol = 2) +
  labs(title = paste0("Mitochondrial % (filter) | threshold = ", QC_MAX_MT, "%"),
       x = "% Mitochondrial reads", y = "Density")
print(p_mt_dist)
plots_filter[["p_mt_dist"]] <- p_mt_dist

rm(p_mt_dist)
```


### UMIs vs. genes detected (coloured by MT%)

Two metrics that are often evaluated together are the number of UMIs and
the number of genes detected per cell. Here, we have plotted the number
of genes versus the numnber of UMIs coloured by the fraction of
mitochondrial reads. Mitochondrial read fractions are only high (light
blue color) in particularly low count cells with few detected genes.
This could be indicative of damaged/dying cells whose cytoplasmic mRNA
has leaked out through a broken membrane, and thus, only mRNA located in
the mitochondria is still conserved. These cells are filtered out by our
count and gene number thresholds. Jointly visualizing the count and gene
thresholds shows the joint filtering effect.

Cells that are poor quality are likely to have low genes and UMIs per
cell, and correspond to the data points in the bottom left quadrant of
the plot. Good cells will generally exhibit both higher number of genes
per cell and higher numbers of UMIs.

With this plot we also evaluate the slope of the line, and any scatter
of data points in the bottom right hand quadrant of the plot. These
cells have a high number of UMIs but only a few number of genes. These
could be dying cells, but also could represent a population of a low
complexity celltype (i.e red blood cells).

```{r plot_umi_vs_genes}
p_umi_vs_genes <- combined_meta_filter %>%
  ggplot(aes(x = nCount_RNA, y = nFeature_RNA, color = percent.MT)) +
  geom_point(size = 0.3, alpha = 0.5) +
  scale_colour_gradient(low = "grey90", high = "red") +
  scale_x_log10() + scale_y_log10() +
  geom_vline(xintercept = QC_MIN_COUNTS, linetype = "dashed") +
  geom_hline(yintercept = QC_MIN_FEATURES, linetype = "dashed") +
  facet_wrap(~region, ncol = 2) +
  theme_classic(base_size = 9) +
  labs(title = "UMIs vs Genes per cell (filter, 50k sample)",
       x = "nCount_RNA (log10)", y = "nFeature_RNA (log10)")
print(p_umi_vs_genes)
plots_filter[["p_umi_vs_genes"]] <- p_umi_vs_genes

rm(p_umi_vs_genes)
```


### Complexity (log10 Genes per UMI)


We can see the samples where we sequenced each cell less have a higher
overall complexity, that is because we have not started saturating the
sequencing for any given gene for these samples. Outlier cells in these
samples might be cells that have a less complex RNA species than other
cells. Sometimes we can detect contamination with low complexity cell
types like red blood cells via this metric. Generally, we expect the
novelty score to be above 0.80.

```{r plot_complexity}
p_complexity <- combined_meta_filter %>%
  ggplot(aes(x = log10GenesPerUMI, color = library, fill = library)) +
  geom_density(alpha = 0.15) +
  geom_vline(xintercept = QC_MIN_COMPLEXITY, linetype = "dashed", color = "red") +
  theme_classic(base_size = 10) +
  theme(legend.position = "none") +
  facet_wrap(~region, ncol = 2) +
  labs(title = "Transcriptome complexity (filter)",
       x = "log10(Genes / UMI)", y = "Density")

print(p_complexity)

plots_filter[["p_complexity"]] <- p_complexity

rm(p_complexity)
```

```{r}
base::saveRDS(plots_filter, file.path(output_path, "RData", "QC_plots_filter.rds"))
plots_filter <- readRDS(file.path(output_path,"RData/QC_plots_filter.rds"))

rm(combined_meta_filter)
gc()
```

## Plots QC
### Distribution of features (pre vs post filter)
#### nCount_RNA distribution per library (pre vs post filter)
```{r}
nCount_RNA_plot <- plots_prefilter[["ncount_rna_prefilter"]] + plots_filter[["ncount_rna_filter"]] +
  plot_annotation(title = "nCount_RNA per library (pre vs post filter)",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")),
                  tag_levels = "A")
print(nCount_RNA_plot)
save_plot("nCount_RNA_per_library_prefilter_vs_filter", nCount_RNA_plot, subdir = "QC/")
rm(nCount_RNA_plot)
```


#### nFeature_RNA distribution per library (pre vs post filter)
```{r}
nFeature_RNA_plot <- plots_prefilter[["nFeature_RNA_prefilter"]] + plots_filter[["nFeature_RNA_filter"]] +
  plot_annotation(title = "nFeature_RNA per library (pre vs post filter)",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")),
                  tag_levels = "A")
print(nFeature_RNA_plot)
save_plot("nFeature_RNA_per_library_prefilter_vs_filter", nFeature_RNA_plot, subdir = "QC/")
rm(nFeature_RNA_plot)
```


#### percent.MT distribution per library (pre vs post filter)
```{r}
percent.MT_plot <- plots_prefilter[["percent.MT_prefilter"]] + plots_filter[["percent.MT_filter"]] +
  plot_annotation(title = "Mitochondrial % per library (pre vs post filter)",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")),
                  tag_levels = "A")
print(percent.MT_plot)
save_plot("percent.MT_plot_per_library_prefilter_vs_filter", percent.MT_plot, subdir = "QC/")
rm(percent.MT_plot)
```


#### log10GenesPerUMI distribution per library (pre vs post filter)
```{r}
log10GenesPerUMI_plot <- plots_prefilter[["log10GenesPerUMI_prefilter"]] + plots_filter[["log10GenesPerUMI_filter"]] +
  plot_annotation(title = "log10GenesPerUMI per library (pre vs post filter)",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")),
                  tag_levels = "A")
print(log10GenesPerUMI_plot)
save_plot("log10GenesPerUMI_per_library_prefilter_vs_filter", log10GenesPerUMI_plot, subdir = "QC/")
rm(log10GenesPerUMI_plot)
```


#### percent.Ribo distribution per library (pre vs post filter)

```{r}
percent.ribosomal_plot <- plots_prefilter[["percent.ribosomal_prefilter"]] + plots_filter[["percent.ribosomal_filter"]] +
  plot_annotation(title = "Ribosomal % per library (pre vs post filter)",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")),
                  tag_levels = "A")
print(percent.ribosomal_plot)
save_plot("percent.ribosomal_plot_library_prefilter_vs_filter", percent.ribosomal_plot, subdir = "QC/")
rm(percent.ribosomal_plot)
```



### Cells per library
```{r}
cells_library_plot <- plots_prefilter[["p_cells_library"]] + plots_filter[["p_cells_library"]] +
  plot_annotation(title = "Cells per library (pre vs post filter)",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")),
                  tag_levels = "A")
print(cells_library_plot)
save_plot("cells_per_library_prefilter_vs_filter", cells_library_plot, subdir = "QC/")
rm(cells_library_plot)
```

### UMI counts distribution
```{r}
umi_counts_plot <- plots_prefilter[["p_UMI_dist"]] + plots_filter[["p_UMI_dist"]]+
  plot_annotation(title = "UMI counts distribution (pre vs post filter)",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")),
                  tag_levels = "A")
print(umi_counts_plot)
save_plot("UMI_counts_distribution_per_library_prefilter_vs_filter", umi_counts_plot, subdir = "QC/")
rm(umi_counts_plot)
```

### Genes detected per cell
```{r}
genes_detected_plot <- plots_prefilter[["p_genes_dist"]] + plots_filter[["p_genes_dist"]] +
  plot_annotation(title = "Genes detected per cell (pre vs post filter)",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")),
                  tag_levels = "A")
print(genes_detected_plot)
save_plot("Genes_detected_per_cell_prefilter_vs_filter", genes_detected_plot, subdir = "QC/")
rm(genes_detected_plot)
```

### Mitochondrial percentage
```{r}
mitochondrial_plot <- plots_prefilter[["p_mt_dist"]] + plots_filter[["p_mt_dist"]] +
  plot_annotation(title = "Mitochondrial percentage (pre vs post filter)",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")),
                  tag_levels = "A")
print(mitochondrial_plot)
save_plot("mitochondrial_percentage_per_library_prefilter_vs_filter", mitochondrial_plot, subdir = "QC/")
rm(mitochondrial_plot)
```

### UMIs vs. genes detected (coloured by MT%)
```{r}
umi_vs_genes_plot <- plots_prefilter[["p_umi_vs_genes"]] + plots_filter[["p_umi_vs_genes"]] +
  plot_annotation(title = "UMI counts vs. genes detected (pre vs post filter)",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")),
                  tag_levels = "A")
print(umi_vs_genes_plot)
save_plot("UMI_counts_vs_genes_detected_per_library_prefilter_vs_filter", umi_vs_genes_plot, subdir = "QC/")
rm(umi_vs_genes_plot)
```

### Complexity (log10 Genes per UMI)
```{r}
complexity_plot <- plots_prefilter[["p_complexity"]] + plots_filter[["p_complexity"]] +
  plot_annotation(title = "Complexity (pre vs post filter)",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")),
                  tag_levels = "A")
print(complexity_plot)
save_plot("Complexity_per_library_prefilter_vs_filter", complexity_plot, subdir = "QC/")
rm(complexity_plot)
```

```{r}
rm(plots_filter, plots_prefilter)
gc()
```

# 4. Doublet Detection (DoubletFinder)

DoubletFinder is run **per library** (before integration) on QC-filtered cells.
Each result is checkpointed. The Demuxlet classification already filtered
inter-sample doublets; DoubletFinder catches intra-sample technical doublets.

> **Memory note:** Processing runs sequentially (1 library at a time) to stay
> within 16 GB RAM. Each library is unloaded after processing.

```{r doubletfinder, message=TRUE}
message("[DOUBLETS] Running scDblFinder per sample...")
# Run DoubletFinder on each library (checkpoint per library)
#df_results <- lapply(names(seurat_filtered), function(lib_name) {
#  run_doubletfinder_bp(seurat_filtered[[lib_name]], lib_name, output_path)
#})
#names(df_results) <- names(seurat_filtered)

# Run DoubletFinder on each library (checkpoint per library). Using BPCells matrixs for faster processing.
  df_results  <- lapply(names(seurat_filtered), function(nm)
                   run_doubletfinder_bp(seurat_filtered[[nm]], nm, output_path))
  names(df_results) <- names(seurat_filtered)

  seurat_singlets <- build_seurat_singlets(seurat_filtered, df_results) # Save singlet-only Seurat objects for downstream use
  # Save "06_seurat_singlets.rds" checkpoint for downstream use
  rm(seurat_filtered, df_results)
  gc()

```


# 5. MERGE + NORMALISE + VARIABLE FEATURES + SCALE + PCA

```{r}

singlets_path <- file.path(
  output_path, "RData",
  paste0(CHECKPOINT_PREFIX, "06_seurat_singlets.rds")
)
if (!file.exists(singlets_path)) {
  stop(
    "[ABORT] checkpoint_06_seurat_singlets.rds not found at: ", singlets_path,
    "\n  Run QC + doublet removal steps first."
  )
}

log_msg("[LOAD] Loading checkpoint_06_seurat_singlets from: ", singlets_path)
seurat_singlets <- readRDS(singlets_path)
sample_names <- names(seurat_singlets)
log_msg("[LOAD] ", length(sample_names), " samples: ", paste(sample_names, collapse = ", "))
log_msg("[RAM]  After load: ", ram_mb(), " MB")

```

```{r}
log_msg("[INFO]  output_path = ", output_path)
log_msg("[INFO]  N_INTEGRATION_FEATURES = ", N_INTEGRATION_FEATURES)
log_msg("[INFO]  N_PCS_INTEGRATION = ", N_PCS_INTEGRATION)
```

Merge, Normalization, Variable Feature Selection, Scaling, and PCA are all performed in a single step to minimize RAM usage. Seurat v5's handling of split layers allows us to avoid iterative merging and the associated ".1"/".2" suffix accumulation, while still applying normalization independently per sample.
```{r}
if (check_checkpoint("lognorm_07_merged")) {
  log_msg("[SKIP] lognorm_07_merged found — loading...")
  merged <- load_checkpoint("lognorm_07_merged")
} else {
  # --- Merge (single call) ----------------------------------------------------
  # Single merge() avoids ".1"/".2" suffix accumulation from iterative merging.
  # RNA assay stays Assay5 + BPCells on disk throughout.
  log_msg("[MERGE] Single merge call — ", length(sample_names), " objects...")
  merged <- merge(
    seurat_singlets[[1]],
    y = seurat_singlets[-1]
  )
  rm(seurat_singlets)
  gc()

  log_msg(sprintf(
    "[MERGE] Done: %d cells | %d genes | RAM: %d MB",
    ncol(merged), nrow(merged), ram_mb()
  ))

  
# --- Normalise (per-layer, streaming over BPCells) --------------------------
  # Seurat v5 detects split layers and applies log-normalisation independently
  # per sample — equivalent to normalising each sample separately before merging.
  log_msg("[NORM] NormalizeData (LogNormalize, scale.factor = 10000)...")
  merged <- NormalizeData(merged,
    normalization.method = "LogNormalize",
    scale.factor         = 10000,
    verbose              = FALSE
  )
  log_msg("[RAM]  After NormalizeData: ", ram_mb(), " MB")

  # --- Variable features (consensus across samples) ---------------------------
  # FindVariableFeatures on split-layer objects computes HVGs per sample and
  # returns a consensus ranked set. Avoids one sample dominating HVG selection.
  log_msg("[HVG] FindVariableFeatures (nfeatures = ", N_INTEGRATION_FEATURES, ")...")
  merged <- FindVariableFeatures(merged,
    selection.method = "vst",
    nfeatures        = N_INTEGRATION_FEATURES,
    verbose          = FALSE
  )
  log_msg("[HVG] ", length(VariableFeatures(merged)), " variable features selected.")
  log_msg("[RAM]  After FindVariableFeatures: ", ram_mb(), " MB")

  # --- Scale (variable genes only, loads into RAM) ----------------------------
  # Only ~N_INTEGRATION_FEATURES genes are scaled — manageable RAM cost.
  # Required before PCA because NormalizeData does not centre or scale variance.
  log_msg("[SCALE] ScaleData on variable features...")
  merged <- ScaleData(merged,
    features = VariableFeatures(merged),
    verbose  = FALSE
  )
  log_msg("[RAM]  After ScaleData: ", ram_mb(), " MB")

  # --- PCA --------------------------------------------------------------------
  log_msg("[PCA] RunPCA (npcs = ", N_PCS_MAX, ")...")
  merged <- RunPCA(merged,
    assay   = "RNA",
    npcs    = N_PCS_MAX,
    verbose = FALSE
  )
  log_msg("[RAM]  After RunPCA: ", ram_mb(), " MB")

  # Variance explained by PCs (for elbow plot)
  pca_var    <- Stdev(merged, reduction = "pca")^2
  total_var  <- sum(pca_var)
  var_table  <- data.frame(
    PC                = seq_along(pca_var),
    Variance          = round(pca_var, 4),
    ExplainedVariance = round(pca_var / total_var * 100, 2),
    CumulativeVar     = round(cumsum(pca_var / total_var * 100), 2)
)
# Print first 30 PCs
print(head(var_table, 50))
writexl::write_xlsx(var_table, paste0(output_path, "tables/Variance_by_PC.xlsx"))
  
  # Calculate Number of PC.
  pcs <- n_pcs(merged)
  #Add to merged object metadata
  merged@misc$pca_selection <- list(n_pcs = pcs, method = "elbow_cumulative")
  
  log_msg("[Number of PC]:", pcs)
  
  # Elbow plot to guide choice of N_PCS_INTEGRATION
  p_elbow <- ElbowPlot(merged, ndims = N_PCS_MAX) +
     geom_vline(xintercept = pcs, linetype = "dashed", color = "red")
  
  save_plot("lognorm_elbow", p_elbow, width = 8, height = 5, subdir = "integration/")
print(p_elbow)
  save_checkpoint(merged, "lognorm_07_merged")
  log_msg("[CHECKPOINT] lognorm_07_merged saved.")
}

```

# 6. Integration
Integration is performed separately for RPCA and Harmony methods to allow comparison. Both methods start from the same merged object (pre-joined layers) to ensure a fair comparison. The RPCA integration is performed first, followed by Harmony, with separate checkpoints for each.

## RPCA INTEGRATION

```{r}

if (check_checkpoint("lognorm_08_rpca")) {
  log_msg("[SKIP] lognorm_08_rpca found — loading...")
  obj_rpca <- load_checkpoint("lognorm_08_rpca")
} else {
  log_msg("[RPCA] IntegrateLayers — method = RPCAIntegration...")
  merged <- load_checkpoint("lognorm_07_merged") # Start from merged object (pre-joined layers)
  n_pcs <- merged@misc$n_pcs
  log_msg("[INFO]  group.by.vars = sample_id | dims = 1:", n_pcs)
  
}
  obj_rpca <- IntegrateLayers(
    object               = merged,
    method               = RPCAIntegration,
    orig.reduction       = "pca",
    normalization.method = "LogNormalize",
    dims                 = 1:n_pcs,
    verbose              = TRUE
  )
  gc()
  log_msg("[RAM]  After RPCA integration: ", ram_mb(), " MB")

  # --- Join layers (required before FindMarkers and most visualisation) -------
  log_msg("[LAYERS] JoinLayers on RNA assay...")
  obj_rpca[["RNA"]] <- JoinLayers(obj_rpca[["RNA"]])
  gc()
  log_msg("[RAM]  After JoinLayers: ", ram_mb(), " MB")

  # --- Clustering + UMAP (RPCA) -----------------------------------------------
  log_msg("[CLUSTER] FindNeighbors + FindClusters (RPCA)...")
  obj_rpca <- FindNeighbors(obj_rpca,
    reduction = "integrated.dr",
    dims      = 1:n_pcs,
    verbose   = FALSE
  )

  for (res in CLUSTERING_RESOLUTIONS) {
    obj_rpca <- FindClusters(obj_rpca,
      resolution = res,
      algorithm  = 1,
      verbose    = FALSE
    )
    log_msg(sprintf(
      "[CLUSTER] resolution %.2f → %d clusters",
      res, length(unique(obj_rpca@meta.data[[paste0("RNA_snn_res.", res)]]))
    ))
  }

  log_msg("[UMAP] RunUMAP on integrated.rpca (dims = 1:", pcs, ")...")
  obj_rpca <- RunUMAP(obj_rpca,
    reduction      = "integrated.dr",
    dims           = 1:n_pcs,
    umap.method    = "uwot",
    reduction.name = "umap.rpca",
    verbose        = FALSE
  )

  # Plots per resolution
  for (res in CLUSTERING_RESOLUTIONS) {
    col_nm <- paste0("RNA_snn_res.", res)
    Idents(obj_rpca) <- col_nm
    p <- DimPlot(obj_rpca,
      reduction = "umap.rpca",
      label     = TRUE,
      repel     = TRUE,
      raster = FALSE
    ) +
      ggtitle(paste0("RPCA | resolution ", res))
    save_plot(
      paste0("lognorm_rpca_umap_res", gsub("\\.", "", as.character(res))),
      p,
      width = 10, height = 8, subdir = "clustering"
    )
  }

  p_sample <- DimPlot(obj_rpca,
    reduction  = "umap.rpca",
    group.by   = "sample_id",
    label      = FALSE
  ) +
    ggtitle("RPCA | by sample")
  save_plot("lognorm_rpca_umap_sample", p_sample,
    width = 10, height = 8,
    subdir = "clustering/"
  )
  print(p_sample)
  
  for (dims_range in list(1:6, 7:12, 13:18)) {
  DimHeatmap(obj_rpca, dims = dims_range, cells = 500, balanced = TRUE)
  }
  
  
  save_checkpoint(obj_rpca, "lognorm_08_rpca")
  log_msg("[CHECKPOINT] lognorm_08_rpca saved.")
  log_msg("[RAM]  Final RPCA block: ", ram_mb(), " MB")



```

### Clustering Resolution Sweep

Tests combinations of PCs (10, 20, 30) × resolutions (0.1, 0.2, 0.4, 0.8)
to find the optimal clustering. Results visualized with **clustree**.

> **Memory note:** Each combination saves its `.rds` file individually.
> Already-computed combinations are skipped on re-run.

```{r clustering_sweep, message=TRUE}
if (check_checkpoint("RPCA_cluster_sweep_metadata")) {
  combined_metadata <- load_checkpoint("RPCA_cluster_sweep_metadata")
  message("[CLUSTERING] Loaded sweep results from checkpoint.")
} else {
  obj_rpca <- load_checkpoint("lognorm_08_rpca") # Ensure obj_rpca is in memory for clustering sweep
  combined_metadata <- data.frame()
  sweep_dir <- file.path(output_path, "RData")

  for (dims in CLUSTERING_DIMS) {
    for (res in CLUSTERING_RESOLUTIONS) {
      sweep_name <- paste0("Dims_", dims, "_Res_", res)
      rds_file   <- paste0(sweep_dir, "/sweep_rpca_", sweep_name, ".rds")

      if (file.exists(rds_file)) {
        message(sprintf("[SWEEP] Loading cached: dims=%d res=%.1f", dims, res))
        tmp <- readRDS(rds_file)
      } else {
        message(sprintf("[SWEEP] Computing: dims=%d res=%.1f  [%s]",
                        dims, res, format(Sys.time(), "%H:%M:%S")))
        tic()
        tmp <- FindNeighbors(obj_rpca, dims = 1:dims, verbose = FALSE)
        tmp <- FindClusters(tmp,  resolution = res,  verbose = FALSE)
        tmp <- RunUMAP(tmp, dims = 1:dims, verbose = FALSE)

        # Save UMAP plot
        p <- DimPlot(tmp, reduction = "umap", label = TRUE, raster = FALSE) +
          ggtitle(paste("dims =", dims, "| res =", res)) + NoLegend()
        print(p)
        save_plot(paste0("UMAP_sweep_", sweep_name), p, subdir = "clustering")

        saveRDS(tmp, file = rds_file)
        toc()
        gc()
      }

      # Collect cluster assignments for clustree
      cell_df <- data.frame(
        cell       = colnames(tmp),
        cluster    = as.character(Idents(tmp)),
        dims       = as.character(dims),
        resolution = as.character(res),
        stringsAsFactors = FALSE
      )
      combined_metadata <- dplyr::bind_rows(combined_metadata, cell_df)
      rm(tmp)
      gc()
    }
  }

  save_checkpoint(combined_metadata, "RPCA_cluster_sweep_metadata")
}
```

### Clustree

Visualizes cluster stability across resolutions for each dimension setting.

```{r clustree_plot, fig.height=10}
  for (d in CLUSTERING_DIMS) {
    subset_meta <- combined_metadata %>%
      dplyr::filter(dims == as.character(d)) %>%
      tidyr::pivot_wider(
        id_cols     = cell,
        names_from  = resolution,
        values_from = cluster,
        names_prefix = "Res_"
      ) 
      
    p_tree <- clustree::clustree(subset_meta, prefix = "Res_") +
      ggtitle(paste("Clustree — PCs =", d))
    print(p_tree)
    save_plot(paste0("Clustree_dims", d), p_tree,
              width = 10, height = 12, subdir = "clustering")
  } 
```

## HARMONY INTEGRATION

```{r}
if (check_checkpoint("lognorm_09_harmony")) {
  log_msg("[SKIP] lognorm_09_harmony found — nothing to do.")
} else {
  log_msg("[HARMONY] IntegrateLayers — method = HarmonyIntegration...")
  merged <- load_checkpoint("lognorm_07_merged") # Start from merged object (pre-joined layers)
  n_pcs <- merged@misc$n_pcs
  log_msg("[INFO]  group.by.vars = sample_id | dims = 1:", n_pcs)
  

  # Work from the pre-joined merged object so Harmony gets the split layers
  # (required for IntegrateLayers — JoinLayers must not be done beforehand).
  obj_harmony <- IntegrateLayers(
    object               = merged,
    method               = HarmonyIntegration,
    orig.reduction       = "pca",
    group.by.vars        = "sample_id",
    normalization.method = "LogNormalize",
    dims                 = 1:n_pcs,
    verbose              = TRUE
  )
  gc()
  log_msg("[RAM]  After Harmony integration: ", ram_mb(), " MB")

  # --- Join layers ------------------------------------------------------------
  log_msg("[LAYERS] JoinLayers on RNA assay...")
  obj_harmony[["RNA"]] <- JoinLayers(obj_harmony[["RNA"]])
  gc()

  # --- Clustering + UMAP (Harmony) --------------------------------------------
  log_msg("[CLUSTER] FindNeighbors + FindClusters (Harmony)...")
  obj_harmony <- FindNeighbors(obj_harmony,
    reduction = "harmony",
    dims      = 1:n_pcs,
    verbose   = FALSE
  )

  for (res in CLUSTERING_RESOLUTIONS) {
    obj_harmony <- FindClusters(obj_harmony,
      resolution = res,
      algorithm  = 1,
      verbose    = FALSE
    )
    log_msg(sprintf(
      "[CLUSTER] resolution %.2f → %d clusters",
      res, length(unique(obj_harmony@meta.data[[paste0("RNA_snn_res.", res)]]))
    ))
  }

  log_msg("[UMAP] RunUMAP on harmony (dims = 1:", n_pcs, ")...")
  obj_harmony <- RunUMAP(obj_harmony,
    reduction      = "harmony",
    dims           = 1:n_pcs,
    umap.method    = "uwot",
    reduction.name = "umap.harmony",
    verbose        = FALSE
  )

  # Plots per resolution
  for (res in CLUSTERING_RESOLUTIONS) {
    col_nm <- paste0("RNA_snn_res.", res)
    Idents(obj_harmony) <- col_nm
    p <- DimPlot(obj_harmony,
      reduction = "umap.harmony",
      label     = TRUE,
      repel     = TRUE,
      raster = FALSE
    ) +
      ggtitle(paste0("Harmony | resolution ", res))
    save_plot(
      paste0("lognorm_harmony_umap_res", gsub("\\.", "", as.character(res))),
      p,
      width = 10, height = 8, subdir = "clustering"
    )
  }

  p_sample <- DimPlot(obj_harmony,
    reduction  = "umap.harmony",
    group.by   = "sample_id",
    label      = FALSE,
    raster = FALSE
  ) +
    ggtitle("Harmony | by sample")
  print(p_sample)
  save_plot("lognorm_harmony_umap_sample", p_sample,
    width = 10, height = 8,
    subdir = "clustering"
  )

  for (dims_range in list(1:6, 7:12, 13:18)) {
  p <- DimHeatmap(obj_harmony, dims = dims_range, cells = 500, balanced = TRUE)
}
  
  save_checkpoint(obj_harmony, "lognorm_09_harmony")
  log_msg("[CHECKPOINT] lognorm_03_harmony saved.")
  log_msg("[RAM]  Final Harmony block: ", ram_mb(), " MB")
}

message("\n[DONE] Step 06 complete.")
message("  RPCA result  : checkpoint_lognorm_08_rpca.rds")
message("  Harmony result: checkpoint_lognorm_09_harmony.rds")
```


### Clustering Resolution Sweep

Tests combinations of PCs (10, 20, 30) × resolutions (0.1, 0.2, 0.4, 0.8)
to find the optimal clustering. Results visualized with **clustree**.

> **Memory note:** Each combination saves its `.rds` file individually.
> Already-computed combinations are skipped on re-run.

```{r clustering_sweep, message=TRUE}
if (check_checkpoint("Harmony_cluster_sweep_metadata")) {
  combined_metadata <- load_checkpoint("Harmony_cluster_sweep_metadata")
  message("[CLUSTERING] Loaded sweep results from checkpoint.")
} else {
  obj_harmony <- load_checkpoint("lognorm_09_harmony") # Ensure obj_rpca is in memory for clustering sweep
  combined_metadata <- data.frame()
  sweep_dir <- file.path(output_path, "RData")

  for (dims in CLUSTERING_DIMS) {
    for (res in CLUSTERING_RESOLUTIONS) {
      sweep_name <- paste0("Dims_", dims, "_Res_", res)
      rds_file   <- paste0(sweep_dir, "/sweep_harmony_", sweep_name, ".rds")

      if (file.exists(rds_file)) {
        message(sprintf("[SWEEP] Loading cached: dims=%d res=%.1f", dims, res))
        tmp <- readRDS(rds_file)
      } else {
        message(sprintf("[SWEEP] Computing: dims=%d res=%.1f  [%s]",
                        dims, res, format(Sys.time(), "%H:%M:%S")))
        tic()
        tmp <- FindNeighbors(obj_harmony, dims = 1:dims, verbose = FALSE)
        tmp <- FindClusters(tmp,  resolution = res,  verbose = FALSE)
        tmp <- RunUMAP(tmp, dims = 1:dims, verbose = FALSE)

        # Save UMAP plot
        p <- DimPlot(tmp, reduction = "umap", label = TRUE, raster = FALSE) +
          ggtitle(paste("dims =", dims, "| res =", res)) + NoLegend()
        print(p)
        save_plot(paste0("UMAP_sweep_", sweep_name), p, subdir = "clustering")

        saveRDS(tmp, file = rds_file)
        toc()
        gc()
      }

      # Collect cluster assignments for clustree
      cell_df <- data.frame(
        cell       = colnames(tmp),
        cluster    = as.character(Idents(tmp)),
        dims       = as.character(dims),
        resolution = as.character(res),
        stringsAsFactors = FALSE
      )
      combined_metadata <- dplyr::bind_rows(combined_metadata, cell_df)
      rm(tmp)
      gc()
    }
  }

  save_checkpoint(combined_metadata, "Harmony_cluster_sweep_metadata")
}
```

### Clustree

Visualizes cluster stability across resolutions for each dimension setting.

```{r clustree_plot, fig.height=10}
  for (d in CLUSTERING_DIMS) {
    subset_meta <- combined_metadata %>%
      dplyr::filter(dims == as.character(d)) %>%
      tidyr::pivot_wider(
        id_cols     = cell,
        names_from  = resolution,
        values_from = cluster,
        names_prefix = "Res_"
      ) 
      
    p_tree <- clustree::clustree(subset_meta, prefix = "Res_") +
      ggtitle(paste("Clustree — PCs =", d))
    print(p_tree)
    save_plot(paste0("Clustree_dims", d), p_tree,
              width = 10, height = 12, subdir = "clustering")
  } 
```


# 7. HVG before and after integration

```{r}
# Load Seurat objects
merged <- load_checkpoint("lognorm_07_merged")
obj_rpca <- load_checkpoint("lognorm_08_rpca")
obj_harmony <- load_checkpoint("lognorm_09_harmony")
#Before integration
merged <- FindVariableFeatures(merged, selection.method = "vst", nfeatures = 2000)
# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(merged), 10)
# plot variable features with and without labels
plot_hgv_merged <- LabelPoints(plot = VariableFeaturePlot(merged), points = top10, repel = TRUE) +
  ggtitle("Merged object")



#RPCA Integration
obj_rpca <- FindVariableFeatures(obj_rpca, selection.method = "vst", nfeatures = 2000)
# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(obj_rpca), 10)
# plot variable features with and without labels
plot_hgv_rpca <- LabelPoints(plot = VariableFeaturePlot(obj_rpca), points = top10, repel = TRUE)+
  ggtitle("RPCA integration")




#Harmony integration
obj_harmony <- FindVariableFeatures(obj_harmony, selection.method = "vst", nfeatures = 2000)
# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(obj_harmony), 10)
# plot variable features with and without labels
plot_hgv_harmony <- LabelPoints(plot = VariableFeaturePlot(obj_harmony), points = top10, repel = TRUE) +
  ggtitle("Harmony integration")


plot_hgv <- plot_hgv_merged / plot_hgv_rpca / plot_hgv_harmony 



print(plot_hgv)
save_plot("HVGs", plot_hgv, height = 20)

rm(plot_hgv, plot_hgv_merged, plot_hgv_rpca, plot_hgv_harmony)
```


# Select one specific integrated object resolution and number of dimensions

```{r}
#remove non-desired objects
rm(merged, obj_rpca, obj_harmony)
gc()

dims <- 30
res <- 0.4

data_select <- readRDS(paste0(output_path, "RData/sweep_harmony_Dims_", dims, "_Res_", res, ".rds"))
data_select@meta.data$condition <- factor(data_select@meta.data$condition, levels = c("Ctrl", "Inulin"))
rm("data_int_list")
```

# 6. Plot UMAP & UMAP Features

```{r}
tic()

## UMAP Visualize clusters
p1 <- DimPlot(data_select, reduction = "umap", group.by = "condition")
print(p1)
save_plot("clusters_integrated", p1)
  #UMAP
 # note that you can set `label = TRUE` or use the LabelClusters function to help label individual clusters
  plot <- DimPlot(data_select, reduction = "umap", group.by = "seurat_clusters", label = TRUE, repel = TRUE) + NoLegend()
  print(plot)
  save_plot("UMAP", plot)
#UMAP by condition
  plot <- DimPlot(data_select, reduction = "umap", split.by="condition", ncol = 2, label = TRUE)
  print(plot)
  save_plot("UMAP_ident", plot)
  
  #UMAP by features
  features <- c("nFeature_RNA", "nCount_RNA")
  feature_plot <- FeaturePlot(data_select, features = features)
  print(feature_plot)
  save_plot("Feature_plot", feature_plot)
  
toc()

```

## UMAP HVGs

```{r}
top <-head(VariableFeatures(data_select),10)

  #UMAP by features
  feature_plot <- FeaturePlot(data_select, features = top, ncol = 2)
    print(feature_plot)
  save_plot("Feature_plot_HVGs", feature_plot,width = 8, height = 18)
  
```

# 7. Finding differentially expressed features (cluster biomarkers)

Seurat can help you find markers that define clusters via differential
expression (DE). By default, it identifies positive and negative markers
of a single cluster (specified in ident.1), compared to all other cells.
FindAllMarkers() automates this process for all clusters, but you can
also test groups of clusters vs. each other, or against all cells.

In Seurat v5, we use the presto package (as described here and available
for installation here), to dramatically improve the speed of DE
analysis, particularly for large datasets. For users who are not using
presto, you can examine the documentation for this function
(?FindMarkers) to explore the min.pct and logfc.threshold parameters,
which can be increased in order to increase the speed of DE testing.

## Subset Clusters and Compare Genotypes

```{r}
options(future.globals.maxSize = 28000 * 1024^2) # 16 GB limit
plan("sequential") # Use 4 workers
gc()

# Normalizar RNA
DefaultAssay(data_select) <- "RNA"
data_select <- NormalizeData(data_select, normalization.method = "LogNormalize", scale.factor = 10000)

# Unir las capas del assay RNA antes de ejecutar FindMarkers
data_select <- JoinLayers(data_select, assay = "RNA")

```

```{r}
# Verificar que 'condition' está en metadata
if (!"condition" %in% colnames(data_select@meta.data)) {
  stop("Error: Column 'condition' does not exist in metadata. Verify that 'WT' & 'KO' are correctly assigned.")
}
# Establecer los clústeres como identidades antes de la iteración
Idents(data_select) <- "seurat_clusters"  # Set cluster identities

# List to store results
cluster_list <- list()

# Loop through each cluster
cluster_ids <- levels(data_select$seurat_clusters)

for (cluster in cluster_ids) {
  # Subset cells belonging to the current cluster
  cluster_cells <- subset(data_select, idents = cluster)
  
  Idents(cluster_cells) <- cluster_cells$condition  # Asigna 'WT' y 'KO' directamente como identidades                                    # Verificar que el clúster tiene ambas condiciones antes de correr FindMarkers
  if (length(unique(Idents(cluster_cells))) < 2) {
    warning(paste("Skipping cluster", cluster, "because does not have cells in both conditions (Ctlr & Inulin)."))
    next  # Salta al siguiente clúster
  }                                 
  # Compare WT vs KO within the cluster
  res <- FindMarkers(cluster_cells, 
                     ident.1 = "Ctrl", 
                     ident.2 = "Inulin", 
                     test.use = "wilcox", # Cambia si prefieres otro test
                     assay = "RNA",          # Use RNA assay
                     min.pct = 0.1,         # Filter genes expressed in at least 25% of cells
                     logfc.threshold = 0.1) # Minimum log fold-change threshold
  
  # Add cluster info to results
  res$gene <- rownames(res)
  res$cluster <- cluster
  cluster_list[[cluster]] <- res
}

# Save using the name
save(cluster_list, file = paste0(output_path, "/RData/", "cluster_list", ".RData"))  
cluster_list %>% 
write_xlsx(paste0(output_path,"/tables/","cluster_list.xlsx"))
# Load results ####
load(paste0(output_path,"/RData/cluster_list.RData"))

# Convertir la lista en un data.frame con una columna adicional 'cluster' (valor numérico)
cluster_markers_df <- do.call(rbind, lapply(names(cluster_list), function(cluster) {
  # Extraer el dataframe del cluster actual y añadir una columna con el número del cluster
  markers <- cluster_list[[cluster]]
  markers$cluster <- as.numeric(gsub("Cluster_", "", cluster)) # Eliminar el prefijo y convertir a numérico
  return(markers)
}))

# Reiniciar los nombres de las filas
rownames(cluster_markers_df) <- NULL




cluster_markers_df <- cluster_markers_df %>% 
  dplyr::select(gene, everything())

# Ver el resultado
head(cluster_markers_df)

#Generate a list 

cluster_markers_df %>% 
write_xlsx(paste0(output_path,"/tables/","cluster_markers_df.xlsx"))
# Save using the name
save(cluster_markers_df, file = paste0(output_path, "/RData/", "cluster_markers_df", ".RData"))  

### Load RDatas

# Load results ####
load(paste0(output_path,"/RData/cluster_markers_df.RData"))



```

## FindAllMarkers

```{r}
plan("sequential")  # Usa 6 núcleos (ajusta según tu PC)
options(future.globals.maxSize = 14000 * 1024^2) # 16 GB limit
gc()
# find markers for every cluster compared to all remaining cells, report only the positive
# ones
DefaultAssay(data_select) <- "RNA" #SCT
#data_select <- PrepSCTFindMarkers(data_select)
Idents(data_select) <- "seurat_clusters"  # Set cluster identities. Look for differences between cluster. Dont look for differences between genotype.
#Idents(data_select) <- "condition"  # Set condition identities. Look for differences between genotype and dont look for differenes between clusters


# Encuentra marcadores de todos los clusters
cluster.all.markers <- FindAllMarkers(data_select, 
                                      test.use = "wilcox",  # Test estadístico Wilcox,wilcox_limma, LR, MAST, DESeq2
                                      only.pos = FALSE, 
                                      min.pct = 0.1, 
                                      logfc.threshold = 0.1)


cluster.all.markers %>% 
write_xlsx(paste0(output_path,"/tables/","cluster.all.markers.xlsx"))

save(cluster.all.markers, file = paste0(output_path, "/RData/", "cluster.all.markers", ".RData")) # Save using the name
load(paste0(output_path, "/RData/", "cluster.all.markers", ".RData"))
plan("sequential")  # Usa 6 núcleos (ajusta según tu PC)
```

## HeatMap

```{r}
DefaultAssay(data_select) <- "RNA" #integrated
top10_all <- cluster.all.markers %>%
  group_by(cluster) %>%
  dplyr::filter(avg_log2FC > FC, p_val_adj < 0.05) %>%
  slice_head(n = 10) %>%
  ungroup()

top10 <- cluster_markers_df %>%
  group_by(cluster) %>%
  dplyr::filter(avg_log2FC > FC) %>%
  slice_head(n = 10) %>%
  ungroup()



plot <- DoHeatmap(data_select, features = top10_all$gene) + NoLegend()
print(plot)
save_plot("HeatMap_cluster_plot", plot, height = 18)


```

# 8. Manual Clustering Association
## Brain cell types (Rubio-Pastor F. PhD Thsis 2025)

We used Uniform Manifold Approximation and Projection (UMAP) to visualize all cell clusters and classified and annotated distinct cell types based on known marker genes for each major brain cell type. Following the initial cell clustering, we identified 7 major populations based on the expression of common cell-type markers (Haney et al., 2024): excitatory neurons (Slc17a7), inhibitory neurons (Gad1), oligodendrocytes (Mbp), oligodendrocyte precursor cells (OPCs, Cacng4), microglia (Siglech), astrocytes (Aqp4) and vascular cells (Cldn5) (Figure 67).
We then subdivided both excitatory and inhibitory neurons into distinct cell subtypes, annotating them further based on known markers and cortical cytoarchitecture, resulting in 14 clusters (Figure 68).
For excitatory neurons, clustering was performed using their gene expression patterns and known markers (Allen Institute, n.d.; Matho et al., 2021), as classification by cortical layer location proved challenging and produced a large number of distinct subtypes. We identified 5 different clusters (Figure 69): ExN1 (Lamp5+ Cux2+) enriched in L2/3, ExN2 (Ndst4+) enriched in L2/3, ExN3 (Rorb+ Cux1+) enriched in L4, ExN4 (Tox+ Hs3st2+) enriched in L4/5, and ExN5 (Tle4+ Fexf2+) enriched in L5/6. We classified inhibitory neurons into subtypes based on previous studies (Bugeon et al., 2022; Tasic et al., 2018; Yao et al., 2021), resulting in four groups: InN1 (Pvalb+), InN2 (Sst+), InN3 (Vip+), and InN4 (Lamp5+).

## Brain cell types ChatGPT
 - Excitatory Neurons: Slc17a7
    - ExN1 enriched L2/3: Lamp5, Cux2
    - ExN2 enriched L2/3: Ndst4
    - ExN3 enriched L4: Rorb, Cux2
    - ExN4 enriched L4/5: Tox, Hs3st2
    - ExN5 enriched L5/6: Tle4, Fexf2
 - Inhibitory Neurons: Gad1
    - InN1: Pvalb
    - InN2: Sst
    - InN3: Vip
    - InN4: Lamp5
 - Oligodendrocytes: Mbp
 - OPC (oligodendrocyte precursor cells): Cacng4
 - Microglia: Siglech
 - Astrocytes: Aqp4
 - Vascular cells: Cldn5
 

Neuronales:

- Neuronas de Purkinje: Calbindina (CALB1) 
- Neuronas granulares: GABRA6 (subunidad α6 del receptor GABA-A) 
- Neuronas inhibitorias(interneuronas): Parvalbúmina (PV), Somatostatina (SST), Calretinina(CALB2) 
- Neuronas excitatorias (proyección): Vesicular Glutamate Transporter 1 (VGLUT1), Vesicular Glutamate Transporter 2 (VGLUT2)
- Neuronas dopaminérgicas: Tirosina hidroxilasa (TH) 
- Neuronas serotoninérgicas: Tryptophan Hydroxylase 2 (TPH2) 
- Neuronas colinérgicas: Colina acetiltransferasa (ChAT) 

No neuronales:

- Astrocitos: Proteína ácida fibrilar glial (GFAP), S100β
- Oligodendrocitos: Proteína proteolipídica (PLP1), Mielina básica proteína (MBP) 
- Células microgliales: Ionized Calcium Binding Adapter Molecule 1 (IBA1, Aif1), CD11b (Cdk11b),Cx3cr1 
- Células endoteliales: Cluster of Differentiation 31 (CD31), Claudina-5 (CLDN5)

## Cerebellum cell types ()
 - Purkinje cells: Ppp1r17
 - Granule: Gabra6
 - UBC: Eomes
 - MLI1: Lypd6
 - MLI2: Prkcd
 - PLI: Klhl1
 - Golgi: Lgi2
 - Bergmannn: Gdf10
 - Astrocyte: Aqp4
 - ODC: Mobp
 - OPC: Ppfibp1
 - Fibroblast: Dcn
 - Endothelial Mural: Kcnj8
 - Choroid: Ttr
 - Macrophage: Mrc1
 - Microglia: C1qa
 - Endothelial: Flt1
 - Ependymal: Foxj1

### Violin Plot & Features plot

```{r}

top1 <- cluster.all.markers %>%
  group_by(cluster) %>%
  dplyr::filter(avg_log2FC > FC) %>%
  slice_head(n = 1) %>%
  ungroup()
# Cerebellum cell types
features = c("Ppp1r17", "Gabra6", "Eomes", "Lypd6", "Prkcd", "Klhl1", "Lgi2", "Gdf10", "Aqp4", "Mobp", "Ppfibp1", "Dcn", "Kcnj8", "Ttr", "Mrc1", "C1qa", "Flt1", "Foxj1")
# Brain cell types ChatGPT
features = c("Calb1", "Gabra6", "Pvalb", "Sst", "Calb2", "Tph2",  "Gfap", "Plp1","Mbp","Aif1", "Cdk11b", "Cx3cr1", "Pecam1", "Cldn5")

# Brain Cell types FRP
features = c("Slc17a7", "Lamp5", "Cux2", "Ndst4", "Rorb", "Tox", "Hs3st2", "Tle4", "Fexf2", "Gad1", "Pvalb", "Sst", "Vip", "Cacng4", "Siglech", "Cx3cr1", "Aqp4", "Cldn5")

features = c("Ppp1r17", "Gabra6", "Eomes", "Lypd6", "Prkcd", "Klhl1", "Lgi2", "Gdf10", "Aqp4", "Mobp", "Ppfibp1", "Dcn", "Kcnj8", "Ttr", "Mrc1", "C1qa", "Flt1", "Foxj1", "Calb1", "Pvalb", "Sst", "Calb2", "Tph2",  "Gfap", "Plp1","Mbp","Aif1", "Cdk11b", "Cx3cr1", "Pecam1", "Cldn5", "Slc17a7", "Lamp5", "Cux2", "Ndst4", "Rorb", "Tox", "Hs3st2", "Tle4", "Fexf2", "Gad1",  "Vip", "Cacng4", "Siglech")

features_presents <- features[features %in% rownames(data_select)]

for(feature in features_presents){
  #VlnPlot
  plot<- VlnPlot(data_select, features = feature, split.by = "condition")
  print(plot)
  save_plot(paste0("Vln_plot_",feature), plot)
  #UMAP by features
  feature_plot <- FeaturePlot(data_select, features = feature)
  print(feature_plot)
  save_plot("Feature_plot_Markers", feature_plot,width = 8, height = 6)
}
```
## DotPlot

```{r}

dotplot <- DotPlot(data_select, features = features_presents, cols = c("blue", "red"), dot.scale = 8, split.by = "condition") +
    RotatedAxis()
print(dotplot)
save_plot("DotPlot_features", dotplot, width = 16 ,height = 18)
```

# 8. Saving

```{r}
#Save Normalized and dimentionally reduced data
tic()
# Export results ####
# Save using the name
  save(data_select, file = paste0(output_path, "RData/", "data_select", ".RData"))  

### Load RDatas

# Load results ####
load(paste0(output_path,"RData/data_select.RData"))
toc()

gc()
```

# 9. Automatic Clustering association

1.  Use 'SingleR' is a widely used tool that compares gene expression
    profiles of your clusters to predefined reference databases. Load an
    appropriate reference database. For example:

-   MouseRNAseqData: Reference for murine tissues.
-   ImmGenData: Immunological reference.
-   HumanPrimaryCellAtlasData: For human cells.

2.  Use scType for fast annotation 'scType' is a simpler, more
    customizable tool that uses user-defined or database-available
    marker lists.

    Comparison between 'SingleR' and 'scType'

    | Feature           | SingleR                        | scType            |
    |-------------------|--------------------------------|-------------------|
    | **Ease of use**   | More robust but more technical | Simpler           |
    | **Database**      | Use predefined references      | Use marker lists  |
    | **Customization** | Limited                        | Very customizable |
    | **Speed**         | Moderate                       | Very fast         |

    : Recommendation If you have a well-characterized dataset (such as
    mouse brain transcriptomics), use SingleR with a database such as
    MouseRNAseqData. If you prefer greater flexibility or speed, try
    scType with specific marker lists.

## Allen Brain Atlas clustering

Download matrix Gene expression matrix and Table cell metadata from
Allen Brain Atlas Cell Types Database: RNA-Seq Data
(<https://portal.brain-map.org/atlases-and-data/rnaseq>) Whole Cortex &
Hippocampus - 10x genomics (2020) with 10x-SMART-seq taxonomy (2021)

## Generate Matrix from FindAllMarkers

```{r}
# Extraer la matriz de expresión y las identidades de los clusters
#cluster.markers <- GetAssayData(data_select, slot = "data")
#cluster.ids <- Idents(data_select)
```

```{r}

# Asegúrate de que el assay está configurado correctamente
DefaultAssay(data_select) <- "integrated"

# Extraer los genes significativos encontrados por FindAllMarkers
significant_genes <- cluster.all.markers %>% 
  dplyr::filter(p_val_adj < 0.05) %>% 
  dplyr::select(gene) %>% 
  dplyr::distinct()%>% 
  dplyr::pull(gene)
  
# Extraer la matriz de expresión para esos genes (utilizando 'layer' en lugar de 'slot')
expression_matrix <- GetAssayData(data_select, layer = "data")[significant_genes, ]

# Extraer las identidades de los clústeres
cluster.ids <- Idents(data_select)

# Crear un objeto SummarizedExperiment con la matriz de expresión
cluster.markers <- SummarizedExperiment(
  assays = list(logcounts = as.matrix(expression_matrix)),
  rowData = DataFrame(gene = rownames(expression_matrix)),
  colData = DataFrame(cluster = cluster.ids)
)



```

### Medians

```{r}
# Cargar los archivos CSV
median_data <- fread("rawdata/medians.csv")
trimmed_means_data <- fread("rawdata/trimmed_means.csv")

# Verificar las primeras filas
head(median_data)
head(trimmed_means_data)

# Revisar dimensiones
dim(median_data)
dim(trimmed_means_data)

# Convertir a matriz
median_matrix <- as.matrix(median_data[ , -1, with = FALSE])  # Asumiendo que la primera columna son los nombres de genes
rownames(median_matrix) <- median_data[[1]]  # Asignar nombres de genes a las filas

trimmed_means_matrix <- as.matrix(trimmed_means_data[ , -1, with = FALSE])  # Similar al anterior
rownames(trimmed_means_matrix) <- trimmed_means_data[[1]]

# Crear referencia de SingleR
# Crear un nuevo objeto SummarizedExperiment con logcounts
reference_median <- SummarizedExperiment(
    assays = list(logcounts = log1p(median_matrix)), # Aplicar log-transformation
    rowData = DataFrame(gene = rownames(median_matrix)),
    colData = DataFrame(label.main = colnames(median_matrix))
)
# Revisar las etiquetas
colData(reference_median)
# Confirmar que `label.main` contiene los nombres de los tipos celulares
reference_median$label.main <- colnames(median_matrix)

```

```{r}
# Comparar con la base de datos de referencia
annotations_median <- SingleR(test = cluster.markers, 
                       ref = reference_median, 
                       labels = reference_median$label.main,
                       clusters = cluster.ids)


# Revisar resultados
head(annotations_median)


```

#### Plot Median

```{r}
new.cluster.median <- annotations_median$labels
names(new.cluster.median) <- levels(data_select)
data_select_median <- RenameIdents(data_select, new.cluster.median)

# Convertir las anotaciones a un data.frame para unirlas con cluster_markers_df
annotations_median_df <- as.data.frame(annotations_median)
annotations_median_df$cluster <- rownames(annotations_median_df)  # Añadir los IDs de los clusters como columna

# Añadir columna 'labels'
cell_types_median <- annotations_median_df$labels
cluster_markers_df$labels_median <- cell_types_median[cluster_markers_df$cluster + 1]

```

```{r}
# Visualiza con las nuevas identidades
plot <- DimPlot(data_select_median, reduction = "umap", label = TRUE, repel = TRUE)
print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_UMAP_median_cluster_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  image_number <<- image_number + 1  # Use <<- to update the external variable

```

### Trimmed_Means

```{r}

reference_trimmed <- SummarizedExperiment(
    assays = list(logcounts = log1p(trimmed_means_matrix)), # Aplicar log-transformation
    rowData = DataFrame(gene = rownames(trimmed_means_matrix)),
    colData = DataFrame(label.main = colnames(trimmed_means_matrix))
)
# Revisar las etiquetas
colData(reference_trimmed)

# Confirmar que `label.main` contiene los nombres de los tipos celulares
reference_trimmed$label.main <- colnames(trimmed_means_matrix)
```

```{r}
# Comparar con la base de datos de referencia
annotations_trimmed <- SingleR(test = cluster.markers, 
                       ref = reference_trimmed, 
                       labels = reference_trimmed$label.main,
                       clusters = cluster.ids)


# Revisar resultados
head(annotations_trimmed)
```

#### Plot Trimmed

```{r}
new.cluster.trimmed <- annotations_trimmed$labels
names(new.cluster.trimmed) <- levels(data_select)
data_select_trimmed <- RenameIdents(data_select, new.cluster.trimmed)
# Convertir las anotaciones a un data.frame para unirlas con cluster_markers_df
annotations_trimmed_df <- as.data.frame(annotations_trimmed)
annotations_trimmed_df$cluster <- rownames(annotations_trimmed_df)  # Añadir los IDs de los clusters como columna

# Añadir columna 'labels'
cell_types_trimmed <- annotations_trimmed_df$labels
cluster_markers_df$labels_trimmed <- cell_types_trimmed[cluster_markers_df$cluster + 1]
```

```{r}
# Visualiza con las nuevas identidades
plot <- DimPlot(data_select_trimmed, reduction = "umap", label = TRUE, repel = TRUE)
print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_UMAP_trimmed_cluster_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  image_number <<- image_number + 1  # Use <<- to update the external variable

```

## MouseRNAseqData

```{r}


ref <- MouseRNAseqData()  # Change according needs
# Extraer la matriz de expresión y las identidades de los clusters
cluster.markers <- GetAssayData(data_select, slot = "data")
cluster.ids <- Idents(data_select)

# Comparar con la base de datos de referencia
annotations_ms <- SingleR(test = cluster.markers, 
                       ref = ref, 
                       labels = ref$label.main,
                       clusters = cluster.ids)

# Ver los resultados
head(annotations_ms)
```

```{r}
new.cluster.ids <- annotations_ms$labels
names(new.cluster.ids) <- levels(data_select)
data_select_ms <- RenameIdents(data_select, new.cluster.ids)
# Convertir las anotaciones a un data.frame para unirlas con cluster_markers_df
annotations_ms_df <- as.data.frame(annotations_ms)
annotations_ms_df$cluster <- rownames(annotations_ms_df)  # Añadir los IDs de los clusters como columna
# Añadir columna 'labels'
cell_types_ms <- annotations_ms_df$labels
cluster_markers_df$labels_ms <- cell_types_ms[cluster_markers_df$cluster + 1]
```

```{r}
# Visualiza con las nuevas identidades
plot <- DimPlot(data_select_ms, reduction = "umap", label = TRUE, repel = TRUE)
print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "UMAP_simple_cluster_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  image_number <<- image_number + 1  # Use <<- to update the external variable

```

## Cerebellum types

Taken from supplementary material from: Kozareva, V., Martin, C.,
Osorno, T. et al. A transcriptomic atlas of mouse cerebellar cortex
comprehensively defines cell types. Nature 598, 214–219 (2021).
<https://doi.org/10.1038/s41586-021-03220-z>

Supplementary Table 2: Highly differentially expressed genes across all
clusters. Genes are limited to top 500 differentially expressed genes
per subtype. Column names represent the following: “avgExpr” is the mean
gene expression in that subtype; “logFC” is the log fold change between
cells within the subtype compared to remaining cells; “pval” is the
p-value, calculated with the two-tailed Gaussian approximation of U
statistic; “padj” is the Benjamini-Hochberg adjusted p-value; “pct_in”
is the percent of cells in the subtype expressing the gene; “pct_out” is
the percent of cells outside the subtype expressing the gene.

```{r}
# Leer el archivo Excel (asumiendo que la primera hoja contiene los datos)
file_path <- "rawdata/cb_cell_types.xlsx"
data <- read_excel(file_path)

# Transformar los datos: agrupar subtipos similares y pivotar
cerebellum_data <- data %>%
  dplyr::select(gene, subtype, avgExpr) %>%  # Seleccionar solo las columnas necesarias
  dplyr::mutate(subtype = str_replace(subtype, "_\\d+$", "")) %>%  # Eliminar sufijos _1, _2, etc.
  dplyr::group_by(gene, subtype) %>%  # Agrupar por gen y subtipo
  dplyr::summarise(avgExpr = sum(avgExpr, na.rm = TRUE), .groups = "drop") %>%  # Sumar valores de expresión
  pivot_wider(
    names_from = subtype,  # Usar la columna "subtype" como nombres de columnas
    values_from = avgExpr,  # Usar los valores de la columna "avgExpr" como contenido de las celdas
    values_fill = list(avgExpr = 0)  # Asignar 0 a los valores NA
  )%>%
  dplyr::mutate(Purkinje = coalesce(Purkinje_Aldoc, 0) + coalesce(Purkinje_Anti_Aldoc, 0)) %>%  # Combinar columnas Purkinje_Aldoc y Purkinje_Anti_Aldoc
  dplyr::select(-Purkinje_Aldoc, -Purkinje_Anti_Aldoc)  # Eliminar las columnas originales
# Verificar las primeras filas
head(cerebellum_data)
# Revisar dimensiones
dim(cerebellum_data)

# Convertir a matriz
cerebellum_matrix <- as.matrix(cerebellum_data[ , -1, with = FALSE])  # Asumiendo que la primera columna son los nombres de genes
rownames(cerebellum_matrix) <- cerebellum_data[[1]]  # Asignar nombres de genes a las filas

# Crear referencia de SingleR
# Crear un nuevo objeto SummarizedExperiment con logcounts
reference_cerebellum <- SummarizedExperiment(
    assays = list(logcounts = log1p(cerebellum_matrix)), # Aplicar log-transformation
    rowData = DataFrame(gene = rownames(cerebellum_matrix)),
    colData = DataFrame(label.main = colnames(cerebellum_matrix))
)
# Revisar las etiquetas
colData(reference_cerebellum)

# Confirmar que `label.main` contiene los nombres de los tipos celulares
reference_cerebellum$label.main <- colnames(cerebellum_matrix)

```

```{r}
# Comparar con la base de datos de referencia
annotations_cerebellum <- SingleR(test = cluster.markers, 
                       ref = reference_cerebellum, 
                       labels = reference_cerebellum$label.main,
                       clusters = cluster.ids)


# Revisar resultados
head(annotations_cerebellum)


plot <- plotScoreHeatmap(annotations_cerebellum)
print(plot)
save_plot("Cerebellum_annotation", plot)
```

```{r}
new.cluster.cerebellum <- annotations_cerebellum$labels
names(new.cluster.cerebellum) <- levels(data_select)
data_select_cerebellum <- RenameIdents(data_select, new.cluster.cerebellum)

# Convertir las anotaciones a un data.frame para unirlas con cluster_markers_df
annotations_cerebellum_df <- as.data.frame(annotations_cerebellum)
annotations_cerebellum_df$cluster <- rownames(annotations_cerebellum_df)  # Añadir los IDs de los clusters como columna

cell_types_cerebellum <- annotations_cerebellum_df$labels
cluster_markers_df$labels_cerebellum <- cell_types_cerebellum[cluster_markers_df$cluster + 1]

```

```{r}
# Visualiza con las nuevas identidades
plot <- DimPlot(data_select_cerebellum, reduction = "umap", label = TRUE, repel = TRUE)
print(plot)
  # Set filename with incremented figure number
  filename <- paste0(output_path, "figures/", sprintf("%03d", image_number), "_UMAP_cerebellum_cluster_plot")
  # Save in TIFF and PDF formats
  ggsave(paste0(filename, tiff_extension), plot, width = 12, height = 6, units = "in")
  ggsave(paste0(filename, pdf_extension), plot, width = 12, height = 6, units = "in")
  # Increment the figure counter
  image_number <<- image_number + 1  # Use <<- to update the external variable

```
## Save Clusters

```{r}
#Generate a list 
cluster_list <- cluster_markers_df %>%
  group_split(cluster) %>%
  set_names(paste0("cluster_", unique(cluster_markers_df$cluster)))
#Save xlsx
cluster_list %>% 
write_xlsx(paste0(output_path,"tables/","cluster_list.xlsx"))
#Save RData

# Save using the name
save(cluster_list, file = paste0(output_path, "RData/", "cluster_list", ".RData"))
```

```{r}
#Clean enviroment and memory
# Especifica los nombres de los objetos que deseas conservar
keep_objects <- c("cluster_list", "image_number", "output_path", "data_select", "cluster.all.markers")
# Elimina todos los objetos excepto los especificados
rm(list = setdiff(ls(), keep_objects))
gc() #Clean memory
# Set up parameters
set.seed(123)
options(scipen = -1)  # Habilitar notación científica en R  
source("code/global_variables.R")  
### Load RDatas
# Load results ####
load(paste0(output_path,"RData/cluster_list.RData"))

```

Now, cluster_list will be used as input for the next steps.

# 9. Vulcano Plots

```{r}
filtration <- 0
source("code/02_vulcano_plots.R")

```

# 10. GO

```{r}

source("code/001_GO.R")
```

# 11. Strings

```{r}
source("code/003_strings.R")
```

# 12. GSE

```{r}
source("code/004_gse.R")
```

#13. KEGG

```{r}
source("code/09_gseKEGG.R")
```

# 14. HeatMaps

```{r}

# Gene list
gene_list <- c("Hmgcr", "Srebf2", "Apoe","Lrp1", "Lrp1b", "Abca1")

source("code/06_Heatmap.R")
```

# 15. HeatMaps GO Types

```{r}
#source("code/07_HeatMap_GO_types.R")
```

# 16. EnrichR

```{r}
source("code/08_EnrichR.R")
```