---
title: "SingleCell RNA seq general pipeline"
author: "Santiago López Begines"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: tango
    fig_width: 10
    fig_height: 8
    fig_caption: true
    code_folding: hide
    number_sections: true
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
editor_options:
  markdown:
    wrap: sentence
---

# 0. Setup

## Environment and global parameters

```{r env_setup}
rm(list = ls())   # Clean environment
gc()              # Release memory

# --- Global parameters --------------------------------------------------------
image_number <- 1
output_path  <- "output/GSE262881/SCT/"

# Data paths
DATA_DIR  <- "../rawdata/GSE262881_RAW"

# Random seed for reproducibility
set.seed(123)
options(scipen = 3, digits = 3)

# Memory: sequential plan (i7-7560U: 2 physical cores, 16 GB RAM)
# Parallel overhead outweighs benefits on 2-core hardware; protects RAM
library(future)
plan("sequential")
```

## Packages

```{r packages}
#| message: false
#| warning: false  
# renv::restore()  # Uncomment to restore packages from renv.lock
source("../code/00_packages.R")
# renv::snapshot()  # Snapshot current package state to renv.lock (uncomment if changes made
```


```{r load_globals}
source("../code/global_variables.R")
source("../code/01_aux_functions.R")
source("../code/02_sc_functions.R")
```


# Input data

**Dataset:** GSE262881 — Wang et al., BMC Biology (2025) 23:124  
**Paper:** A single-cell transcriptomic atlas of all cell types in the brain of 5xFAD Alzheimer mice in response to dietary inulin supplementation

---

## Full Processing Workflow (per Methods, p. 19)

```
FASTQ  →  CellRanger 7.1.0  →  raw_feature_bc_matrix/
                                        ↓
                              CellBender 0.2.2
                                        ↓
                    *_CellBender_feature_bc_matrix_filtered.h5   ← .h5 files on GEO

FASTQ  →  CellRanger 7.1.0 (BAM)  →  velocyto 0.17.17
                                        ↓
                                    *.loom.gz   ← .loom files on GEO
```

We start by reading in the data. The Read10X() function reads in the
output of the cellranger pipeline from 10X, returning a unique molecular
identified (UMI) count matrix. The values in this matrix represent the
number of molecules for each feature (i.e. gene; row) that are detected
in each cell (column). Note that more recent versions of cellranger now
also output using the h5 file format, which can be read in using the
Read10X_h5() function in Seurat.

Data from Wang et al., BMC Biology (2025) 23:124 comes from CellBender filtering. The paper check 2 conditions, Control vs. Insulin; in 4 different parts of brain: Us


## Read 10X matrices (with checkpoint)

Each library is read and saved individuallyto avoid holding all count matrices in RAM simultaneously.

```{r import_raw, message=TRUE}
# define output_path y CHECKPOINT_PREFIX in global_variables                                                                                                    

# Checkpoint: all raw Seurat objects
# Con checkpoint (recomendado — 24 ficheros h5 tardan)    
  seurat_list <- load_h5_samples(data_path, checkpoint = "01_raw_data")                                          # Only desired region, without checkpoint
# seurat_list <- load_h5_samples(data_path, regions = "forebrain")
gc()          
     
```

## Import summary

```{r import_summary}
# Quick check of matching quality (cells found in metadata)
import_tbl <- do.call(rbind, lapply(names(seurat_list), function(nm) {
  obj <- seurat_list[[nm]]
  n_total     <- ncol(obj)
  #n_matched   <- sum(!is.na(obj$Status))
  #n_healthy   <- sum(obj$Status == "Healthy", na.rm = TRUE)
  #n_psa       <- sum(obj$Status == "PSA",     na.rm = TRUE)
  #n_pso       <- sum(obj$Status == "PSO",     na.rm = TRUE)
  data.frame(Library = nm, N_Cells = n_total,
             row.names = NULL)
}))

print(import_tbl)
cat("\nTotal barcodes imported:", sum(import_tbl$N_Cells), "\n")
```
# 2. QC Metrics Calculation

Add human-specific QC metrics.
Key difference from mouse data: mitochondrial genes use **`^MT-`** (uppercase).

```{r add_qc_metrics}
if (check_checkpoint("02_seurat_qc")) {
  seurat_qc <- load_checkpoint("02_seurat_qc")
} else {
  #seurat_raw <- load_checkpoint("01_raw_seurat")
  seurat_qc <- lapply(seurat_list, function (x){
    add_qc_metrics(x, pattern_mito = mito_pattern , pattern_ribo = ribo_pattern)})
  rm(seurat_list) 
  gc()
  save_checkpoint(seurat_qc, "02_seurat_qc")
}
```
## Pre-filter QC overview (combined across all libraries)
```{r qc_combined_meta}
combined_meta_prefilter <- build_combined_meta(seurat_qc)
cat("Total barcodes with QC metrics:", nrow(combined_meta_prefilter), "\n")
cat("Libraries:", length(unique(combined_meta_prefilter$library)), "\n")

# Save metadata_prefilter for downstream use
save_checkpoint(combined_meta_prefilter, "02_combined_meta_prefilter")

plots_prefilter <- list()  # Initialize list to store pre-filter QC plots
plots_filter <- list()     # Initialize list to store post-filter QC plots
```

### Cells per library

```{r plot_ncells_prelim}
p_cells_library <- combined_meta_prefilter %>%
  ggplot(aes(x = library, fill = condition)) +
  geom_bar() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("Number of Cells per Library") +
  xlab("Library") +
  ylab("Number of Cells") 
  # + facet_wrap(~region)


print(p_cells_library)
plots_prefilter[["p_cells_library"]] <- p_cells_library

rm(p_cells_library)
```
### UMI counts distribution

The UMI counts per cell (nCount_RNA) should generally be above 500, that
is the low end of what we expect. If UMI counts are between 500-1000
counts, it is usable but the cells probably should have been sequenced
more deeply.

```{r plot_umi_dist}
p_UMI_dist <- combined_meta_prefilter %>%
  ggplot(aes(x = nCount_RNA, color = library, fill = library)) +
  geom_density(alpha = 0.15) +
  scale_x_log10() +
  geom_vline(xintercept = QC_MIN_COUNTS, linetype = "dashed", color = "red") +
  geom_vline(xintercept = QC_MAX_COUNTS, linetype = "dashed", color = "red") +
  theme_classic(base_size = 10) +
  theme(legend.position = "none") +
  facet_wrap(~region, ncol = 2) +
  labs(title = "UMI distribution per library (pre-filter)",
       x = "nCount_RNA (log10)", y = "Density")

print(p_UMI_dist)
plots_prefilter[["p_UMI_dist"]] <- p_UMI_dist

rm(p_UMI_dist)
```

### Genes detected per cell
We have similar expectations for gene detection as for UMI detection,
although it may be a bit lower than UMIs. For high quality data, the
proportional histogram should contain a single large peak that
represents cells that were encapsulated. If we see a small shoulder to
the right of the major peak (not present in our data), or a bimodal
distribution of the cells, that can indicate a couple of things. It
might be that there are a set of cells that failed for some reason. It
could also be that there are biologically different types of cells (i.e.
quiescent cell populations, less complex cells of interest), and/or one
type is much smaller than the other (i.e. cells with high counts may be
cells that are larger in size). Therefore, this threshold should be
assessed with other metrics that we describe in this lesson.

```{r plot_genes_dis}
p_genes_dist <-  combined_meta_prefilter %>%
  ggplot(aes(x = nFeature_RNA, color = library, fill = library)) +
  geom_density(alpha = 0.15) +
  scale_x_log10() +
  geom_vline(xintercept = QC_MIN_COUNTS, linetype = "dashed", color = "red") +
  geom_vline(xintercept = QC_MAX_COUNTS, linetype = "dashed", color = "red") +
  theme_classic(base_size = 10) +
  theme(legend.position = "none") +
  facet_wrap(~region, ncol = 2) +
  labs(title = "Distribution of number of genes (pre-filter)",
       x = "nFeature_RNA", y = "Density")


print(p_genes_dist)
plots_prefilter[["p_genes_dist"]] <- p_genes_dist

rm(p_genes_dist)
```


### Mitochondrial percentage

This metric can identify whether there is a large amount of
mitochondrial contamination from dead or dying cells. We define poor
quality samples for mitochondrial counts as cells which surpass the 0.2
mitochondrial ratio mark, unless of course you are expecting this in
your sample.


```{r plot_mt_dist}
p_mt_dist <- combined_meta_prefilter %>%
  dplyr::filter(percent.MT<=5) %>%  # Filter out extreme outliers for better visualization)
  ggplot(aes(x = percent.MT, color = library, fill = library)) +
  geom_density(alpha = 0.15) +
  geom_vline(xintercept = QC_MAX_MT, linetype = "dashed", color = "red") +
  theme_classic(base_size = 10) +
  theme(legend.position = "none") +
  facet_wrap(~region, ncol = 2) +
  labs(title = paste0("Mitochondrial % (pre-filter) | threshold = ", QC_MAX_MT, "%"),
       x = "% Mitochondrial reads", y = "Density")
print(p_mt_dist)
plots_prefilter[["p_mt_dist"]] <- p_mt_dist

rm(p_mt_dist)
```


### UMIs vs. genes detected (coloured by MT%)

Two metrics that are often evaluated together are the number of UMIs and
the number of genes detected per cell. Here, we have plotted the number
of genes versus the numnber of UMIs coloured by the fraction of
mitochondrial reads. Mitochondrial read fractions are only high (light
blue color) in particularly low count cells with few detected genes.
This could be indicative of damaged/dying cells whose cytoplasmic mRNA
has leaked out through a broken membrane, and thus, only mRNA located in
the mitochondria is still conserved. These cells are filtered out by our
count and gene number thresholds. Jointly visualizing the count and gene
thresholds shows the joint filtering effect.

Cells that are poor quality are likely to have low genes and UMIs per
cell, and correspond to the data points in the bottom left quadrant of
the plot. Good cells will generally exhibit both higher number of genes
per cell and higher numbers of UMIs.

With this plot we also evaluate the slope of the line, and any scatter
of data points in the bottom right hand quadrant of the plot. These
cells have a high number of UMIs but only a few number of genes. These
could be dying cells, but also could represent a population of a low
complexity celltype (i.e red blood cells).

```{r plot_umi_vs_genes}
p_umi_vs_genes <- combined_meta_prefilter %>%
  ggplot(aes(x = nCount_RNA, y = nFeature_RNA, color = percent.MT)) +
  geom_point(size = 0.3, alpha = 0.5) +
  scale_colour_gradient(low = "grey90", high = "red") +
  scale_x_log10() + scale_y_log10() +
  geom_vline(xintercept = QC_MIN_COUNTS, linetype = "dashed") +
  geom_hline(yintercept = QC_MIN_FEATURES, linetype = "dashed") +
  facet_wrap(~region, ncol = 2) +
  theme_classic(base_size = 9) +
  labs(title = "UMIs vs Genes per cell (pre-filter, 50k sample)",
       x = "nCount_RNA (log10)", y = "nFeature_RNA (log10)")
print(p_umi_vs_genes)
plots_prefilter[["p_umi_vs_genes"]] <- p_umi_vs_genes

rm(p_umi_vs_genes)
```


### Complexity (log10 Genes per UMI)


We can see the samples where we sequenced each cell less have a higher
overall complexity, that is because we have not started saturating the
sequencing for any given gene for these samples. Outlier cells in these
samples might be cells that have a less complex RNA species than other
cells. Sometimes we can detect contamination with low complexity cell
types like red blood cells via this metric. Generally, we expect the
novelty score to be above 0.80.

```{r plot_complexity}
p_complexity <- combined_meta_prefilter %>%
  ggplot(aes(x = log10GenesPerUMI, color = library, fill = library)) +
  geom_density(alpha = 0.15) +
  geom_vline(xintercept = QC_MIN_COMPLEXITY, linetype = "dashed", color = "red") +
  theme_classic(base_size = 10) +
  theme(legend.position = "none") +
  facet_wrap(~region, ncol = 2) +
  labs(title = "Transcriptome complexity (pre-filter)",
       x = "log10(Genes / UMI)", y = "Density")

print(p_complexity)

plots_prefilter[["p_complexity"]] <- p_complexity

rm(p_complexity)
```

```{r}
base::saveRDS(plots_prefilter, file.path(output_path, "RData", "QC_plots_prefilter.rds"))
plots_prefilter <- readRDS(paste0(output_path,"RData/QC_plots_prefilter.rds"))
rm(combined_meta_prefilter)
gc()
```


# 3. Cell Filtering

Three-layer filtering strategy:

1. **Demuxlet singlet filter** — Remove barcodes classified as doublets
   by Demuxlet (included in paper metadata: `DemuxletType == "SNG"`)
2. **Study inclusion filter** — Keep only cells the authors included
   (`IncludedInStudy == TRUE`)
3. **QC thresholds** — Standard human PBMC thresholds

```{r}
cat("Filtering Parameters:\n")
cat("===========================\n")
cat("Genes per cell: Minimun: ", QC_MIN_FEATURES,  ", Maximum: ", QC_MAX_FEATURES, "\n")
cat("Genes Counts: Minimun: ", QC_MIN_COUNTS,  ", Maximum: ", QC_MAX_COUNTS, "\n")
cat("Ribosomal % Counts: Minimun: ", QC_MIN_RIBO,  ", Maximum: ", QC_MAX_RIBO, "\n")
cat("Maximum % mitochondrial reads:", QC_MAX_MT,"\n\n" )
```

```{r filter_cells}
if (check_checkpoint("03_seurat_filtered")) {
  seurat_filtered <- load_checkpoint("03_seurat_filtered")
} else {
  libs_dir <- paste0(output_path, "RData/filtered_libs/")
  dir.create(libs_dir, recursive = TRUE, showWarnings = FALSE)

  seurat_qc  <- load_checkpoint("02_seurat_qc")
  lib_names  <- names(seurat_qc)

  # --- Sequential loop: filter one library at a time --------------------------
  # After filtering each library it is saved individually and its slot in
  # seurat_qc is nulled so peak RAM stays bounded (~size of seurat_qc alone).
  for (lib_name in lib_names) {
    lib_rds <- paste0(libs_dir, "filtered_", lib_name, ".rds")

    if (file.exists(lib_rds)) {
      message(sprintf("[SKIP] Already filtered: %s", lib_name))
      seurat_qc[[lib_name]] <- NULL   # free slot even when skipping
      gc()
      next
    }

    obj <- seurat_qc[[lib_name]]
    n0  <- ncol(obj)
    
    # Layer 1: Demuxlet singlet filter
    

    # Layer 2: Paper inclusion filter
    # NOTE: cells with IncludedInStudy == FALSE may still pass QC;
    #        we include them here but flag them for optional downstream exclusion
    # (keep all singlets regardless of IncludedInStudy for our own QC analysis)

    # Layer 3: Standard human PBMC QC thresholds
    obj <- subset(obj,
                  subset = nFeature_RNA     >= QC_MIN_FEATURES   &
                           nFeature_RNA     <= QC_MAX_FEATURES   &
                           nCount_RNA       >= QC_MIN_COUNTS     &
                           nCount_RNA       <= QC_MAX_COUNTS     &
                           percent.MT       <= QC_MAX_MT         &
                           log10GenesPerUMI >= QC_MIN_COMPLEXITY)
    n1 <- ncol(obj)

    message(sprintf("[FILTER] %s: %d → filter %d (kept %.1f%%)",
                    lib_name, n0, n1,  n1 / n0 * 100))

    saveRDS(obj, file = lib_rds)
    message(sprintf("[SAVED]  %s", lib_rds))

    # Free this library's slot before loading the next one
    seurat_qc[[lib_name]] <- NULL
    rm(obj); gc()
  }
  rm(seurat_qc); gc()

  # --- Assemble final list from individual checkpoints ------------------------
  message("\n[ASSEMBLE] Loading individual filtered libraries...")
  seurat_filtered <- lapply(lib_names, function(lib_name) {
    readRDS(paste0(libs_dir, "filtered_", lib_name, ".rds"))
  })
  names(seurat_filtered) <- lib_names

  save_checkpoint(seurat_filtered, "03_seurat_filtered")

  # To remove individual per-library files after the combined checkpoint is
  # confirmed good, uncomment and run manually:
  # unlink(paste0(libs_dir, "filtered_", lib_names, ".rds"))
}
```

## Post-filter QC summary

```{r qc_combined_meta}
combined_meta_filter <- build_combined_meta(seurat_filtered)
cat("Total barcodes with QC metrics:", nrow(combined_meta_filter), "\n")
cat("Libraries:", length(unique(combined_meta_filter$library)), "\n")

# Save metadata_prefilter for downstream use
save_checkpoint(combined_meta_filter, "03_combined_meta_filter")
```

### Cells per library

```{r plot_ncells_prelim}
p_cells_library <- combined_meta_filter %>%
  ggplot(aes(x = library, fill = condition)) +
  geom_bar() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  ggtitle("Number of Cells per Library") +
  xlab("Library") +
  ylab("Number of Cells") 
  # + facet_wrap(~region)


print(p_cells_library)
plots_filter[["p_cells_library"]] <- p_cells_library

rm(p_cells_library)
```
### UMI counts distribution

The UMI counts per cell (nCount_RNA) should generally be above 500, that
is the low end of what we expect. If UMI counts are between 500-1000
counts, it is usable but the cells probably should have been sequenced
more deeply.

```{r plot_umi_dist}
p_UMI_dist <- combined_meta_filter %>%
  ggplot(aes(x = nCount_RNA, color = library, fill = library)) +
  geom_density(alpha = 0.15) +
  scale_x_log10() +
  geom_vline(xintercept = QC_MIN_COUNTS, linetype = "dashed", color = "red") +
  geom_vline(xintercept = QC_MAX_COUNTS, linetype = "dashed", color = "red") +
  theme_classic(base_size = 10) +
  theme(legend.position = "none") +
  facet_wrap(~region, ncol = 2) +
  labs(title = "UMI distribution per library (filter)",
       x = "nCount_RNA (log10)", y = "Density")

print(p_UMI_dist)
plots_filter[["p_UMI_dist"]] <- p_UMI_dist

rm(p_UMI_dist)
```

### Genes detected per cell
We have similar expectations for gene detection as for UMI detection,
although it may be a bit lower than UMIs. For high quality data, the
proportional histogram should contain a single large peak that
represents cells that were encapsulated. If we see a small shoulder to
the right of the major peak (not present in our data), or a bimodal
distribution of the cells, that can indicate a couple of things. It
might be that there are a set of cells that failed for some reason. It
could also be that there are biologically different types of cells (i.e.
quiescent cell populations, less complex cells of interest), and/or one
type is much smaller than the other (i.e. cells with high counts may be
cells that are larger in size). Therefore, this threshold should be
assessed with other metrics that we describe in this lesson.

```{r plot_genes_dis}
p_genes_dist <-  combined_meta_filter %>%
  ggplot(aes(x = nFeature_RNA, color = library, fill = library)) +
  geom_density(alpha = 0.15) +
  scale_x_log10() +
  geom_vline(xintercept = QC_MIN_COUNTS, linetype = "dashed", color = "red") +
  geom_vline(xintercept = QC_MAX_COUNTS, linetype = "dashed", color = "red") +
  theme_classic(base_size = 10) +
  theme(legend.position = "none") +
  facet_wrap(~region, ncol = 2) +
  labs(title = "Distribution of number of genes (filter)",
       x = "nFeature_RNA", y = "Density")


print(p_genes_dist)
plots_filter[["p_genes_dist"]] <- p_genes_dist

rm(p_genes_dist)
```


### Mitochondrial percentage

This metric can identify whether there is a large amount of
mitochondrial contamination from dead or dying cells. We define poor
quality samples for mitochondrial counts as cells which surpass the 0.2
mitochondrial ratio mark, unless of course you are expecting this in
your sample.


```{r plot_mt_dist}
p_mt_dist <- combined_meta_filter %>%
  dplyr::filter(percent.MT<=5) %>%  # Filter out extreme outliers for better visualization)
  ggplot(aes(x = percent.MT, color = library, fill = library)) +
  geom_density(alpha = 0.15) +
  geom_vline(xintercept = QC_MAX_MT, linetype = "dashed", color = "red") +
  theme_classic(base_size = 10) +
  theme(legend.position = "none") +
  facet_wrap(~region, ncol = 2) +
  labs(title = paste0("Mitochondrial % (filter) | threshold = ", QC_MAX_MT, "%"),
       x = "% Mitochondrial reads", y = "Density")
print(p_mt_dist)
plots_filter[["p_mt_dist"]] <- p_mt_dist

rm(p_mt_dist)
```


### UMIs vs. genes detected (coloured by MT%)

Two metrics that are often evaluated together are the number of UMIs and
the number of genes detected per cell. Here, we have plotted the number
of genes versus the numnber of UMIs coloured by the fraction of
mitochondrial reads. Mitochondrial read fractions are only high (light
blue color) in particularly low count cells with few detected genes.
This could be indicative of damaged/dying cells whose cytoplasmic mRNA
has leaked out through a broken membrane, and thus, only mRNA located in
the mitochondria is still conserved. These cells are filtered out by our
count and gene number thresholds. Jointly visualizing the count and gene
thresholds shows the joint filtering effect.

Cells that are poor quality are likely to have low genes and UMIs per
cell, and correspond to the data points in the bottom left quadrant of
the plot. Good cells will generally exhibit both higher number of genes
per cell and higher numbers of UMIs.

With this plot we also evaluate the slope of the line, and any scatter
of data points in the bottom right hand quadrant of the plot. These
cells have a high number of UMIs but only a few number of genes. These
could be dying cells, but also could represent a population of a low
complexity celltype (i.e red blood cells).

```{r plot_umi_vs_genes}
p_umi_vs_genes <- combined_meta_filter %>%
  ggplot(aes(x = nCount_RNA, y = nFeature_RNA, color = percent.MT)) +
  geom_point(size = 0.3, alpha = 0.5) +
  scale_colour_gradient(low = "grey90", high = "red") +
  scale_x_log10() + scale_y_log10() +
  geom_vline(xintercept = QC_MIN_COUNTS, linetype = "dashed") +
  geom_hline(yintercept = QC_MIN_FEATURES, linetype = "dashed") +
  facet_wrap(~region, ncol = 2) +
  theme_classic(base_size = 9) +
  labs(title = "UMIs vs Genes per cell (filter, 50k sample)",
       x = "nCount_RNA (log10)", y = "nFeature_RNA (log10)")
print(p_umi_vs_genes)
plots_filter[["p_umi_vs_genes"]] <- p_umi_vs_genes

rm(p_umi_vs_genes)
```


### Complexity (log10 Genes per UMI)


We can see the samples where we sequenced each cell less have a higher
overall complexity, that is because we have not started saturating the
sequencing for any given gene for these samples. Outlier cells in these
samples might be cells that have a less complex RNA species than other
cells. Sometimes we can detect contamination with low complexity cell
types like red blood cells via this metric. Generally, we expect the
novelty score to be above 0.80.

```{r plot_complexity}
p_complexity <- combined_meta_filter %>%
  ggplot(aes(x = log10GenesPerUMI, color = library, fill = library)) +
  geom_density(alpha = 0.15) +
  geom_vline(xintercept = QC_MIN_COMPLEXITY, linetype = "dashed", color = "red") +
  theme_classic(base_size = 10) +
  theme(legend.position = "none") +
  facet_wrap(~region, ncol = 2) +
  labs(title = "Transcriptome complexity (filter)",
       x = "log10(Genes / UMI)", y = "Density")

print(p_complexity)

plots_filter[["p_complexity"]] <- p_complexity

rm(p_complexity)
```

```{r}
base::saveRDS(plots_filter, file.path(output_path, "RData", "QC_plots_filter.rds"))
plots_filter <- readRDS(paste0(output_path,"RData/QC_plots_filter.rds"))

rm(combined_meta_filter)
gc()
```

## Plots QC

### Cells per library
```{r}
cells_library_plot <- plots_prefilter[["p_cells_library"]] + plots_filter[["p_cells_library"]] +
  plot_annotation(title = "Cells per library (pre vs post filter)",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")),
                  tag_levels = "A")
print(cells_library_plot)
save_plot("cells_per_library_prefilter_vs_filter", cells_library_plot)
```

### UMI counts distribution
```{r}
umi_counts_plot <- plots_prefilter[["p_UMI_dist"]] + plots_filter[["p_UMI_dist"]]+
  plot_annotation(title = "UMI counts distribution (pre vs post filter)",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")),
                  tag_levels = "A")
print(umi_counts_plot)
save_plot("UMI_counts_distribution_per_library_prefilter_vs_filter", umi_counts_plot)
```

### Genes detected per cell
```{r}
genes_detected_plot <- plots_prefilter[["p_genes_dist"]] + plots_filter[["p_genes_dist"]] +
  plot_annotation(title = "Genes detected per cell (pre vs post filter)",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")),
                  tag_levels = "A")
print(genes_detected_plot)
save_plot("Genes_detected_per_cell_prefilter_vs_filter", genes_detected_plot)
```

### Mitochondrial percentage
```{r}
mitochondrial_plot <- plots_prefilter[["p_mt_dist"]] + plots_filter[["p_mt_dist"]] +
  plot_annotation(title = "Mitochondrial percentage (pre vs post filter)",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")),
                  tag_levels = "A")
print(mitochondrial_plot)
save_plot("mitochondrial_percentage_per_library_prefilter_vs_filter", mitochondrial_plot)
```

### UMIs vs. genes detected (coloured by MT%)
```{r}
umi_vs_genes_plot <- plots_prefilter[["p_umi_vs_genes"]] + plots_filter[["p_umi_vs_genes"]] +
  plot_annotation(title = "UMI counts vs. genes detected (pre vs post filter)",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")),
                  tag_levels = "A")
print(umi_vs_genes_plot)
save_plot("UMI_counts_vs_genes_detected_per_library_prefilter_vs_filter", umi_vs_genes_plot)
```

### Complexity (log10 Genes per UMI)
```{r}
complexity_plot <- plots_prefilter[["p_complexity"]] + plots_filter[["p_complexity"]] +
  plot_annotation(title = "Complexity (pre vs post filter)",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")),
                  tag_levels = "A")
print(complexity_plot)
save_plot("Complexity_per_library_prefilter_vs_filter", complexity_plot)
```

```{r}
rm(plots_filter, plots_prefilter)
gc()
```





